---
title: "Regular expressions (regex)"
subtitle: "Thursday, June 18, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day23-regex-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day23-regex-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(dcData)
data(BabyNames)

NameList <- BabyNames |>
  group_by(name, sex) |>
  summarise(total = sum(count), .groups = "drop") |>
  arrange(desc(total))
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Thursday June 18]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Reading** (Canvas): [*Data Computing* Ch. 17](https://psu.instructure.com/courses/2487468/assignments/18325379) §17.1–17.2 — Regular expressions

**Plan for today**

- **Project oral check-in #1**; work on the activity if your group doesn't go first.
- What a **regex** and its roles (**detect, replace, extract**)
- literals, `.`, character classes `[ ]`, alternation `|`
- **Quantifiers** `? * + {n}`, **anchors** `^ $`, and **escaping**
- **Capturing** with `( )` and pulling pieces out
- In-class activity

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*DC* Ch. 17](https://dtkaplan.github.io/DataComputingEbook/chap-regular-expressions.html) · [stringr cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/strings.pdf) · [regex101.com](https://regex101.com/)
:::

# What is a regex? {background-color="#001E44"}

## Patterns in strings {.scrollable .nostretch}

A **regular expression** (regex) is a compact way to describe a **pattern** in text. Lots of everyday strings follow patterns:

| Kind | Examples |
|---|---|
| Phone numbers | `(651) 696-6000` · `+1 651.696.6000` |
| Times | `10:30 AM` · `13:15.54` |
| Durations | `1:59:40` (Kipchoge's sub-2-hour marathon) |
| US postal codes | `16801` · `55105-1362` |
| Stray whitespace | `"  henry  "` |

A regex lets you say "match anything that has this particular shape" without having to list out every possible case (which is often impossible).

## Three jobs regex does {.scrollable .nostretch}

Across all of data cleaning, regex does three things:

| Job | base R | stringr (tidyverse) | Often with |
|---|---|---|---|
| **Detect** a pattern | `grepl()` | `str_detect()` | `filter()` |
| **Replace** a pattern | `gsub()` | `str_replace_all()` | `mutate()` |
| **Extract** a match | `regmatches()` | `str_extract()` | `mutate()` |

The textbook uses base R (`grepl`, `gsub`); the analogous versions from the stringr package do the same things but have tidier names.

## grepl(): does the pattern appear? {.scrollable .nostretch}

`grepl(pattern, x)` returns `TRUE`/`FALSE` for each string. The **pattern is the quoted string**. Here: names containing "shine" (`ignore.case = TRUE` ignores capitalization):

```{r}
#| echo: TRUE
NameList |>
  filter(grepl("shine", name, ignore.case = TRUE)) |>
  head(5)
```

`str_detect(name, regex("shine", ignore_case = TRUE))` does the same thing.

# Regex building blocks {background-color="#001E44"}

## Literals & special characters {.scrollable .nostretch}

The simplest patterns are **literal**: `"able"` matches any string containing "able" ("cap**able**", "t**able**d") but is **case-sensitive** by default ("Able" won't match unless `ignore.case = TRUE`).

These characters are **special** in that they mean something other than themselves:

``` text
.   ^   $   [   ]   {   }   |   (   )   +   *   ?
```

For example `.` means "any single character", so `".ble"` matches "able", "sol**uble**", "fung**ible**", "B**ible**".

## Character classes `[ ]` {.scrollable .nostretch}

Square brackets mean "**any one** of these characters":

- `[aeiou]` $\rightarrow$ any vowel
- `[^aeiou]` $\rightarrow$ **not** a vowel (the `^` *inside* brackets means "except") $\rightarrow$ a consonant
- `[0-9]` $\rightarrow$ any digit; `[a-z]` $\rightarrow$ any lowercase letter

```{r}
#| echo: TRUE
NameList |>
  filter(grepl("mn", name, ignore.case = TRUE)) |>   # literal "mn"
  head(3)
```

::: {.callout-tip}
A vertical bar `|` means **"or"**: `"rain|snow|sleet|hail"` matches any of those four words.
:::

## Putting characters in a row {.scrollable .nostretch}

Two patterns side by side means "this **then** that". `^[^aeiou].[^aeiou].[^aeiou]` reads as:

- `^` start of string, then
- `[^aeiou]` a consonant, `.` anything, `[^aeiou]` a consonant, `.` anything, `[^aeiou]` a consonant

i.e. the 1st, 3rd, and 5th letters are consonants:

```{r}
#| echo: TRUE
NameList |>
  filter(grepl("^[^aeiou].[^aeiou].[^aeiou]", name, ignore.case = TRUE)) |>
  head(3)
```

# Quantifiers {background-color="#001E44"}

## How many times? {.scrollable .nostretch}

A quantifier follows a pattern and says **how many times** it repeats:

| Quantifier | Means |
|---|---|
| `?` | zero or **one** time |
| `*` | zero or **more** times |
| `+` | **one** or more times |
| `{n}` | exactly `n` times |
| `{n,m}` | between `n` and `m` times |
| `{n,}` | `n` or more times |

::: {.callout-warning}
The *Data Computing* reading swaps the definitions of `+` and `*`. **In R (and every real regex engine):** `*` = zero-or-more, `+` = one-or-more.
:::

## Quantifiers in action {.scrollable .nostretch}

```{r}
#| echo: TRUE
x <- c("Mr.", "Ms", "Mrs.", "Miss", "Man", "Miocene")
x[grepl("M[aeiou]+", x)]   # M then ONE or more vowels
x[grepl("M[aeiou]*", x)]   # M then ZERO or more vowels (matches Mr.)
```

Three or more vowels in a row, `[aeiou]{3,}`:

```{r}
#| echo: TRUE
NameList |>
  filter(grepl("[aeiou]{3,}", name, ignore.case = TRUE)) |>
  head(3)
```

# Anchors & escaping {background-color="#001E44"}

## Start `^` and end `$` {.scrollable .nostretch}

Outside of square brackets, `^` and `$` are **anchors**:

- `^` = the **start** of the string
- `$` = the **end** of the string

Names that **end** in a vowel, `[aeiou]$`:

```{r}
#| echo: TRUE
NameList |>
  filter(grepl("[aeiou]$", name)) |>
  group_by(sex) |>
  summarise(total = sum(total), .groups = "drop")
```

Girls' names end in a vowel almost five times as often as boys' names.

## Escaping special characters {.scrollable .nostretch}

To match a special character **literally**, **escape** it with `\\`. So `\\.` means a real period, and `\\$` a real dollar sign (not the end-of-string anchor).

```{r}
#| echo: TRUE
currency <- c("$100.95", "45¢")
gsub("^\\$|¢$", "", currency)   # strip a leading $ or trailing ¢
```

Here `^\\$` is a literal `$` at the start; `¢$` is a literal `¢` at the end. Without the `\\`, that first `$` would mean "end of string".

# Capturing & extracting {background-color="#001E44"}

## Parentheses capture {.scrollable .nostretch}

Wrap part of a pattern in `( )` to **capture** it; in other words, pull that piece out. `str_extract()` returns the matched text; with capture groups, `gsub()` can refer back to group 1 as `\\1`.

Names with **four vowels in a row**, extracting those vowels:

```{r}
#| echo: TRUE
BabyNames |>
  group_by(name) |>
  summarise(total = sum(count), .groups = "drop") |>
  mutate(match = str_extract(name, "([aeiouAEIOU]{4})")) |>
  filter(!is.na(match)) |>
  arrange(desc(total)) |>
  head(4)
```

## Backreferences with `\\1` {.scrollable .nostretch}

The regex `".*([aeiou])$"` = "anything, then a captured final vowel". In `gsub()`, `replacement = "\\1"` keeps **only** the captured group, so the whole name collapses to its last vowel:

```{r}
#| echo: TRUE
re <- ".*([aeiou])$"
NameList |>
  filter(grepl(re, name)) |>
  mutate(vowel = gsub(re, "\\1", name)) |>
  group_by(sex, vowel) |>
  summarise(total = sum(total), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = total)
```

Names ending in "a", "e", "i" skew female; "o" skews male.

# Cleaning with regex {background-color="#001E44"}

## Stripping junk out of numbers {.scrollable .nostretch}

Scraped numbers arrive with `$`, `,`, `%`, units. A character class lists everything to delete at once:

```{r}
#| echo: TRUE
debt <- tibble(
  country = c("World", "United States*", "Japan"),
  debt    = c("$56,308 billion", "$17,607 billion", "$9,872 billion"),
  percGDP = c("64%", "73.60%", "214.30%")
)
debt |>
  mutate(
    debt    = gsub("[$,%]|billion", "", debt),
    percGDP = gsub("[,%]", "", percGDP)
  )
```

`[$,%]` deletes any `$`, `,`, or `%`; `|billion` also deletes the word "billion".

## Trimming whitespace {.scrollable .nostretch}

`^ +| +$` matches one-or-more spaces at the **start** *or* the **end**:

```{r}
#| echo: TRUE
s <- "   My name is Julia     "
gsub("^ +| +$", "", s)
```

::: {.callout-tip}
stringr has friendly shortcuts for these: `str_remove_all()`, `str_trim()` (ends), and `str_squish()` (ends + collapse internal runs). Same regex ideas underneath.
:::

# Activity {background-color="#7C0A02"}

## In-class activity {.scrollable .nostretch}

Open `day23-activity.R` in RStudio and work through the parts in order. (Work on the activity when it isn't your group's turn for **oral check-in #1**.)

- **Part 1**: **detect** patterns with `grepl()` / `str_detect()` + `filter()` on `BabyNames`.
- **Part 2**: build patterns from **character classes, quantifiers, and anchors** (including a "Caitlin"-style name and a phone-number regex).
- **Part 3 (challenge)**: **extract** with capture groups, and **clean** messy strings with `gsub()`.

[Activity file](day23-activity.R){download="day23-activity.R"}

## Wrap-up & looking ahead {.scrollable .nostretch}

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- A regex describes a **pattern**; it does three jobs: **detect / replace / extract**
- `.` any char · `[aeiou]` class · `[^aeiou]` negation · `|` or
- Quantifiers `?` `*` `+` `{n}` `{n,m}` `{n,}` (R: `+` = one-or-more, `*` = zero-or-more)
- Anchors `^` start, `$` end; **escape** specials with `\\`
- Capture with `( )`; pull pieces with `str_extract()` or `\\1` in `gsub()`
- Clean numbers/whitespace with `gsub()` (or `str_remove_all`/`str_squish`)

**Upcoming**

- **Fri Jun 19:** **No class — Juneteenth**
- **Quiz 5** rescheduled to **Mon Jun 22**
- Keep building your **project**; oral check-in #1 today
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*DC* Ch. 17](https://dtkaplan.github.io/DataComputingEbook/chap-regular-expressions.html) · [regex101.com](https://regex101.com/)
:::
::::
