---
title: "Writing functions in R"
subtitle: "Monday, June 15, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day20-functions-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day20-functions-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(nycflights13)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Monday June 15]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Reading** (Canvas): [*R4DS* Ch. 25](https://r4ds.hadley.nz/functions): Functions
- **HW 4** due tonight, 11:59 p.m.

**Plan for today**

- *Why* functions: the cost of copy-and-paste
- **Vector functions**: `rescale01()`, mutate helpers, summary helpers
- **Data frame functions** and the embracing operator `{{ }}`
- **Plot functions** with `ggplot2`
- Naming & style; in-class activity

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 25](https://r4ds.hadley.nz/functions) · [tidyverse style guide](https://style.tidyverse.org/functions.html)
:::

# Why write functions? {background-color="#001E44"}

## The rule of three {.scrollable .nostretch}

> *"You should consider writing a function whenever you've copied and pasted a block of code more than twice."* - *R4DS*

Functions have four big advantages over copy-and-paste:

1. You give the code an **evocative name**, so it's easier to read.
2. As requirements change, you update in **one place** rather than many.
3. You **eliminate copy-paste mistakes** (changing a name in one spot but not another).
4. You can **reuse work** across projects.

Writing functions allows us to go from *using* tidyverse verbs to *writing our own tools*.

## Spot the bug {.scrollable .nostretch}

This code rescales every column to run from 0 to 1. Can you find the copy-paste bug?

```{r}
#| echo: TRUE
#| eval: FALSE
df <- tibble(a = rnorm(5), b = rnorm(5), c = rnorm(5), d = rnorm(5))

df |> mutate(
  a = (a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE)),
  b = (b - min(a, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE)),
  c = (c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE)),
  d = (d - min(d, na.rm = TRUE)) / (max(d, na.rm = TRUE) - min(d, na.rm = TRUE)),
)
```

::: {.callout-warning}
Line `b` divides by `min(a, ...)` instead of `min(b, ...)`. Copy-paste invites this kind of typo, but writing this as a function prevents the possibility of this type of bug altogether.
:::

# Vector functions {background-color="#001E44"}

## What varies vs. stays the same {.scrollable .nostretch}

Pull the repeated line out and line the copies up. The pattern jumps out:

``` r
(a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE))
(b - min(b, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE))
(c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE))
```

Replace the part that **changes** with `█`:

``` r
(█ - min(█, na.rm = TRUE)) / (max(█, na.rm = TRUE) - min(█, na.rm = TRUE))
```

The only thing that varies is the vector. That becomes our function **argument**.

## The anatomy of a function {.scrollable .nostretch}

To write a function you need three things:

- a **name**: here `rescale01` (it rescales a vector to 0–1),
- the **arguments**: the things that vary; here just one, call it `x`,
- the **body**: the code that's repeated.

``` r
name <- function(arguments) {
  body
}
```

```{r}
#| echo: TRUE
rescale01 <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}
```

## Always test on simple inputs {.scrollable .nostretch}

Before trusting a function, test run it with inputs where you already know the answer:

```{r}
#| echo: TRUE
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

Now the `mutate()` is short, readable, and more bug-resistant:

```{r}
#| echo: TRUE
df <- tibble(a = rnorm(5), b = rnorm(5), c = rnorm(5), d = rnorm(5))
df |> mutate(a = rescale01(a), b = rescale01(b),
             c = rescale01(c), d = rescale01(d))
```

## Changing a function in one place {.scrollable .nostretch}

Because the logic is in *one* spot rather than 4+, improving it is easy. `range()` finds the min and max in a single step:

```{r}
#| echo: TRUE
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(1:10, Inf))
```

::: {.callout-tip}
`finite = TRUE` tells `range()` to ignore `Inf`. We fixed two behaviors at once.
:::

## "Mutate" functions {.scrollable .nostretch}

A **mutate function** returns an output the *same length* as its input, so it slots into `mutate()` and `filter()`.

```{r}
#| echo: TRUE
# z-score: center and scale to mean 0, sd 1
z_score <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

# clamp values into [min, max]
clamp <- function(x, min, max) {
  case_when(
    x < min ~ min,
    x > max ~ max,
    .default = x
  )
}
clamp(1:10, min = 3, max = 7)
```

## Mutate functions aren't just for numbers {.scrollable .nostretch}

```{r}
#| echo: TRUE
# capitalize the first letter
first_upper <- function(x) {
  str_sub(x, 1, 1) <- str_to_upper(str_sub(x, 1, 1))
  x
}
first_upper("hello")
```

```{r}
#| echo: TRUE
# strip $, commas, % then convert to a number
clean_number <- function(x) {
  is_pct <- str_detect(x, "%")
  num <- x |>
    str_remove_all("%") |> str_remove_all(",") |> str_remove_all(fixed("$")) |>
    as.numeric()
  if_else(is_pct, num / 100, num)
}
clean_number(c("$12,300", "45%"))
```

## "Summary" functions {.scrollable .nostretch}

A **summary function** collapses a vector to a *single value*, so it slots into `summarize()`.

```{r}
#| echo: TRUE
# coefficient of variation: sd relative to the mean
cv <- function(x, na.rm = FALSE) {
  sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
}

# a memorable name for a common pattern
n_missing <- function(x) {
  sum(is.na(x))
}

flights |> summarize(
  n_missing_delay = n_missing(dep_delay),
  cv_delay        = cv(dep_delay, na.rm = TRUE)
)
```

::: {.callout-tip}
Note the **default argument** `na.rm = FALSE`. You can override it, but don't have to. If you don't specify, it won't automatically fail, but rather falls back to the default. 
:::

# Data frame functions {background-color="#001E44"}

## Repeating verbs, not just expressions {.scrollable .nostretch}

Vector functions pull out code repeated *inside* a verb. But often you repeat the **verbs themselves**. A **data frame function**:

- takes a data frame as its **first argument**,
- takes extra arguments saying what to do,
- returns a data frame (or a vector).

The natural first attempt looks like this, but it fails:

```{r}
#| echo: TRUE
#| eval: FALSE
grouped_mean <- function(df, group_var, mean_var) {
  df |>
    group_by(group_var) |>
    summarize(mean(mean_var))
}
flights |> grouped_mean(origin, dep_delay)
#> Error: Column `group_var` is not found.
```

## The problem: indirection {.scrollable .nostretch}

dplyr uses **tidy evaluation** so you can write `dep_delay` instead of `flights$dep_delay`. That's great most of the time, but inside a function, `group_by(group_var)` looks for a column *named* `group_var` that doesn't exist.

We need to tell dplyr: *don't use the argument's name, look* **inside** *it.*

The solution is **embracing** using `{{ }}`.

::: {.callout-tip}
`{{ var }}` is analogous to looking **down a tunnel**: it makes dplyr look inside `var` rather than searching for a column called `var`.
:::

## Embracing fixes it {.scrollable .nostretch}

```{r}
#| echo: TRUE
grouped_mean <- function(df, group_var, mean_var) {
  df |>
    group_by({{ group_var }}) |>
    summarize(mean = mean({{ mean_var }}, na.rm = TRUE), .groups = "drop")
}
flights |> grouped_mean(origin, dep_delay)
```

Now `grouped_mean(origin, dep_delay)` really does group by `origin` and average `dep_delay`.

## When do I need to embrace? {.scrollable .nostretch}

Two kinds of tidy evaluation:

| Type | Used by | You can... |
|---|---|---|
| **Data-masking** | `filter()`, `arrange()`, `summarize()`, `group_by()`, `mutate()` | *compute* with variables (`x + 1`) |
| **Tidy-selection** | `select()`, `relocate()`, `rename()`, `pivot_wider()` | *select* variables (`a:x`, `starts_with()`) |

If an argument gets passed to a verb that uses tidy evaluation, typically we want to **embrace it**.

## A handy summary helper {.scrollable .nostretch}

Wrap up the summaries you always run during exploration:

```{r}
#| echo: TRUE
summary6 <- function(data, var) {
  data |> summarize(
    min = min({{ var }}, na.rm = TRUE),
    mean = mean({{ var }}, na.rm = TRUE),
    median = median({{ var }}, na.rm = TRUE),
    max = max({{ var }}, na.rm = TRUE),
    n = n(),
    n_miss = sum(is.na({{ var }})),
    .groups = "drop"
  )
}
diamonds |> group_by(cut) |> summary6(carat)
```

Because it wraps `summarize()`, it works on both **grouped** data and on **computed** variables (`summary6(log10(carat))`).

## count() with proportions {.scrollable .nostretch}

```{r}
#| echo: TRUE
count_prop <- function(df, var, sort = FALSE) {
  df |>
    count({{ var }}, sort = sort) |>
    mutate(prop = n / sum(n))
}
diamonds |> count_prop(clarity)
```

Only `var` is embraced; here, `df` and `sort` are ordinary arguments.

## Selecting inside a data-masking function {.scrollable .nostretch}

What if you want to group by *several* user-supplied columns? `group_by()` is data-masking, not tidy-selection, so `c(year, month, day)` breaks. The fix is **`pick()`**:

```{r}
#| echo: TRUE
count_missing <- function(df, group_vars, x_var) {
  df |>
    group_by(pick({{ group_vars }})) |>
    summarize(n_miss = sum(is.na({{ x_var }})), .groups = "drop")
}
flights |> count_missing(c(year, month, day), dep_time)
```

`pick()` lets you use **tidy-selection** inside a **data-masking** function.

# Plot functions {background-color="#001E44"}

## Functions that return a plot {.scrollable .nostretch}

`aes()` is a data-masking function, so the *same* `{{ }}` trick builds plotting functions:

```{r}
#| echo: TRUE
#| fig-height: 3
histogram <- function(df, var, binwidth = NULL) {
  df |>
    ggplot(aes(x = {{ var }})) +
    geom_histogram(binwidth = binwidth)
}
diamonds |> histogram(carat, 0.1)
```

## You can still add layers {.scrollable .nostretch}

A plot function returns a ggplot object, so keep building with `+` (not `|>`):

```{r}
#| echo: TRUE
#| fig-height: 3
diamonds |>
  histogram(carat, 0.1) +
  labs(x = "Size (in carats)", y = "Number of diamonds")
```

## More variables {.scrollable .nostretch}

Embrace each variable you pass to `aes()`:

```{r}
#| echo: TRUE
#| fig-height: 3
#| warning: FALSE
#| message: FALSE
linearity_check <- function(df, x, y) {
  df |>
    ggplot(aes(x = {{ x }}, y = {{ y }})) +
    geom_point() +
    geom_smooth(method = "loess", formula = y ~ x, color = "red", se = FALSE) +
    geom_smooth(method = "lm", formula = y ~ x, color = "blue", se = FALSE)
}
mpg |> linearity_check(displ, hwy)
```

## The walrus operator `:=` {.scrollable .nostretch}

To put an embraced name on the **left** of `=`, use `:=`:

```{r}
#| echo: TRUE
#| fig-height: 3
sorted_bars <- function(df, var) {
  df |>
    mutate({{ var }} := fct_rev(fct_infreq({{ var }}))) |>
    ggplot(aes(y = {{ var }})) +
    geom_bar()
}
diamonds |> sorted_bars(clarity)
```

R's syntax won't allow `{{ var }} =`, so tidy evaluation uses `:=` to mean the same thing.

## Labeling with the variable name {.scrollable .nostretch}

`rlang::englue()` works like `str_glue()` but also understands `{{ }}`, inserting the variable's *name* into a string:

```{r}
#| echo: TRUE
#| fig-height: 2.8
histogram <- function(df, var, binwidth) {
  label <- rlang::englue("A histogram of {{var}} with binwidth {binwidth}")
  df |>
    ggplot(aes(x = {{ var }})) +
    geom_histogram(binwidth = binwidth) +
    labs(title = label)
}
diamonds |> histogram(carat, 0.1)
```

# Naming & style {background-color="#001E44"}

## Names are for humans {.scrollable .nostretch}

R doesn't care what you name things, but smart naming makes our lives easier.

- Function names should be **verbs**; arguments should be **nouns**.
  - exceptions: well-known nouns (`mean()`), or accessors (`coef()`).
- Prefer **clear over short** since autocomplete can type the long name for you.

``` r
f()                      # too short, meaningless
my_awesome_function()    # not a verb, not descriptive
impute_missing()         # long but clear
collapse_years()         # long but clear
```

## Whitespace conventions {.scrollable .nostretch}

- `function(...)` is always followed by `{ }`.
- The body is indented **two extra spaces**.
- Put spaces inside `{{ }}` so the embracing stands out.

``` r
# good
density <- function(color, facets, binwidth = 0.1) {
  diamonds |>
    ggplot(aes(x = carat, y = after_stat(density), color = {{ color }})) +
    geom_freqpoly(binwidth = binwidth) +
    facet_wrap(vars({{ facets }}))
}
```

Skimming the left-hand margin should reveal the structure of your code.

# Activity {background-color="#7C0A02"}

## In-class activity {.scrollable .nostretch}

Open `day20-activity.R` in RStudio and work through the parts in order.

- **Part 1**: write & test **vector functions** (`rescale01`, a mutate helper, a summary helper).
- **Part 2**: write **data frame functions** using embracing `{{ }}` on `nycflights13`.
- **Part 3 (challenge)**: write a **plot function**, use `:=`, and label it with `rlang::englue()`.

[Activity file](day20-activity.R){download="day20-activity.R"}

## Wrap-up & looking ahead {.scrollable .nostretch}

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- Write a function once you've copy-pasted **more than twice**
- **Vector functions**: same-length-out (mutate) vs single-value-out (summary)
- Always **test on known inputs**; improve in **one place**
- **Data frame functions**: embrace data-masking args with `{{ }}`
- `pick()` for tidy-selection inside data-masking; `:=` for names
- **Plot functions**: embrace inside `aes()`; label with `rlang::englue()`
- Names are **verbs**, clear beats short

**Upcoming**

- **Tue Jun 16:** control flow: `if/else`, `for`, `while`
- **Reading before Tue:** *R4DS* §26.1–26.3
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 25](https://r4ds.hadley.nz/functions) · [tidyverse style guide](https://style.tidyverse.org/functions.html)
:::
::::
