---
title: "Iteration: working across many columns"
subtitle: "Tuesday, June 16, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day21-iteration-across-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day21-iteration-across-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(nycflights13)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Tuesday June 16]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Reading**: [*R4DS* Ch. 26](https://psu.instructure.com/courses/2487468/assignments/18325377) §26.1–26.3

**Plan for today**

- *Iteration* intro
- The copy-and-paste problem across **columns**
- **`across()`**: `.cols`, `.fns`, `.names`
- Selecting columns with `where()`; passing functions *without* `()`
- Anonymous functions `\(x)`, multiple functions, `if_any()` / `if_all()`
- In-class activity

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 26](https://r4ds.hadley.nz/iteration) · [dplyr `across()` reference](https://dplyr.tidyverse.org/reference/across.html)
:::

# Why iteration? {background-color="#001E44"}

## R's built-in iteration {.scrollable .nostretch}

**Iteration** is repeatedly performing the same action on different objects. In R, much of it is *implicit*. We do it without even realizing. 

To double every element of a vector, you just write:

```{r}
#| echo: TRUE
x <- c(1, 2, 3, 4)
2 * x
```

In many other languages you'd write a loop. In R, the operation is **vectorized**.

You've also already used tools that repeat an action over **subsets**:

- `facet_wrap()` / `facet_grid()`: one plot per group
- `group_by()` + `summarize()`: one summary per group
- `unnest_wider()` / `unnest_longer()`: one row/column per list element

## Functional programming {.scrollable .nostretch}

**Functional programming** tools are functions that *take other functions as inputs*.

Last class, you learned to *write* functions. Now you'll *hand them to other functions* to do the repeating for you.

For the first of two iteration lectures (today), we'll focus on:

> **Doing the same thing to many columns at once.**

The key tool is dplyr's **`across()`**.

# Modifying multiple columns {background-color="#001E44"}

## The copy-and-paste problem {.scrollable .nostretch}

Suppose we want the median of every column of this tibble:

```{r}
#| echo: TRUE
set.seed(1014)
df <- tibble(a = rnorm(10), b = rnorm(10), c = rnorm(10), d = rnorm(10))

df |> summarize(
  n = n(),
  a = median(a),
  b = median(b),
  c = median(c),
  d = median(d)
)
```

This breaks the **rule of three** from last class. With four columns, it isn't that bad, but imagine copying and pasting the same line 100 times. 

## across() {.scrollable .nostretch}

`across()` repeats an action over a set of columns:

```{r}
#| echo: TRUE
df |> summarize(
  n = n(),
  across(a:d, median)
)
```

`across()` main arguments:

| Argument | Role |
|---|---|
| `.cols` | **which** columns to iterate over (uses `select()` syntax) |
| `.fns` | **what** to do to each column |
| `.names` | (optional) **how** to name the output columns |

## .cols: choosing columns {.scrollable .nostretch}

`.cols` accepts everything `select()` does: `a:d`, `starts_with()`, `ends_with()`. Two selectors are especially handy with `across()`:

- **`everything()`**: every (non-grouping) column
- **`where()`**: select by *type*: `where(is.numeric)`, `where(is.character)`, `where(is.Date)`, ...

```{r}
#| echo: TRUE
set.seed(1014)
df <- tibble(
  grp = sample(2, 10, replace = TRUE),
  a = rnorm(10), b = rnorm(10), c = rnorm(10), d = rnorm(10)
)

df |>
  group_by(grp) |>
  summarize(across(everything(), median))
```

Grouping columns (`grp`) are excluded automatically since `summarize()` already preserves them.

## where(): select by type {.scrollable .nostretch}

`where()` is very helpful for handling real data with mixed column types. Here, average every **numeric** column of `flights`:

```{r}
#| echo: TRUE
flights |>
  summarize(across(where(is.numeric), \(x) mean(x, na.rm = TRUE))) |>
  glimpse()
```

You can combine selectors with Boolean logic: `!where(is.numeric)` (all non-numeric), or `starts_with("dep") & where(is.numeric)`.

## .fns {.scrollable .nostretch}

We pass `median` to `across`. This is *one function handed to another function*, which is an example of functional programming.

::: {.callout-warning}
Pass the **name only** with no parentheses. For example: `across(a:d, median)`, not `across(a:d, median())`.
:::

`median()` with no argument throws an error, because you've *called* it with nothing to work on:

```{r}
#| echo: TRUE
#| error: TRUE
median()
```

## Adding arguments with anonymous functions {.scrollable .nostretch}

What if there are `NA`s? `median()` propagates them:

```{r}
#| echo: TRUE
df_miss <- tibble(a = c(1, 2, NA, 4), b = c(NA, 5, 6, 7))
df_miss |> summarize(across(a:b, median))
```

We want `na.rm = TRUE` to stop this propagation. We can wrap `median()` in a tiny **anonymous function**, `\(x)`. Here, `\(x)` is shorthand for `function(x)`, aka, a function that takes argument(s) `x`: 

```{r}
#| echo: TRUE
df_miss |> summarize(across(a:b, \(x) median(x, na.rm = TRUE)))
```

In English, you can read this as `\(x) median(x, na.rm = TRUE)` as "given `x`, return its median, dropping NAs."

## What across() expands to {.scrollable .nostretch}

It can help to picture what `across()` writes for you. This:

```{r}
#| echo: TRUE
#| eval: FALSE
df_miss |> summarize(across(a:b, \(x) median(x, na.rm = TRUE)))
```

is exactly equivalent to:

```{r}
#| echo: TRUE
#| eval: FALSE
df_miss |> summarize(
  a = median(a, na.rm = TRUE),
  b = median(b, na.rm = TRUE)
)
```

`across()` is just a more concise way to write the copy-pasted version. The extra bonus is that the `across()` approach prevents copy-paste mistakes.

## Multiple functions at once {.scrollable .nostretch}

We can calculate the median *and* count the missings by passing a **named list** to `.fns`:

```{r}
#| echo: TRUE
df_miss |>
  summarize(
    across(a:b, list(
      median = \(x) median(x, na.rm = TRUE),
      n_miss = \(x) sum(is.na(x))
    )),
    n = n()
  )
```

The new columns are named `{.col}_{.fn}`, e.g. `a_median`, `a_n_miss`. This syntax is from the **glue** package, and you can adjust the naming. For example, you could switch the order and instead write `.names = "{.fn}_{.col}"`.

## across() in mutate() {.scrollable .nostretch}

`across()` works well inside `mutate()` too. Recall `rescale01()` from last class. This function lets us rescale four columns in one line:

```{r}
#| echo: TRUE
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

set.seed(1014)
df <- tibble(a = rnorm(5), b = rnorm(5), c = rnorm(5), d = rnorm(5))
df |> mutate(across(a:d, rescale01))
```

The punchline: **write a function once**, then **apply it everywhere you want in one line** with `across()`.

## Filtering with if_any() and if_all() {.scrollable .nostretch}

`across()` works in `summarize()` and `mutate()`. For `filter()`, we use similar helpers:

- **`if_any()`**: keep a row if the condition holds for *any* of the columns
- **`if_all()`**: keep a row if it holds for *all* of the columns

```{r}
#| echo: TRUE
# keep only rows with NO missing values in any of these timing columns
flights |>
  filter(if_all(c(dep_time, arr_time, dep_delay, arr_delay), \(x) !is.na(x))) |>
  nrow()
```

# Activity {background-color="#7C0A02"}

## In-class activity {.scrollable .nostretch}

Open `day21-activity.R` in RStudio and work through the parts in order.

- **Part 1**: replace copy-pasted `summarize()` with `across()`; choose columns with `.cols`.
- **Part 2**: handle `NA`s with anonymous functions `\(x)`; apply *multiple* functions with a named list and control `.names`.
- **Part 3 (challenge)**: `across()` inside `mutate()` (reusing a function you write), plus `if_all()` / `if_any()` for filtering.

[Activity file](day21-activity.R){download="day21-activity.R"}

## Wrap-up & looking ahead {.scrollable .nostretch}

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- Iteration = the same action over many things; R gives you a lot for **free** (vectorization, faceting, `group_by`)
- **Functional programming**: hand a function to another function
- **`across(.cols, .fns, .names)`** repeats an action over columns
- `.cols` uses `select()` syntax; `everything()` and `where(is.numeric)`
- Pass the function name with **no `()`**; use `\(x) f(x, ...)` to add args
- Named list `.fns` for **multiple** functions; `{.col}_{.fn}` naming
- `if_any()` / `if_all()` bring the same idea to `filter()`

**Upcoming**

- **Wed Jun 17:** iterating over *files* and *list-columns* with `purrr::map()`
- **Reading before Wed:** [*R4DS* §26.4–26.5](https://psu.instructure.com/courses/2487468/assignments/18325377)
- **Homework 5**: due Friday June 19 at 11:59 p.m. on [Canvas](https://psu.instructure.com/courses/2487468/assignments/18343523)
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 26](https://r4ds.hadley.nz/iteration) · [`across()` reference](https://dplyr.tidyverse.org/reference/across.html)
:::
::::
