---
title: "Day 8 Activity: Plot-matching with `penguins`"
subtitle: "Thursday, May 28, 2026"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
```

Today's activity uses the `penguins` dataset (`palmerpenguins` package). Glimpse it before you start:

```{r}
glimpse(penguins)
```

---

# Instructions

Below are **four target plots**, generated for you in advance. Your job is to **recreate each one as closely as you can** using `ggplot2`, then write **one sentence** about what each plot reveals about the penguins.

You won't get pixel-perfect matches — themes, exact colors, and tiny details may differ. Aim to match:

- The frame (which variables are on x and y)
- The glyph (which geom)
- The aesthetic mappings (color, shape, facets, etc.)
- Any axis transformations or labels
- Anything else that's clearly *intentional* in the target

Look closely. A target plot with three colors and three facets is showing you something different than one with three colors and no facets.

---

## Target 1 — Bills, with species and a smoother

```{r target1, echo=FALSE, fig.width=7, fig.height=4, warning=FALSE, message=FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2, alpha = 0.7, na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, na.rm = TRUE) +
  labs(
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    title = "Bill dimensions by species",
    color = "Species"
  ) +
  theme_minimal()
```

```{r recreate-1}
# Your code:

```

**What does this plot reveal?**

[Write one sentence.]

---

## Target 2 — Body mass distribution by species and sex

```{r target2, echo=FALSE, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.6) +
  facet_grid(sex ~ species) +
  labs(
    x = "Body mass (g)",
    y = "Count",
    title = "Body mass distribution by species and sex",
    fill = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r recreate-2}
# Your code:

```

*Hint: the legend is hidden because the fill is redundant with the facet columns. You can use `show.legend = FALSE` on the geom, or `theme(legend.position = "none")` to do this.*

**What does this plot reveal?**

[Write one sentence.]

---

## Target 3 — Flipper length vs. body mass, with continuous color

```{r target3, echo=FALSE, fig.width=7, fig.height=4, warning=FALSE, message=FALSE}
ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, color = bill_length_mm)) +
  geom_point(size = 2.5, na.rm = TRUE) +
  scale_color_viridis_c(option = "plasma") +
  facet_wrap(~species) +
  labs(
    x = "Body mass (g)",
    y = "Flipper length (mm)",
    title = "Flipper length vs. body mass, colored by bill length",
    color = "Bill length (mm)"
  ) +
  theme_minimal()
```

```{r recreate-3}
# Your code:

```

*Hint: this plot encodes **four** variables. The continuous color scale is from the viridis package — see the slide on continuous color.*

**What does this plot reveal?**

[Write one sentence.]

---

## Target 4 — Counts of penguins by island and species

```{r target4, echo=FALSE, fig.width=7, fig.height=4, warning=FALSE, message=FALSE}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  labs(
    x = "Island",
    y = "Number of penguins",
    title = "Where each species lives",
    fill = "Species"
  ) +
  theme_minimal()
```

```{r recreate-4}
# Your code:

```

*Hint: pay attention to the position adjustment. Stacked, dodged, and filled bars all look different.*

**What does this plot reveal?**

[Write one sentence.]

---

# Wrap-up

Once you've matched all four, take a moment to reflect:

**Which target was hardest?** Was it the one with the most variables, or one with a subtler design choice?

[Write 1–2 sentences.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
