---
title: "Day 8 Activity: ggplot2 internals with `storms`"
subtitle: "Wednesday, May 27, 2026"
author: "Your Name Here"
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
execute:
  enabled: true
  freeze: auto
editor: source
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

We'll use the `storms` dataset that comes built into the tidyverse (via `dplyr`). It contains 6-hourly observations of every named Atlantic storm since 1975, from NOAA's HURDAT2 reanalysis.

```{r}
glimpse(storms)
```

Useful variables for today:

- `name`, `year`, `month`, `day`, `hour` — when the observation was taken
- `lat`, `long` — where the storm was
- `status` — `tropical depression`, `tropical storm`, `hurricane`, ...
- `category` — Saffir-Simpson scale (1–5), `NA` for non-hurricanes
- `wind` — maximum sustained wind speed, knots
- `pressure` — central pressure, millibars (lower = stronger)

---

# Part 1 — Guided warmup (~10 minutes)

Fill in the blanks in each chunk. Each one practices a concept from the slides.

## 1.1 — Mapping vs. setting

The chunk below colors every point the same shade of teal. **Without changing anything else,** modify the code so points are colored by `status` instead.

```{r}
#| eval: FALSE
ggplot(storms, aes(x = wind, y = pressure)) +
  geom_point(color = "#1c7293", alpha = 0.4)
```

```{r}
# Your fixed version:

```

In **one sentence**, what's the difference between putting `color = "#1c7293"` *inside* `aes()` versus *outside* it?

[Your answer here.]

## 1.2 — Local mappings

The plot below colors *both* the points *and* the smoothers by `status` (so you get one smoother per status). Modify it so **only the points are colored**, but the smoother is a single curve fit to all the data.

```{r}
#| eval: FALSE
ggplot(storms, aes(x = wind, y = pressure, color = status)) +
  geom_point(alpha = 0.2) +
  geom_smooth()
```

```{r}
# Your fixed version:

```

## 1.3 — The `group` aesthetic

Run this code first. It tries to draw the path of every storm in 2005, but it produces a single zigzagging line that connects unrelated storms together.

```{r}
#| eval: FALSE
ggplot(storms |> filter(year == 2005), aes(x = long, y = lat)) +
  geom_path()
```

Add **one aesthetic mapping** so each storm gets its own path. (Hint: which variable identifies a storm?)

```{r}
# Your fixed version:

```

## 1.4 — `after_stat()` for proportions

Make a bar chart showing the **proportion** (not count) of observations in each `status` category. You should not pre-compute the proportions with `dplyr` — let ggplot do it via `after_stat()`.

*Hint: you'll need `group = 1`. The slide on `after_stat()` explains why.*

```{r}
# Your code:

```

## 1.5 — Position adjustments

Run the code below. It stacks hurricane counts by category, within each month.

```{r}
#| eval: FALSE
hurricanes <- storms |> filter(status == "hurricane")

ggplot(hurricanes, aes(x = month, fill = factor(category))) +
  geom_bar()
```

Make **two** more versions:

(a) One where each bar is normalized so every month has the same height, and segments show the *proportion* of each category.

(b) One where the categories are placed *side by side* within each month, not stacked.

```{r}
# (a) position = "fill"

```

```{r}
# (b) position = "dodge"

```

In **one sentence**, which of the three versions (original stacked, fill, dodge) would you use to answer the question *"In which month is Category 5 most common as a share of all hurricanes?"*

[Your answer here.]

---

# Part 2 — Open exploration (~15 minutes)

**Pick ONE** of the three questions below. Build a plot that you think answers it well, then write 2–3 sentences explaining what you see.

You're encouraged to use:

- multiple geoms in layers
- local vs. global mappings
- a stat or position adjustment we covered today
- a thoughtful color or alpha choice

Don't try to use all of them — pick the ones that fit the question.

## Question A — Wind vs. pressure

It's well-known among meteorologists that storm wind speed and central pressure have a very tight relationship. Build a plot that **shows this relationship clearly** and lets us see how `status` (or `category`) fits in.

*Suggestion: layered geoms with a smoother on top, or color by category and an alpha tuned for the overplotting.*

## Question B — When is hurricane season?

Build a plot that shows how **hurricane activity varies through the year**. You could look at counts by month, intensity by month, or how the distribution of categories shifts as the season progresses.

*Suggestion: this is a job for `geom_bar()` with a position adjustment, or possibly a faceted version of something simpler.*

## Question C — One season's storm tracks

Pick a single year (say, 2005 — a famously bad year — or any other). Build a map of `lat` vs. `long` showing the **paths of every storm** that season. Highlight one or two notable storms.

*Suggestion: this needs the `group` aesthetic, and probably two `geom_path()` layers with different data.*

```{r}
# Your plot:

```

**Your 2–3 sentence interpretation:**

[Write here.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
