---
title: "Multivariable graphics · specifying graphs · axes & labels"
subtitle: "Thursday, May 28, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day8-multivariable-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day8-multivariable-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Thursday May 28]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Assignment**: [*DC* §8.1–8.3](https://psu.instructure.com/courses/2487468/assignments/18317726)

**Plan for today**

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 9](https://r4ds.hadley.nz/layers) · [*DC* Ch. 8](https://dtkaplan.github.io/DataComputingEbook/chap-graphics-commands.html#chap:graphics-commands) · [ggplot2 reference](https://ggplot2.tidyverse.org/reference)
:::

### 2. TODO: main segment (~25 min)

TODO: hands-on practice — students follow along.

```{r}
#| eval: false
# TODO: example to walk through
```

### 3. In-class activity (~20 min)

TODO: brief description, or link to a separate handout.

### 4. Wrap-up (~5 min)

- TODO: key takeaway
- TODO: preview next class

## Resources

- TODO: slides PDF
- TODO: in-class code .R file
- TODO: activity handout

## After class

- TODO: due tomorrow
- TODO: optional supplementary reading

::: {.callout-tip collapse="true"}
## Instructor notes

TODO: notes to self — pacing, common student questions, things to watch for.
:::
