---
title: "Day 13 Activity: Reshaping data"
subtitle: "Thursday, June 4, 2026"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
```

Today's verbs are `pivot_longer()` and `pivot_wider()`. We'll use a few small built-in tables plus `penguins`.

---

# Part 1 --- Tidy, wide, or narrow?

Look at `table1` and `table2` (both built into tidyr):

```{r}
table1
table2
```

**1a.** For each, describe what **one row** represents.

[table1: ... ] — [table2: ... ]

**1b.** Which one is **tidy** (each variable a column, each observation a row)? Which is the **narrower** of the two, and how can you tell?

[Your answer.]

---

# Part 2 --- pivot_longer() on a tiny table

Build this wide table by hand:

```{r build-wide}
bp_wide <- tribble(
  ~subject, ~before, ~after,
  "BHO",        160,    160,
  "GWB",        115,    135,
  "WJC",        105,    145
)
bp_wide
```

**2a. Predict first.** If you pivot `before` and `after` into a long form, how many **rows** will the result have? How many **columns**?

Prediction: [ ... rows / ... columns ]

**2b.** Now do it. Use `names_to = "when"` and `values_to = "sbp"`. Check against your prediction.

```{r longer-bp}
# Your code:

```

---

# Part 3 --- pivot_longer() on real data

The `billboard` data (in tidyr) has one column per chart week (`wk1`, `wk2`, ...).

**3a.** Pivot all the `wk` columns into `week` / `rank`, dropping the empty weeks. (Hint: `cols = starts_with("wk")`, `values_drop_na = TRUE`.)

```{r longer-billboard}
# Your code:

```

**3b.** Add `mutate(week = parse_number(week))` so `"wk3"` becomes `3`, then plot `rank` (y) against `week` (x), one faint line per `track`. Use `scale_y_reverse()` so rank 1 sits on top.

```{r billboard-plot}
# Your code:

```

**One sentence:** what does the plot show about how long songs stay on the charts?

[Your answer.]

---

# Part 4 --- pivot_wider() for comparison

Start from a **long** summary (this is the kind of table you built on Wednesday):

```{r long-summary, message=FALSE}
penguin_summary <- penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(avg_mass = mean(body_mass_g, na.rm = TRUE), .groups = "drop")
penguin_summary
```

**4a. Predict first.** If you widen this with `names_from = sex`, `values_from = avg_mass`, how many rows and columns will you get?

Prediction: [ ... rows / ... columns ]

**4b.** Do the pivot, then add a column `diff` = `male` − `female`.

```{r wider-penguins}
# Your code:

```

**Which species has the biggest mass gap between sexes?**

[Your answer.]

---

# Part 5 --- The round trip, and a trap

**5a.** Take your widened penguin table from Part 4 (before the `diff` column) and pivot it **back** to long form. Do you recover the original `penguin_summary`? (`pivot_longer()` and `pivot_wider()` are inverses.)

```{r round-trip}
# Your code:

```

**5b. Debug it.** Someone tries to make `bp_wide` longer like this and gets a strange result:

```{r too-narrow}
#| eval: false
bp_wide |>
  pivot_longer(
    cols = c(subject, before, after),
    names_to = "when",
    values_to = "sbp"
  )
```

Run it. Why is this "too narrow" / wrong? What single change to `cols` fixes it?

[Your explanation and fix.]

---

# Part 6 --- Choose the shape

For each goal, say whether you'd want the data **wide** or **narrow**, and which pivot gets you there. One sentence each.

1. Plot each country's TB cases over time, one line per country.
2. Compute, for each country, the change in cases from 1999 to 2000 as a single new column.
3. Group by year and sum the cases across all countries.

[1: ... ] — [2: ... ] — [3: ... ]

---

# Wrap-up

**One-sentence reflection:** pivoting is the first thing this week that changed the *shape* of the table rather than its contents, what clicked, and what's still fuzzy?

[Write one sentence.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
