---
title: "Day 14 Activity: Get Git working (weekend setup)"
subtitle: "Friday, June 5, 2026 · due 11:59 p.m. Tuesday, June 9"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

By **Tuesday June 9** your project team needs a shared GitHub repo that **every member can push to.**

Work through the steps in order. Each links to the matching chapter of [**Happy Git with R**](https://happygitwithr.com/) (Jenny Bryan); read its instructions there and use this page as the checklist. Part 5 covers the common ones, and feel free to come to office hours if you get stuck.

::: {.callout-note}
We use **RStudio's built-in Git integration** with **HTTPS + a personal access token (PAT)**. That matches *DC* Ch. 9 and Happy Git, and keeps everything inside RStudio.
:::

---

# Part 1 --- Accounts and installs (solo)

- [ ] **Register a GitHub account** (if you don't already have one): [happygitwithr.com/github-acct](https://happygitwithr.com/github-acct). Pick a username that's professional-ish; you may reuse it.
- [ ] **R and RStudio up to date**: [install-r-rstudio](https://happygitwithr.com/install-r-rstudio). You set these up in week 1; just confirm they're current.
- [ ] **Install Git**: [install-git](https://happygitwithr.com/install-git). (Windows: Git for Windows. macOS: easiest is to run `git --version` in the Terminal and accept the install prompt.)
- [ ] **Introduce yourself to Git**: [hello-git](https://happygitwithr.com/hello-git). Set your name and the email tied to your GitHub account. This is easiest from the R console:

```{r intro-git}
#| eval: false
# install.packages("usethis")   # once, if you don't have it
library(usethis)
use_git_config(
  user.name  = "Your Name",          # your name, doesn't have to be your username
  user.email = "you@example.com"      # the SAME email as your GitHub account
)
```

---

# Part 2 --- Connect RStudio, Git, and GitHub (solo)

- [ ] **Create a personal access token (PAT)**: [https-pat](https://happygitwithr.com/https-pat). From the R console:

```{r pat}
#| eval: false
# install.packages(c("usethis", "gitcreds"))   # run this exactly once
usethis::create_github_token()   # opens GitHub; generate the token, COPY it
gitcreds::gitcreds_set()         # paste the token when prompted
```

::: {.callout-warning}
Treat the token like a password. You should copy it the moment you create it, because GitHub won't show it again. If you happen to lose it, just make a new one.
:::

- [ ] **Confirm the connection works**: [push-pull-github](https://happygitwithr.com/push-pull-github).
- [ ] **Confirm RStudio sees Git**: [rstudio-git-github](https://happygitwithr.com/rstudio-git-github) and [rstudio-see-git](https://happygitwithr.com/rstudio-see-git). Check *Tools > Global Options > Git/SVN* shows a Git executable.

---

# Part 3 --- Test repo (solo)

Follow [new-github-first](https://happygitwithr.com/new-github-first):

- [ ] On **GitHub**, create a new repository (check "Add a README").
- [ ] Copy its URL (the green **Code** button, then HTTPS).
- [ ] In **RStudio**: *File > New Project > Version Control > Git*, paste the URL, create.
- [ ] Open `README.md`, add a line like "Test repo for STAT 184", and **save**.
- [ ] In the **Git tab** (top-right pane): tick the box to **stage** `README.md`, click **Commit**, write a message, then **Push** (up arrow).
- [ ] Refresh the repo page on GitHub. **Your line is there.**

If your change shows up on GitHub, your setup is done. The rest of the term is just repeating this loop.

---

# Part 4 --- The team repo (together, asynchronously over the weekend)

Coordinate with your team over the weekend.

**One person (the repo owner):**

- [ ] Create the team repo on GitHub (with a README). Name it something like `stat184-project-teamNN`.
- [ ] *Settings > Collaborators*, then add each teammate by GitHub username.

**Everyone else:**

- [ ] Accept the email/GitHub invitation.
- [ ] Clone the repo into RStudio (*New Project > Version Control > Git*, paste the URL).

**Each member, including the owner:**

- [ ] **Pull** first (down arrow) to get everyone's latest.
- [ ] Add your name on its own line in the README, save.
- [ ] Stage, Commit (message like "Add Charles to README"), then **Push**.

- [ ] **Verify:** on GitHub, the README lists **every** member's name, and the repo's commit history shows **at least one commit per person.**

::: {.callout-tip}
If two of you edit the README at the same time, you may hit a **merge conflict**, that's fine, it's good practice. See Part 5, and remember the rule: **pull before you push.**
:::

---

# Part 5 --- When it goes wrong (troubleshooting)

You will hit at least one of these. None of them mean you broke anything.

| Symptom | What's happening | What to do |
|---|---|---|
| **Push rejected** | A teammate pushed since your last pull | **Pull**, then push again |
| **Merge conflict** | Two people changed the same lines | Open the file, keep the right lines, remove the `<<<<`/`====`/`>>>>` markers, commit |
| Changes don't appear in Git tab | Wrong RStudio **Project** is open | Switch to the Project tied to that repo |
| "File too large" (>50 MB) | A big data file got staged | Don't commit it; add it to `.gitignore` |
| Auth fails on push | PAT missing/expired | Redo `gitcreds::gitcreds_set()` (Part 2) |

: Common setup snags. {tbl-colwidths="[30,38,32]"}

Resources:

- Happy Git's [**troubleshooting**](https://happygitwithr.com/troubleshooting) chapter.
- The ["**Burn it all down**"](https://happygitwithr.com/burn) move: delete the broken local folder and **re-clone** a fresh copy from GitHub. 

Feel free to come to office hours on Monday if you get stuck. 

---

# Turn in

Submit on Canvas:

1. The **URL of your team repo**.
2. Confirmation that **all members have pushed** (the easiest proof is a screenshot of the repo's commit history or contributors list).

**Due 11:59 p.m. Tuesday, June 9.**
