---
title: "Putting it together: from messy data to a summary table"
subtitle: "Wednesday, June 3, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day12-integration-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day12-integration-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(nycflights13)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Wednesday June 3]{style="font-size: 1.3em;"}

**Before class**

- **No new reading**

**Plan for today**

- No new verbs. Today is about **combining and sequencing** them
- The workflow: question --\> target table --\> plan --\> build --\> **check**
- Working on a genuinely messy dataset (`flights`)
- Handling missing values, small groups, wrong order

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*DC* Ch. 7](https://dtkaplan.github.io/DataComputingEbook/chap-wrangling.html#chap:wrangling) · [*DC* Ch. 10](https://dtkaplan.github.io/DataComputingEbook/chap-more-data-verbs.html#chap:more-data-verbs) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::

## What you already have

**Single-table toolkit**:

| Touches     | Verbs                       |
|-------------|-----------------------------|
| **Rows**    | `filter()`, `arrange()`     |
| **Columns** | `select()`, `mutate()`      |
| **Groups**  | `group_by()`, `summarise()` |

: Six verbs, plus `count()` and the pipe.

::: {.callout-tip style="font-size: 1.1em;"}
Today we will practice *choosing which verb, in what order*, and *checking that the answer is trustworthy and rigorous.*
:::

## Workflow {.scrollable .nostretch}

1.  **State the question** in plain English.
2.  **Sketch the target table:** what does one row represent? what columns?
3.  **Plan the verbs** as English sentences, writing one sentence per step.
4.  **Build the pipeline** one verb at a time, looking at the output each time.
5.  **Check it.** Does the count make sense? Are missing values handled? Does the result match your intuition/pass a sanity check?

::: callout-note
Make sure not to skip Step 5. A pipeline that *runs* is not the same as a pipeline that's *correct*.
:::

# A messier dataset {background-color="#001E44"}

## `flights` {.scrollable .nostretch}

You saw `flights` on HW2 and Friday; it's also the running example in R4DS Ch. 3. It contains data on every flight out of NYC area airports in 2013:

```{r}
#| echo: TRUE
glimpse(flights)
```

## Dirty data {.scrollable .nostretch}

This is a common example of what a real dataset often looks like immediately after you load it into R, before you've done any data cleaning:

- **336,776 rows, 19 columns**: `select()` and `n()` are much more relevant here compared to a smaller dataset like `penguins`.
- **Meaningful missing values:** cancelled flights have `NA` for `dep_delay` and `arr_delay`.
- **Codes, rather than long-form labels:** here, `carrier` is `"UA"`, not `"United"`. (Turning codes into names requires a *join*, which is later, so today we'll stick to the codes.)

```{r}
#| echo: TRUE
flights |>
  summarise(
    n           = n(),
    missing_dep = sum(is.na(dep_delay))
  )
```

::: callout-warning
The above result tells us that there are 8,000 rows with a missing departure delay. Every average we compute has to decide what to do with them, or it silently returns `NA`.
:::

# Worked example 1 {background-color="#001E44"}

## Q: which months had the worst delays? {.scrollable .nostretch}

**Plan:**

- We want one row per **month**, with its average departure delay and a count.
- *Group by* month, *then summarise* the mean and `n()`, *then arrange* worst-first.

**Target table:** `month | avg_delay | n`, which should have twelve rows.

```{r}
#| echo: TRUE
flights |>
  group_by(month) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  ) |>
  arrange(desc(avg_delay))
```

## What just happened {.scrollable .nostretch}

We used **three** verbs in sequence, and two safety habits from this week:

- `na.rm = TRUE` so the cancelled flights don't give `NA`s for the means.
- `n = n()` so we can see each average rests on around 25,000 flights at least, rather than a small number.

::: callout-tip
Summer months (6, 7) and December rise to the top, which matches what you'd expect from thunderstorms and holiday travel. This is a good place to think about common sense/sanity checking; if January had come out worst by a mile (no thunderstorms and little holiday travel), it might suggest something is wrong with your code. At the very least, it's a signal to double check your code/investigate further.
:::

# Worked example 2 {background-color="#001E44"}

## Q: worst JFK summer destinations? {.scrollable .nostretch}

**Question:** for flights leaving **JFK in summer**, which **destination** had the worst average arrival delay, among destinations with **enough flights to trust the result**? Let's set an arbitrary cutoff at 100 flights for simplicity.

**Plan:**

- *Filter* to JFK, summer months.
- *Group by* destination, *summarise* mean arrival delay + `n()`.
- *Filter again* to drop tiny destinations.
- *Arrange* worst-first.

## Execution {.scrollable .nostretch}

```{r}
#| echo: TRUE
flights |>
  filter(origin == "JFK", month %in% c(6, 7, 8)) |>
  group_by(dest) |>
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    n             = n(),
    .groups       = "drop" # <-- removes grouping metadata from your data frame
  ) |>
  filter(n >= 100) |>
  arrange(desc(avg_arr_delay))
```

::: callout-note
Notice that `filter()` appears **twice**, operating once on the raw rows (which flights), and once on the summary table (which destinations are "big enough").
:::

## The necessity of the second `filter ()`.scrollable .nostretch}

Without `filter(n >= 100)`, the "worst destination" could be whichever one had a single very-late flight (isn't necessarily a problem with this dataset, but we got lucky). **A mean from a group of 3 is not very reliable.** 

- If we gained more samples, the answer is much more likely to shift significantly in the case of a small sample size.

It's a good idea to have verification steps built-in to all of your pipelines when possible; always including the count column is how you *catch* the problem before it ends up in a report that you have to retract later.

# Worked example 3 {background-color="#001E44"}

## Q: do long hauls leave later? {.scrollable .nostretch}

Now we can leverage `mutate()` before piping our transformed data into a plot.

**English plan:** *mutate* a long/short label from `distance`, *group by* it, *summarise* the average delay, *then plot* the summary table.

```{r}
#| echo: TRUE
flights |>
  mutate(haul = ifelse(distance > 1000, "long", "short")) |>
  group_by(haul) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )
```

## From summary table to plot {.scrollable .nostretch}

Because the summary is just a (tiny) data frame, it's already glyph-ready, so we can pipe it into `ggplot()`:

```{r}
#| echo: TRUE
#| fig-height: 3
flights |>
  mutate(haul = ifelse(distance > 1000, "long", "short")) |>
  group_by(haul) |>
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = haul, y = avg_delay)) +
  geom_col()
```

# Getting it right {background-color="#001E44"}

## Order matters {.scrollable .nostretch}

The same verbs in a different order answer different questions:

::::: columns
::: {.column width="50%"}
**Filter, *then* summarise**

"Average delay *of JFK flights.*" (This is more computationally efficient)

```{r}
#| echo: TRUE
#| eval: FALSE
flights |>
  filter(origin == "JFK") |>
  summarise(
    avg = mean(dep_delay,
               na.rm = TRUE)
  )
```
:::

::: {.column width="50%"}
**Summarise, *then* filter**

"Of the per-origin averages, keep JFK's."

```{r}
#| echo: TRUE
#| eval: FALSE
flights |>
  group_by(origin) |>
  summarise(
    avg = mean(dep_delay,
               na.rm = TRUE)
  ) |>
  filter(origin == "JFK")
```
:::
:::::

Both are valid, and both give the same number in this case. However, they won't always be. The trick is knowing **which question you're actually asking** before you write it.

## A case when the output changes with order

```{r}
#| echo: TRUE
flights |>
  filter(dep_delay > 60) |> 
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE))

flights |>
  summarise(avg_delay = mean(arr_delay, na.rm = TRUE)) |>
  filter(avg_delay > 60)
```

## The three checks {.scrollable .nostretch}

Before trusting any wrangled result, run this checklist:

| Check                                | The verb/habit           |
|--------------------------------------|--------------------------|
| Did missing values get handled?      | `na.rm`, `!is.na()`      |
| Is each number based on enough rows? | `n = n()`                |
| Does the answer pass a smell test?   | compare to what you know |

: Your verification checklist.

::: callout-warning
A pipeline that runs without error has cleared the *lowest* bar; that doesn't *necessarily* mean everything is fine.
:::

## What we did today {.scrollable .nostretch}

- **No new verbs**: just the six from Monday and Tuesday, sequenced
- **The workflow:** question --> target table --> English plan --> build incrementally --> check
- **Real mess:** missing delays, code-not-name columns, tiny groups
- **`filter()` in two roles:** subsetting raw rows *and* trimming a summary table
- **Order matters:** filter-then-summarise =/= summarise-then-filter
- **The full arc:** raw rows --> glyph-ready summary --> plot

# Activity {background-color="#7C0A02"}

## Activity: a flights investigation

Roughly 25 minutes. Individually or with a neighbor.

- [Activity Canvas Link](https://psu.instructure.com/courses/2487468/assignments/)

**Format:** you'll answer a series of questions about `flights` by building pipelines, but each one starts with writing the **plan in words first**. You'll compare two pipelines that look alike but answer different questions, catch a silently-dropped `NA`, and finish with one open-ended question where you choose the verbs and **report your sanity checks**.

**To turn in:** save your `.qmd` and rendered HTML.

## Wrap-up & looking ahead

**Tomorrow (Thursday):** new reading and new content, reshaping data between **wide and narrow** with `pivot_longer()` and `pivot_wider()`.

::::: columns
::: {.column width="55%"}
**Upcoming Deadlines**

- **Thu 6/4, before class**: [*DC* §12.1](https://psu.instructure.com/courses/2487468/assignments/18325371)
- **Thu 6/4, 11:59 p.m.**: [Project: Team formation](https://psu.instructure.com/courses/2487468/discussion_topics/17435844)
- **Fri 6/5, 11:59 p.m.**: [Homework 3: Data Wrangling & Reshaping ](https://psu.instructure.com/courses/2487468/assignments/18327019)
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478\@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228\@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*DC* Ch. 10](https://dtkaplan.github.io/DataComputingEbook/chap-more-data-verbs.html#chap:more-data-verbs) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::
:::::
