---
title: "Putting it together: messy data → summary table"
subtitle: "Wednesday, June 3, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day12-integration-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day12-integration-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(nycflights13)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Wednesday June 3]{style="font-size: 1.3em;"}

**Before class**

- **No new reading** — bring the six verbs from Monday and Tuesday

**Plan for today**

- No new verbs. Today is about **combining and sequencing** them
- The workflow: question → target table → plan → build → **check**
- Working on a genuinely messy dataset (`flights`)
- Not getting fooled: missing values, tiny groups, wrong order

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*DC* Ch. 7](https://dtkaplan.github.io/DataComputingEbook/chap-wrangling.html#chap:wrangling) · [*DC* Ch. 10](https://dtkaplan.github.io/DataComputingEbook/chap-more-data-verbs.html#chap:more-data-verbs) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::

## What you already have

Two days in, you have the **whole single-table toolkit**:

| Touches... | Verbs |
|---|---|
| **Rows** | `filter()`, `arrange()` |
| **Columns** | `select()`, `mutate()` |
| **Groups** | `group_by()`, `summarise()` |

: Six verbs, plus `count()` and the pipe. {tbl-colwidths="[30,70]"}

::: {.callout-tip style="font-size: 1.1em;"}
Today there is **nothing new to memorize.** The skill we're building is *choosing which verb, in what order*, and *checking that the answer is real.*
:::

## The workflow {.scrollable .nostretch}

This is Monday's habit (*DC* §7.1), now run end to end:

1. **State the question** in plain English.
2. **Sketch the target table:** what does one row represent? what columns?
3. **Plan the verbs** as English sentences, one per step.
4. **Build the pipeline** one verb at a time, looking at the output each time.
5. **Check it.** Does the count make sense? Are missing values handled? Does the number pass a smell test?

::: {.callout-note}
Step 5 is the one people skip, and it's the one Friday's AI lesson was about. A pipeline that *runs* is not the same as a pipeline that's *right*.
:::

# A messier dataset {background-color="#001E44"}

## Meet `flights` {.scrollable .nostretch}

You saw `flights` on HW2 and Friday; it's also the running example in R4DS Ch. 3. Every flight out of NYC in 2013:

```{r}
#| echo: TRUE
glimpse(flights)
```

## Where the mess hides {.scrollable .nostretch}

This is what real data looks like after `read_csv("something.csv")`:

- **336,776 rows, 19 columns** — `select()` and `n()` earn their keep here in a way they didn't on `penguins`.
- **Missing values that mean something:** cancelled flights have `NA` for `dep_delay` and `arr_delay`.
- **Codes, not labels:** `carrier` is `"UA"`, not `"United"`. (Turning codes into names needs a *join*, which is later, so today we'll live with the codes.)

```{r}
#| echo: TRUE
flights |>
  summarise(
    n          = n(),
    missing_dep = sum(is.na(dep_delay))
  )
```

::: {.callout-warning}
Over 8,000 rows have a missing departure delay. Every average we compute has to decide what to do with them, or it silently returns `NA`.
:::

# Worked example 1 {background-color="#001E44"}

## Q: which months had the worst delays? {.scrollable .nostretch}

**English plan:**

- One row per **month**, with its average departure delay and a count.
- *Group by* month, *then summarise* the mean and `n()`, *then arrange* worst-first.

**Target table:** `month | avg_delay | n` — twelve rows.

```{r}
#| echo: TRUE
flights |>
  group_by(month) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  ) |>
  arrange(desc(avg_delay))
```

## What just happened {.scrollable .nostretch}

We used **three** verbs in sequence, and two safety habits from this week:

- `na.rm = TRUE` so the cancelled flights don't poison the mean *(Monday)*
- `n = n()` so we can see each average rests on ~25,000 flights, not a handful *(Monday)*

::: {.callout-tip}
Summer months (6, 7) and December rise to the top — which matches what you'd expect from thunderstorms and holiday travel. **That agreement is the smell test passing.** If January had come out worst by a mile, you'd go looking for a bug.
:::

# Worked example 2 {background-color="#001E44"}

## Q: worst JFK summer destinations? {.scrollable .nostretch}

**Question:** for flights leaving **JFK in summer**, which **destination** had the worst average arrival delay, among destinations with **enough flights to trust**?

**English plan:**

- *Filter* to JFK, summer months.
- *Group by* destination, *summarise* mean arrival delay + `n()`.
- *Filter again* to drop tiny destinations.
- *Arrange* worst-first.

## Build it one verb at a time {.scrollable .nostretch}

```{r}
#| echo: TRUE
flights |>
  filter(origin == "JFK", month %in% c(6, 7, 8)) |>
  group_by(dest) |>
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    n             = n(),
    .groups       = "drop"
  ) |>
  filter(n >= 100) |>
  arrange(desc(avg_arr_delay))
```

::: {.callout-note}
Notice `filter()` appears **twice**, and it's the *same verb* doing two different jobs: once on the raw rows (which flights), once on the summary table (which destinations are big enough to believe). Same tool, different point in the pipeline.
:::

## Why the second filter matters {.scrollable .nostretch}

Without `filter(n >= 100)`, the "worst destination" is whichever one had a single very-late flight. That's exactly Monday's lesson: **a mean from a group of 3 is not a fact.**

```{r}
#| echo: TRUE
#| eval: FALSE
# Drop the n >= 100 line and the top of the list is full of
# destinations with n = 1 or n = 2. Impressive-looking, meaningless.
```

This is also a verification move: the count column is how you *catch* the problem before it ends up in a report.

# Worked example 3 {background-color="#001E44"}

## Q: do long hauls leave later? {.scrollable .nostretch}

Now bring in `mutate()` and end with a plot.

**English plan:** *mutate* a long/short label from `distance`, *group by* it, *summarise* the average delay, *then plot* the summary table.

```{r}
#| echo: TRUE
flights |>
  mutate(haul = ifelse(distance > 1000, "long", "short")) |>
  group_by(haul) |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )
```

## From summary table to plot {.scrollable .nostretch}

Because the summary is just a (tiny) data frame, it's already glyph-ready, pipe it into `ggplot()`, switching `|>` to `+`:

```{r}
#| echo: TRUE
#| fig-height: 3
flights |>
  mutate(haul = ifelse(distance > 1000, "long", "short")) |>
  group_by(haul) |>
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = haul, y = avg_delay)) +
  geom_col()
```

This is the full arc of the week: **raw rows → wrangle to a glyph-ready table → graph it.**

# Getting it right {background-color="#001E44"}

## Order matters {.scrollable .nostretch}

The same verbs in a different order answer different questions:

::: {.columns}
::: {.column width="50%"}
**Filter, *then* summarise**

"Average delay *of JFK flights.*"

```{r}
#| echo: TRUE
#| eval: FALSE
flights |>
  filter(origin == "JFK") |>
  summarise(
    avg = mean(dep_delay,
               na.rm = TRUE)
  )
```
:::

::: {.column width="50%"}
**Summarise, *then* filter**

"Of the per-origin averages, keep JFK's."

```{r}
#| echo: TRUE
#| eval: FALSE
flights |>
  group_by(origin) |>
  summarise(
    avg = mean(dep_delay,
               na.rm = TRUE)
  ) |>
  filter(origin == "JFK")
```
:::
:::

Both are valid. The trick is knowing **which question you're actually asking** before you write it.

## The three checks {.scrollable .nostretch}

Before you believe any wrangled result, run the checklist that's been building all week:

| Check | The verb/habit | First seen |
|---|---|---|
| Did missing values get handled? | `na.rm`, `!is.na()` | Mon / Tue |
| Is each number based on enough rows? | `n = n()` | Monday |
| Does the answer pass a smell test? | compare to what you know | Friday |

: Your verification checklist. {tbl-colwidths="[42,33,25]"}

::: {.callout-warning}
A pipeline that runs without error has cleared the *lowest* bar. These three checks are what separate "it ran" from "it's right."
:::

## What we did today {.scrollable .nostretch}

- **No new verbs** — just the six from Monday and Tuesday, sequenced
- **The workflow:** question → target table → English plan → build incrementally → check
- **Real mess:** missing delays, code-not-name columns, tiny groups
- **`filter()` in two roles:** subsetting raw rows *and* trimming a summary table
- **Order matters:** filter-then-summarise ≠ summarise-then-filter
- **The full arc:** raw rows → glyph-ready summary → plot

# Activity {background-color="#7C0A02"}

## Activity: a flights investigation

Roughly 25 minutes. Individually or with a neighbor.

- [Activity Canvas Link](https://psu.instructure.com/courses/2487468/assignments/)

**Format:** you'll answer a series of questions about `flights` by building pipelines, but each one starts with writing the **English plan first**. You'll compare two pipelines that look alike but answer different questions, catch a silently-dropped `NA`, and finish with one open-ended question where you choose the verbs and **report your sanity checks**.

**To turn in:** save your `.qmd` and rendered HTML.

## Wrap-up & looking ahead

**Tomorrow (Thursday):** new reading and new content, reshaping data between **wide and narrow** with `pivot_longer()` and `pivot_wider()`.

:::: {.columns}
::: {.column width="55%"}
**Upcoming Deadlines**

- **Thu 6/4, before class**: [*DC* §12.1](https://psu.instructure.com/courses/2487468/assignments/18325371)
- **Thu 6/4, 11:59 p.m.**: [Project: Team formation](https://psu.instructure.com/courses/2487468/assignments/)
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*DC* Ch. 10](https://dtkaplan.github.io/DataComputingEbook/chap-more-data-verbs.html#chap:more-data-verbs) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::
::::
