---
title: "Wrangling concepts · reduction & transformation · group_by"
subtitle: "Monday, June 1, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day10-wrangling-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day10-wrangling-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(palmerpenguins)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Monday June 1]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Assignment**: [*DC* §7.1–7.7](https://psu.instructure.com/courses/2487468/assignments/18323235)

**Plan for today**

- The wrangling *mindset*: start from the table you want
- The three function families: **reduction**, **transformation**, **data verb**
- First two data verbs: `summarize()` and `group_by()`
- The `na.rm` trap
- The usefulness of `n()` in summaries

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*DC* Ch. 7](https://dtkaplan.github.io/DataComputingEbook/chap-wrangling.html#chap:wrangling) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::

## Where we are

For two weeks we've been **making plots**. But that assumed the data was already in the right shape, with one row per glyph and one column per aesthetic.

::: {.callout-tip style="font-size: 1.1em;"}
Real data almost never arrives glyph-ready. This week is about **getting it there**.
:::

Today we will build the **mental model** that every wrangling function hangs off of.

## Reading the pipe

We've learned `|>`. Read it as the word **"then."**

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |>           # take penguins, then
  group_by(species) |> #   group it, then
  summarize(n = n())   #   summarize it
```

::: {.callout-note}
The *DC* reading writes the pipe as `%>%`. `|>` and `%>%` behave identically, so read them as the same word. We'll use `|>`.
:::

# Wrangling mindset {background-color="#001E44"}

## Start from the table you want

*DC* §7.1 gives the single most useful habit in this whole unit:

> Before any coding, sketch the **glyph-ready table** you need, then plan the steps to get there in **plain English**.

For "how does average body mass differ by species?", the table you *wish* you had is:

| species | avg_mass |
|---|---|
| Adelie | ... |
| Chinstrap | ... |
| Gentoo | ... |

Here we have one row per species and one column for the number you want. **The case is a species, not a penguin.** That shift in what one row represents is from `summarize()` and `group_by()`. 

## Plan in English first

*DC*'s smoking example plans the wrangling as four English sentences *before* any code:

1. Make an age-group variable from age.
2. Drop people under 20.
3. Split into groups by age-decade and sex.
4. For each group, count people and count smokers; divide.

Each sentence maps to **one operation**. Today we learn the vocabulary for sentences like #3 (*split into groups*) and #4 (*for each group, count / average*).

# A framework for wrangling {background-color="#001E44"}

## Three kinds of building blocks

*DC* §7.2 boils all of wrangling down to three shapes of data...

- A **data frame**: rows (cases) and columns (variables)
- A **variable**: one column
- A **scalar**: a single value

and three families of functions, sorted by **what goes in and what comes out**:

| Family | Input → Output | Examples |
|---|---|---|
| **Reduction** | variable → scalar | `mean()`, `sum()`, `n()`, `median()`, `sd()`, `n_distinct()` |
| **Transformation** | variable(s) → variable | `x / y`, `log10()`, `round()`, `ifelse()`, `<`, `==`, `%in%` |
| **Data verb** | data frame → data frame | `summarize()`, `group_by()`, `filter()`, `mutate()`, ... |

: {tbl-colwidths="[22,33,45]"}

## Reduction functions: many → one

A **reduction** takes a whole column and crushes it down to a single number (*DC* §7.3).

```{r}
#| echo: TRUE
mean(penguins$body_mass_g, na.rm = TRUE)  # one number out
```

We've used many of these before: `mean()`, `sum()`, `median()`, `sd()`, `min()`, `max()`. Here are two new ones:

- `n()` — how many cases (takes **no** arguments)
- `n_distinct()` — how many *different* values

::: {.callout-note}
`n()` doesn't look at any column, it simply counts rows. It only works *inside* a data verb like `summarize()`, not on its own.
:::

## Transformation functions: column to column

A **transformation** runs once *per row* and hands back a column the same length as the input (*DC* §7.4).

```{r}
#| echo: TRUE
head(penguins$body_mass_g / 1000)   # grams to kg, one value per row
```

Comparisons are transformations too, they return a column of `TRUE`/`FALSE`:

```{r}
#| echo: TRUE
head(penguins$body_mass_g > 4000)
```

::: {.callout-tip}
**Reduction vs. transformation in one question:** does it give back *one* number, or *one per row*? `mean()` to one. `round()` to one per row.
:::

## Data verbs: data frame to data frame

A **data verb** takes a whole data frame and returns a whole (new) data frame (*DC* §7.5). Here are three rules every dplyr verb obeys:

1. The **first argument is always a data frame**.
2. The other arguments name **columns**, unquoted.
3. The **output is always a new data frame**; the input is never modified.

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |> summarize(avg = mean(body_mass_g, na.rm = TRUE))
# penguins itself is unchanged; you get a NEW data frame back
```

::: {.callout-warning}
Because verbs never modify their input, `penguins |> summarize(...)` on its own just **prints**. To keep a result, assign it: `out <- penguins |> summarize(...)`.
:::

## Each step = one verb + helpers

> Each step in data wrangling involves a **data verb** and one or more **reduction or transformation functions.**

The verb is the *action on the table*; the reduction/transformation is the *detail of what to compute*.

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |>
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
#         └─ data verb ┘ └─ reduction function ┘
```

# summarize() {background-color="#001E44"}

## summarize(): collapse a table to a summary

`summarize()` turns many rows into **one row**, using reduction functions.

```{r}
#| echo: TRUE
penguins |>
  summarize(
    n        = n(),
    avg_mass = mean(body_mass_g, na.rm = TRUE),
    max_mass = max(body_mass_g, na.rm = TRUE)
  )
```

Things to notice:

- Each **named argument** becomes a **column** in the output.
- The output is a **data frame** — one row here, but still a data frame.

## The `na.rm` trap

Watch what happens without `na.rm = TRUE`:

```{r}
#| echo: TRUE
penguins |>
  summarize(avg_mass = mean(body_mass_g))
```

One missing value anywhere in the column makes the **whole answer** `NA`. The reduction can't average around a hole it doesn't know how to fill.

::: {.callout-warning}
This is the AI failure we dissected on Friday: code that runs, looks fine, and quietly returns `NA`. The fix is `na.rm = TRUE`, but the *skill* is noticing the `NA` in the first place.
:::

# group_by() {background-color="#001E44"}

## group_by(): summarize, but per group

`group_by()` doesn't compute anything itself. It **tags** the data frame so the *next* verb runs once per group.

```{r}
#| echo: TRUE
#| message: FALSE
penguins |>
  group_by(species) |>
  summarize(
    n        = n(),
    avg_mass = mean(body_mass_g, na.rm = TRUE)
  )
```

Same `summarize()` as before, now **one row per species** instead of one row total.

::: {.callout-note}
`group_by()` should (almost) always be followed by another verb. On its own it just attaches an invisible label, the work happens in the verb after it.
:::

## Grouping by more than one variable {.scrollable .nostretch}

List several columns to get one row per **combination**:

```{r}
#| echo: TRUE
#| message: FALSE
penguins |>
  group_by(species, sex) |>
  summarize(
    n        = n(),
    avg_mass = mean(body_mass_g, na.rm = TRUE)
  )
```

Two things worth a pause:

- There's an `NA` row for `sex` indicating some penguins have missing sex. `group_by()` keeps them as their own group. (Dropping them needs `filter()`.)
- You may see a message about the output being *"regrouped"*. This is harmless for us, and we can add `.groups = "drop"` to silence it.

## How `group_by()` changes the case {.scrollable .nostretch}

Wrangling changes **what one row means.**

| Pipeline | One row = |
|---|---|
| `penguins` | one penguin |
| `... |> summarize(n = n())` | the whole dataset |
| `... |> group_by(species) |> summarize(...)` | one species |
| `... |> group_by(species, sex) |> summarize(...)` | one species × sex combo |

: `group_by()` resets the meaning of a case. {tbl-colwidths="[60,40]"}

::: {.callout-tip}
Before you run a `group_by() |> summarize()`, predict **how many rows** the result has. It's the number of distinct group combinations. If your prediction is wrong, you've learned something about your data.
:::

# Trust, but verify {background-color="#001E44"}

## Always include an `n()`

Average mass by the finest grouping:

```{r}
#| echo: TRUE
#| message: FALSE
penguins |>
  group_by(species, island, sex) |>
  summarize(
    n        = n(),
    avg_mass = mean(body_mass_g, na.rm = TRUE),
    .groups  = "drop"
  )
```

Some groups have a handful of birds, others have dozens. **A mean from a group of 3 is not the same kind of fact as a mean from a group of 60.**

## Good habit

::: {.callout-warning}
Whenever you compute a group summary, **add `n = n()`.** A surprising-looking average is very often just a tiny group. Without the count, you can't tell a real effect from noise.
:::

This is also a verification move, exactly like Friday: the number that *looks* like an answer might be resting on almost no data.

# Tomorrow {background-color="#001E44"}

## What's next

Today: the framework + `summarize()` + `group_by()`. That's already enough to answer a lot of "how many / what's the average, by group" questions.

Tomorrow (*DC* Ch. 10) we add the four verbs:

- `filter()` — keep some **rows** (e.g. drop the `NA` sexes)
- `select()` — keep some **columns**
- `mutate()` — add a **new column** (this is where transformations live)
- `arrange()` — **reorder** rows

With the two we covered today, these handle "take a messy CSV and produce a useful summary."

## What we covered today

- **The mindset:** sketch the glyph-ready table, plan in English, one sentence per operation
- **The framework:** reduction (variable→scalar), transformation (variable→variable), data verb (frame→frame)
- **`summarize()`:** many rows to one; named args become columns; output is a data frame
- **`na.rm = TRUE`:** one `NA` poisons a whole reduction; spotting it is the skill
- **`group_by()`:** tags the frame so the next verb runs per group; changes what a "case" means
- **`n()`:** include it in every summary so you never trust a mean from three rows

# Activity {background-color="#7C0A02"}

## Activity: summarising the penguins

Roughly 25 minutes. Individually or with a neighbor.

- [Activity Canvas Link](https://psu.instructure.com/courses/2487468/assignments/)

You'll classify functions into the three families, build up `summarize()` and `group_by()` pipelines on `penguins`, deliberately trip the `na.rm` trap, and practice predicting the **shape** of a grouped result before you run it.

**To turn in:** save your `.qmd` and rendered HTML.

## Wrap-up & looking ahead

**Tomorrow (Tuesday):** the row and column verbs, `filter`, `select`, `mutate`, `arrange`, the rest of the dplyr toolkit.

:::: {.columns}
::: {.column width="55%"}
**Upcoming Deadlines**

- **Tue 6/2, before class**: [*DC* §10.1–10.6](https://psu.instructure.com/courses/2487468/assignments/18325370)
- **Thu 6/4, 11:59 p.m.**: [Project: Team formation](https://psu.instructure.com/courses/2487468/assignments/)
- **Fri 6/5, 11:59 p.m.**: [Homework 3: Data Wrangling & Reshaping](https://psu.instructure.com/courses/2487468/assignments/18327019)
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*DC* Ch. 7](https://dtkaplan.github.io/DataComputingEbook/chap-wrangling.html#chap:wrangling) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::
::::
