---
title: "Day 13 Activity: Reshaping data"
subtitle: "Thursday, June 4, 2026"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
```

Today's verbs are `pivot_longer()` and `pivot_wider()`. We'll use a few small built-in tables plus `penguins`.

---

# Part 1

Look at `table1` and `table2` (both built into tidyr):

```{r}
table1
table2
```

**1a.** For each, describe what **one row** represents.

> Table 1: [ Answer here. ]

> Table 2: [ Answer here. ]

**1b.** Which one is **tidy** (each variable a column, each observation a row)? Which is the **narrower** of the two, and how can you tell?

> [Answer here.]

---

# Part 2

Build this wide table by hand:

```{r build-wide}
bp_wide <- tribble(
  ~subject, ~before, ~after,
  "BHO", 160, 160,
  "GWB", 115, 135,
  "WJC", 105, 145
)
bp_wide
```

**2a. Predict first.** If you pivot `before` and `after` into a long form, how many **rows** will the result have? How many **columns**?

Prediction:

> [Pred. # Rows here]

> [Pred. # Cols here]

**2b.** Now pivot `bp_wide` to long form. Use `names_to = "when"` and `values_to = "sbp"`. Check against your prediction.

```{r longer-bp}
# Your code:

```

---

# Part 3

The `billboard` data (in tidyr) has one column per chart week (`wk1`, `wk2`, ...).

**3a.** Pivot all the `wk` columns into `week` / `rank`, dropping the empty weeks. (Hint: `cols = starts_with("wk")`, `values_drop_na = TRUE`.)

```{r longer-billboard}
# Your code:

```

**3b.** Add `mutate(week = parse_number(week))` so `"wk3"` becomes `3`, then plot `rank` (y) against `week` (x), one faint line per `track`. Use `scale_y_reverse()` so rank 1 sits on top.

```{r billboard-plot}
# Your code:

```

**One sentence:** what does the plot show about how long songs stay on the charts?

> [Answer here.]

---

# Part 4

Start from a long summary table:

```{r long-summary, message=FALSE}
penguin_summary <- penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(avg_mass = mean(body_mass_g, na.rm = TRUE), .groups = "drop")
penguin_summary
```

**4a.** If you widen this with `names_from = sex`, `values_from = avg_mass`, how many rows and columns will you get?

Prediction:

> [Pred. # Rows here]

> [Pred. # Cols here]

**4b.** Pivot, then add a column `diff` = `male` - `female`.

```{r wider-penguins}
# Your code:

```

**Which species has the biggest mass gap between sexes?**

> [Answer here.]

---

# Part 5

**5a.** Take your widened penguin table from Part 4 (before the `diff` column) and pivot it **back** to long form. Do you recover the original `penguin_summary`? (`pivot_longer()` and `pivot_wider()` are inverses.)

```{r round-trip}
# Your code:

```

**5b.** Someone tries to make `bp_wide` longer like this and gets a strange result:

```{r too-narrow}
#| eval: false
bp_wide |>
  pivot_longer(
    cols = c(subject, before, after),
    names_to = "when",
    values_to = "sbp"
  )
```

Run it. Why is this "too narrow" / wrong? What single change to `cols` fixes it?

> [Answer here.]

---

# Part 6 --- Choose the shape

For each goal, say whether you'd want the data **wide** or **narrow**, and which pivot gets you there.

1. Plot each country's TB cases over time, one line per country.
2. Compute, for each country, the change in cases from 1999 to 2000 as a single new column.
3. Group by year and sum the cases across all countries.

> 1. [Answer here.]
>
> 2. [Answer here.]
>
> 3. [Answer here.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
