---
title: "Day 10 Activity: Summarizing the `penguins`"
subtitle: "Monday, June 1, 2026"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
```

Today's activity uses the `penguins` dataset (`palmerpenguins` package). Glimpse it before you start:

```{r}
glimpse(penguins)
```

We will practice using **two data verbs**, `summarize()` and `group_by()`, as well as reduction functions like `n()` and `mean()`.

---

# Part 1

For each operation below, say which family it belongs to: **reduction** (variable to one number), **transformation** (variable to one value *per row*), or **neither** (a whole-table data verb).

| # | Operation | Family |
|---|---|---|
| 1 | The largest body mass in the dataset | [Your answer.] |
| 2 | Body mass converted from grams to kilograms | [Your answer.] |
| 3 | The number of penguins | [Your answer.] |
| 4 | Whether each penguin weighs more than 4000 g | [Your answer.] |
| 5 | The number of *distinct* islands | [Your answer.] |
| 6 | Labeling each penguin `"heavy"` or `"light"` | [Your answer.] |
| 7 | Keeping only the Gentoo penguins | [Your answer.] |
| 8 | The standard deviation of flipper length | [Your answer.] |

: Fill in the third column. {tbl-colwidths="[8,62,30]"}

::: {.callout-tip}
The fastest test: does it return **one** number (reduction), **one per row** (transformation), or a **whole new table** (neither, that's a data verb)? Number 7 is tricky.
:::

---

# Part 2

**2a.** Write a single `summarize()` that returns the number of penguins, the mean body mass, and the mean flipper length.

```{r summary-1}
# Your code:

```

**2b.** Run this, then explain in one sentence *why* the result is what it is:

```{r summary-2}
penguins |>
  summarize(avg_mass = mean(body_mass_g))
```

[Your explanation here.]

**2c.** Fix 2b so it returns a real number. What one change did you make, and why does it work?

```{r summary-3}
# Your fixed code:

```

[Your explanation here.]

---

# Part 3

For **3a–3d**, first **write down your prediction** for how many rows the result will have and which columns it will contain. *Then* run it and check.

**3a.**

```{r group-a}
#| eval: false
penguins |>
  group_by(species) |>
  summarize(n = n(), avg_mass = mean(body_mass_g, na.rm = TRUE))
```

Prediction (rows / columns): [Your answer.]

Actual: [Your answer.]

**3b.**

```{r group-b}
#| eval: false
penguins |>
  group_by(species, sex) |>
  summarize(n = n(), avg_mass = mean(body_mass_g, na.rm = TRUE),
            .groups = "drop")
```

Prediction (rows / columns): [Your answer.]

Actual: [Your answer.]

Did anything about the **number of rows** surprise you? (Look closely at the `sex` column.)

[Your answer.]

**3c.** This one is a trap. Predict *before* you run it, then run it:

```{r group-c}
#| eval: false
penguins |>
  group_by(body_mass_g) |>
  summarize(n = n(), .groups = "drop")
```

Prediction (rows / columns): [Your answer.]

Actual: [Your answer.]

What went wrong, and what does it tell you about which *kinds* of variables you should group by?

[Your answer.]

**3d.** Without grouping at all:

```{r group-d}
#| eval: false
penguins |>
  summarize(n = n())
```

Prediction (rows / columns): [Your answer.]

Actual: [Your answer.]

Explain the relationship between 3a, 3b, and 3d in terms of "what does one row of the output represent?"

[Your answer.]

---

# Part 4 --- Don't trust a mean without its `n()`

Run the finest grouping we can make from these three variables:

```{r trust}
#| eval: false
penguins |>
  group_by(species, island, sex) |>
  summarize(n = n(), avg_mass = mean(body_mass_g, na.rm = TRUE),
            .groups = "drop")
```

**4a.** Which group has the **fewest** penguins? Would you be comfortable quoting that group's `avg_mass` in a report? Why or why not?

[Your answer.]

**4b.** In one or two sentences: why is `n = n()` worth adding to *every* group summary you write, even when nobody asked for a count?

[Your answer.]

---

# Wrap-up

You now have the two verbs that change *what a row means*. Tomorrow we add the four verbs that change *which rows and columns are present*.

**One-sentence reflection:** what was the single most useful idea from today, the framework, the `na.rm` trap, or predicting the shape of a grouped result?

[Write one sentence.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
