---
title: "Day 11 Activity: Wrangling the `penguins`"
subtitle: "Tuesday, June 2, 2026"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(palmerpenguins)
```

Today's activity uses the `penguins` dataset again. Glimpse it before you start:

```{r}
glimpse(penguins)
```

You now have all six verbs: yesterday's `group_by()` and `summarise()`, and today's `filter()`, `arrange()`, `select()`, and `mutate()`.

---

# Part 1 --- Match the verb

Each task can be done with **one** verb (one task needs two). Name the verb(s) you'd use, no code yet.

| # | Task | Verb(s) |
|---|---|---|
| 1 | Keep only the Gentoo penguins | |
| 2 | Add a column of body mass in kilograms | |
| 3 | Sort the penguins heaviest-first | |
| 4 | Keep only the `species`, `island`, and `body_mass_g` columns | |
| 5 | One row per species, each with its average body mass | |
| 6 | Drop the penguins whose `sex` is missing | |

: Fill in the third column. {tbl-colwidths="[8,62,30]"}

---

# Part 2 --- filter()

Write a **single pipeline** for each. Run each one and glance at the number of rows.

**2a.** Penguins heavier than 5000 g.

```{r filter-a}
# Your code:

```

**2b.** Adelie penguins that live on Dream island (both conditions).

```{r filter-b}
# Your code:

```

**2c.** Penguins that are Adelie *or* Gentoo. Write it **two ways**, once with `|` and once with `%in%`.

```{r filter-c}
# Your code:

```

**2d. Debug it.** This errors. What's the message, why does it happen, and what's the fix?

```{r filter-d}
#| eval: false
penguins |>
  filter(species = "Chinstrap")
```

[Your explanation and fix.]

---

# Part 3 --- mutate() and arrange()

**3a.** Add a column `bill_ratio` = bill length ÷ bill depth. Then `arrange()` so the largest ratio is on top. (Which species ends up at the top?)

```{r mutate-a}
# Your code:

```

[Which species is on top?]

**3b.** Add a column `size` that is `"large"` when `body_mass_g > 4000` and `"small"` otherwise. (Hint: `ifelse()`.)

```{r mutate-b}
# Your code:

```

**3c.** *Reduction or transformation?* You want the **single heaviest** body mass in the whole dataset. Does that belong in `mutate()` or `summarise()`? Write it, and say why.

```{r mutate-c}
# Your code:

```

[Why that verb?]

---

# Part 4 --- select()

**4a.** Keep only `species`, `sex`, and the two bill columns.

```{r select-a}
# Your code:

```

**4b.** Do the same selection of the two bill columns using a **helper** (`starts_with()` or `ends_with()`) instead of naming them.

```{r select-b}
# Your code:

```

**4c.** Select `body_mass_g` but rename it to `mass_g` in the same step.

```{r select-c}
# Your code:

```

---

# Part 5 --- Put it together

Answer this in **one pipeline**: *Among penguins with a known sex, what is the average flipper length for each species/sex combination, and how many penguins is each average based on? Show the longest-flippered group first.*

Before you run it, **predict** how many rows the result has.

Prediction (rows): [ ... ]

```{r together}
# Your pipeline (filter -> group_by -> summarise -> arrange):

```

Now turn the same filtered data into a plot of `body_mass_g` by `species`, filled by `sex`:

```{r together-plot}
# Your plot (remember: switch |> to + when you reach ggplot):

```

**One or two sentences:** what does the table or plot show?

[Your answer.]

---

# Part 6 --- Restyle it

The pipeline below works, but it's unreadable. Rewrite it following the style guide: a space before each `|>`, `|>` at the end of the line, each named argument of `summarise()` on its own line indented two spaces, and the closing `)` on its own line.

```{r restyle}
#| eval: false
penguins|>filter(!is.na(sex))|>group_by(species,sex)|>summarize(n=n(),avg_mass=mean(body_mass_g,na.rm=TRUE))|>arrange(desc(avg_mass))
```

```{r restyle-fixed}
# Your restyled version:

```

::: {.callout-tip}
Once you've done it by hand, try the `styler` package: `Cmd/Ctrl + Shift + P`, then type "styler" and pick *"Style active file."* Did it match yours?
:::

---

# Wrap-up

**One-sentence reflection:** of the four verbs today, which felt most natural and which will you have to look up again next week?

[Write one sentence.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
