---
title: "The data verbs: select · filter · mutate · arrange"
subtitle: "Tuesday, June 2, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day11-data-verbs-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day11-data-verbs-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(palmerpenguins)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Tuesday June 2]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Assignment**: [*DC* §10.1–10.6](https://psu.instructure.com/courses/2487468/assignments/18325370)

**Plan for today**

- Four new data verbs: `filter`, `arrange`, `select`, `mutate`
- Sorting verbs by *what they touch*: rows vs. columns
- Combining verbs with the pipe into real pipelines
- Formatting pipelines so they stay readable

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*R4DS* Ch. 4](https://r4ds.hadley.nz/workflow-style) · [*DC* Ch. 10](https://dtkaplan.github.io/DataComputingEbook/chap-more-data-verbs.html#chap:more-data-verbs) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::

## Recap from yesterday

We discussed the overarching **framework** and two data verbs:

| Family | Input to Output | The verb that uses it |
|---|---|---|
| Reduction | variable to scalar | `summarise()` |
| Transformation | variable to variable | *...today: `mutate()`* |
| Data verb | data frame to data frame | all of them |

: {tbl-colwidths="[28,42,30]"}

We write reduction functions inside of `summarise()`. Today we will cover `mutate()`, which contains **transformation** functions, plus three verbs that reshape *which* rows and columns you keep.

## A map of the verbs

We can organize dplyr by **what each verb operates on**:

| Touches... | Verbs | What changes |
|---|---|---|
| **Rows** | `filter()`, `arrange()` | which rows, and their order |
| **Columns** | `select()`, `mutate()` | which columns, and new ones |
| **Groups** | `group_by()`, `summarise()` | *(yesterday)* |

: The dplyr toolkit. {tbl-colwidths="[20,42,38]"}

::: {.callout-tip style="font-size: 1.1em;"}
These all follow the same three rules: **first argument is a data frame, you name columns without quotes, and the output is a brand-new data frame.**
:::

# Rows: filter() {background-color="#001E44"}

## filter(): keep the rows you want {.scrollable .nostretch}

`filter()` keeps only the rows that pass a test (*DC* §10.2, R4DS §3.2.1):

```{r}
#| echo: TRUE
penguins |>
  filter(species == "Gentoo")
```

The test is a **transformation** that returns `TRUE`/`FALSE` for each row; `filter()` keeps the `TRUE`s.

## The comparison operators {.scrollable .nostretch}

These are the tests you'll use inside `filter()` (*DC* Table 10.9):

| Operator | Means | Example |
|---|---|---|
| `==` | equal to | `species == "Gentoo"` |
| `!=` | not equal to | `sex != "male"` |
| `>` `>=` | greater (or equal) | `body_mass_g >= 5000` |
| `<` `<=` | less (or equal) | `flipper_length_mm < 190` |
| `%in%` | is one of | `island %in% c("Dream", "Biscoe")` |

: Comparisons for `filter()`. {tbl-colwidths="[18,32,50]"}

::: {.callout-warning}
Use `==` to **test** equality, not `=`. A single `=` means "assign," not "test equality".
:::

## Multiple conditions {.scrollable .nostretch}

Separate tests with a comma (or `&`) for **AND**, a vertical bar `|` for **OR** (R4DS §3.2.1):

```{r}
#| echo: TRUE
#| eval: FALSE
# AND: Adelies that are ALSO on Dream
penguins |>
  filter(species == "Adelie", island == "Dream")

# OR: any penguin that is Adelie OR Gentoo
penguins |>
  filter(species == "Adelie" | species == "Gentoo")

# %in% is the clean way to write a long OR
penguins |>
  filter(species %in% c("Adelie", "Gentoo"))
```

::: {.callout-note}
`filter(species == "Adelie" | "Gentoo")` does **not** work the way it reads. Each side of `|` must be a complete test.
:::

## Dropping `NA`s {.scrollable .nostretch}

Yesterday, `group_by(species, sex)` gave us an `NA` group because some penguins have a missing `sex`. Now we can drop them:

```{r}
#| echo: TRUE
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(n = n(), .groups = "drop")
```

`is.na(sex)` is `TRUE` for the missing ones; `!` flips it, so `!is.na(sex)` keeps the rows that **do** have a sex. This removes the `NA` group.

# Rows: arrange() {background-color="#001E44"}

## arrange(): set the order {.scrollable .nostretch}

`arrange()` reorders rows without adding or removing any. This function gives the result in ascending order by default:

```{r}
#| echo: TRUE
penguins |>
  arrange(body_mass_g)
```

You can wrap a column in `desc()` if you want descending order. If there are ties, to break them you can list several columns:

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |>
  arrange(species, desc(body_mass_g))   # by species, then heaviest first
```

# Columns: select() {background-color="#001E44"}

## select(): keep the columns you want {.scrollable .nostretch}

`select()` is to columns what `filter()` is to rows:

```{r}
#| echo: TRUE
penguins |>
  select(species, island, body_mass_g)
```

You can grab a range, drop with `!`, or rename as you go:

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |> select(species:bill_depth_mm)  # a range of columns
penguins |> select(!year)                  # everything EXCEPT year
penguins |> select(mass = body_mass_g)     # rename while selecting
```

::: {.callout-note}
`penguins` has only 8 columns, so `select()` feels optional here. On a 19-column table like `flights` from HW2, or a 76-column one, it's one of the first things you might want to try.
:::

## select() helpers {.scrollable .nostretch}

For wide tables, name columns by pattern instead of one-by-one:

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |> select(starts_with("bill"))    # bill_length_mm, bill_depth_mm
penguins |> select(ends_with("_mm"))       # all the millimetre measurements
penguins |> select(where(is.numeric))      # every numeric column
```

# Columns: mutate() {background-color="#001E44"}

## mutate(): add a new column {.scrollable .nostretch}

`mutate()` adds columns computed from existing ones, keeping every row:

```{r}
#| echo: TRUE
penguins |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  select(species, body_mass_g, body_mass_kg)
```

## mutate() contains transformation functions {.scrollable .nostretch}

> `mutate()` **always** involves a *transformation* function, just as `summarise()` **always** involves a *reduction* function.

```{r}
#| echo: TRUE
#| eval: FALSE
# transformation: one value per row to mutate()
penguins |> mutate(bill_ratio = bill_length_mm / bill_depth_mm)

# reduction: one value total to summarise()
penguins |> summarise(avg_bill = mean(bill_length_mm, na.rm = TRUE))
```

::: {.callout-tip}
Deciding between them is the same question from yesterday: does the calculation give **one value per row** (`mutate`) or **one value for the whole group** (`summarise`)?
:::

## A categorical column with ifelse() {.scrollable .nostretch}

`ifelse()` is a transformation that picks between two values per row, perfect inside `mutate()`:

```{r}
#| echo: TRUE
penguins |>
  mutate(size = ifelse(body_mass_g > 4000, "large", "small")) |>
  select(species, body_mass_g, size)
```

# Combining verbs {background-color="#001E44"}

## The pipe does the real work {.scrollable .nostretch}

One verb does one thing. For most questions, we will need several, chained together with `|>`:

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |>
  filter(!is.na(sex)) |>
  mutate(body_mass_kg = body_mass_g / 1000) |>
  select(species, sex, body_mass_kg) |>
  arrange(desc(body_mass_kg))
```

This can be read in English as: take penguins, **then** drop missing sex, **then** add mass in kg, **then** keep three columns, **then** sort heaviest first.

::: {.callout-tip}
Because each verb returns a new data frame, the output of one is the input to the next. The verbs line up on the left so you can quickly skim to understand what's going on.
:::

## Putting all six together {.scrollable .nostretch}

We can use all six verbs to answer the question, "what's the average mass of each species/sex, biggest first?":

```{r}
#| echo: TRUE
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(
    n        = n(),
    avg_mass = mean(body_mass_g, na.rm = TRUE),
    .groups  = "drop"
  ) |>
  arrange(desc(avg_mass))
```

And because the result is just a data frame, you can pipe it straight into a plot (switching `|>` to `+`):

```{r}
#| echo: TRUE
#| fig-height: 2.7
#| warning: FALSE
penguins |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot()
```

## Formatting your pipelines {.scrollable .nostretch}

A multi-step pipeline is only readable if you format it. The rules that matter most:

- A space before `|>`, and let `|>` be the **last thing on the line**.
- For verbs with **named arguments** (`mutate`, `summarise`), put each argument on its **own line**, indented two spaces.
- Verbs without named arguments (`filter`, `select`, `arrange`) can stay on one line if they fit.
- Name new columns in `snake_case`: `body_mass_kg`, not `BodyMassKG`.

```{r}
#| echo: TRUE
#| eval: FALSE
# Strive for                          # Avoid
penguins |>                           penguins|>filter(!is.na(sex))|>group_by(
  filter(!is.na(sex)) |>              species)|>summarize(n=n(),avg=mean(
  group_by(species) |>                body_mass_g,na.rm=TRUE))
  summarise(
    n   = n(),
    avg = mean(body_mass_g, na.rm = TRUE)
  )
```

::: {.callout-tip}
Don't hand-format forever: the **`styler`** package reformats a whole file for you. In RStudio, `Cmd/Ctrl + Shift + P` then type "styler". Full guide: R4DS Ch. 4.
:::

## A helpful shortcut {.scrollable .nostretch}

Counting rows per group is so common it has its own verb, `count()`:

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |> count(species)

# is exactly the same as
penguins |>
  group_by(species) |>
  summarise(n = n())
```

::: {.callout-note}
`count()` is just `group_by()` + `summarise(n = n())` bundled up. Add `sort = TRUE` to get the biggest groups first.
:::

## What we covered today {.scrollable .nostretch}

- **`filter()`**: keep rows that pass a test; `==` not `=`; combine with `,`/`|`/`%in%`; `!is.na()` to drop missing
- **`arrange()`**: reorder rows; `desc()` for descending; extra columns break ties
- **`select()`**: keep/drop/rename columns; `starts_with()` & friends for wide tables
- **`mutate()`**: add columns from transformations; the per-row twin of `summarise()`
- **The pipe**: chain verbs into a readable top-to-bottom story; pipe results into ggplot
- **Style**: space before `|>`, named args on their own lines, `snake_case`, let `styler` do the rest

# Activity {background-color="#7C0A02"}

## Activity: wrangling the penguins

Roughly 25 minutes. Individually or with a neighbor.

- [Activity Canvas Link](https://psu.instructure.com/courses/2487468/assignments/)

**Format:** a handout that walks each verb, then asks you to chain them. You'll match tasks to verbs, write `filter`/`arrange`/`select`/`mutate` pipelines on `penguins`, debug a `=` vs `==` error, build one multi-verb pipeline that answers a question, and restyle a deliberately unreadable pipeline.

**To turn in:** save your `.qmd` and rendered HTML.

## Wrap-up & looking ahead

**Tomorrow (Wednesday):** no new reading, we put all six verbs together on a messier dataset and turn raw rows into summary tables and plots.

:::: {.columns}
::: {.column width="55%"}
**Upcoming Deadlines**

- **Thu 6/4, before class**: [*DC* §12.1](https://psu.instructure.com/courses/2487468/assignments/18325371)
- **Thu 6/4, 11:59 p.m.**: [Project: Team formation](https://psu.instructure.com/courses/2487468/assignments/)
- **Fri 6/5, 11:59 p.m.**: [Homework 3: Data Wrangling & Reshaping](https://psu.instructure.com/courses/2487468/assignments/18327019)
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 3](https://r4ds.hadley.nz/data-transform) · [*R4DS* Ch. 4](https://r4ds.hadley.nz/workflow-style) · [*DC* Ch. 10](https://dtkaplan.github.io/DataComputingEbook/chap-more-data-verbs.html#chap:more-data-verbs) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::
::::
