---
title: "Day 12 Activity: A `flights` investigation"
subtitle: "Wednesday, June 3, 2026"
author: "Your Name Here"
editor: source
engine: knitr
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
```

Today we use `flights`, every flight out of NYC in 2013. Glimpse it, and note that cancelled flights are denoted using `NA` missing values in the `dep_delay` and `arr_delay` columns.

```{r}
glimpse(flights)
```

Today, we will practice using the six verbs (`filter`, `arrange`, `select`, `mutate`, `group_by`, `summarise`), plus `count()` and the pipe to chain them together.

---

# Part 1 --- Plan before you code

For this question: *Which of the three NYC airports (`origin`) sent out the most flights in 2013?*

**1a.** First, write out your plan in English using one sentence per step. Answer the following questions:

1. Which verbs will you need to use, and in what order? 

2. What does one row of your target table represent?

> [Write your plan here.]

**1b.** Now implement your plan with code. (Recall: `count()` is a shortcut for a `group_by()` + `summarise()`. Both give the same result.)

```{r part1}
# Your code:

```

---

# Part 2 --- Build a pipeline

*Which carrier had the worst average arrival delay, among carriers with at least 1,000 flights?*

Build a pipeline step by step; filter is **not** needed on the raw rows here, but you'll need it on the **summary table** to drop small carriers. Include `n = n()`.

```{r part2}
# Your code:

```

**Which carrier is it, and how many flights is that average based on?**

> [Write your answer here. Note that we only have the 2-letter code for carrier in today's data; turning it into the airline's name needs a *join*, which we've seen but haven't covered in detail yet.]

---

# Part 3 --- mutate(), then plot

**3a.** Create a new variable `time_of_day` that takes on the value `"morning"` when `dep_time < 1200` and `"afternoon/evening"` otherwise. Then find the average departure delay for each, with a count.

```{r part3a}
# Your code:

```

**3b.** Pipe a version of that summary into a `geom_col()` plot of average delay by `time_of_day`.

```{r part3b}
# Your code:

```

In one sentence, explain what your plot shows:

> [Write your answer here.]

---

# Part 4 --- Don't get fooled

**4a. Order matters.** These two pipelines run without error but answer **different questions**. Run both, then explain in one or two sentences how the questions differ.

```{r part4a-i}
#| eval: false
# Pipeline A
flights |>
  filter(dest == "BOS") |>
  summarise(avg = mean(arr_delay, na.rm = TRUE))
```

```{r part4a-ii}
#| eval: false
# Pipeline B
flights |>
  group_by(dest) |>
  summarise(avg = mean(arr_delay, na.rm = TRUE)) |>
  filter(dest == "BOS")
```

> [Write your answer here. Address the following: how do the questions differ? Do the numbers match? Should they?]

**4b. Spot the silent `NA`.** This returns `NA` for some carriers. Why? What single change fixes it?

```{r part4b}
#| eval: false
flights |>
  group_by(carrier) |>
  summarise(avg_delay = mean(dep_delay))
```

> [Write your explanation and the fix in words here.]

---

# Part 5 --- Open investigation

Pick **one** of these questions (or invent your own), and answer it with a pipeline:

- Which destination has the most *variable* arrival delays? (Hint: `sd()` is a reduction, like `mean()`.)
- Do flights later in the day make up time in the air more than morning flights?
- Which day of the week (you would needthe  `month` and `day` variables) is worst to fly? *(Stick to month if day of the week is too much.)*

**5a.** State your question and your English plan.

> [Write your question & plan here.]

**5b.** Your pipeline:

```{r part5}
# Your code:

```

**5c. Report your checks.** Answer the following three questions:

1. How did you handle missing values?
2. What's the smallest sample size `n` in your result, and is it big enough to trust?
3. Does the answer match your intuition (what you would expect)?

> 1. [Write your answer here.]
>
> 2. [Write your answer here.]
>
> 3. [Write your answer here.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
