---
title: "Reshaping: pivot_longer · pivot_wider"
subtitle: "Thursday, June 4, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day13-pivot-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day13-pivot-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(palmerpenguins)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Thursday June 4]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Assignment**: [*DC* §12.1](https://psu.instructure.com/courses/2487468/assignments/18325371)

**Plan for today**

- What makes data **tidy**, and why the tidyverse wants it that way
- **Wide** vs. **narrow (long)** layouts of the *same* data
- Two new verbs: `pivot_longer()` and `pivot_wider()`
- Choosing a shape to fit the task

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 5](https://r4ds.hadley.nz/data-tidy) · [*DC* Ch. 12](https://dtkaplan.github.io/DataComputingEbook/chap-wide-vs-narrow.html#chap:wide-vs-narrow) · [tidyr reference](https://tidyr.tidyverse.org/reference)
:::

## Recap, and what's new

All week you've manipulated data **within** its existing shape, choosing rows, columns, and groups. That worked because the data was already laid out sensibly.

::: {.callout-tip style="font-size: 1.1em;"}
Today is the first new verbs since Monday. `pivot_longer()` and `pivot_wider()` change the **shape** of the table itself, without changing any of the values.
:::

These come from **tidyr**, dplyr's sibling package (both load with `library(tidyverse)`).

# Tidy data {background-color="#001E44"}

## What "tidy" means {.scrollable .nostretch}

R4DS §5.2 gives three rules. A dataset is **tidy** when:

1. Each **variable** is a column.
2. Each **observation** is a row.
3. Each **value** is a cell.

The same numbers can be stored many ways. Only one is tidy:

```{r}
#| echo: TRUE
table1   # tidy: country, year, cases, population each a column
```

## The same data, untidy {.scrollable .nostretch}

```{r}
#| echo: TRUE
table2   # cases and population are stacked into type/count
```

`table2` holds identical information, but `type`/`count` jams two variables into a name/value pair. Computing a rate is awkward here and natural in `table1`.

::: {.callout-note}
Why tidy? `dplyr`, `ggplot2`, and friends are **built** for it. Tidy data is the shape where `mutate()`, `group_by()`, and `aes()` all "just work."
:::

# Wide vs. narrow {background-color="#001E44"}

## Two layouts of one dataset {.scrollable .nostretch}

*DC* §12.1 frames reshaping as **wide ↔ narrow**. Blood pressure, wide:

```{r}
#| echo: TRUE
bp_wide <- tribble(
  ~subject, ~before, ~after,
  "BHO",        160,    160,
  "GWB",        115,    135,
  "WJC",        105,    145
)
bp_wide
```

One row per **subject**; `before`/`after` are separate columns.

## The same data, narrow {.scrollable .nostretch}

```{r}
#| echo: TRUE
bp_narrow <- bp_wide |>
  pivot_longer(
    cols      = c(before, after),
    names_to  = "when",
    values_to = "sbp"
  )
bp_narrow
```

Now one row per **measurement occasion**: the case changed meaning, the numbers didn't.

::: {.callout-tip}
Each shape has strengths. *Wide* makes a before/after change a one-line `mutate(change = after - before)`. *Narrow* makes it easy to add more occasions or more variables, and it's what `ggplot2` usually wants.
:::

# pivot_longer() {background-color="#001E44"}

## The idea {.scrollable .nostretch}

`pivot_longer()` takes several columns and stacks them into **two** new columns: one holding the old **names**, one holding the old **values** (R4DS §5.3).

Three arguments after the data:

- `cols` — which columns to gather (uses `select()` syntax: `starts_with()`, ranges, `!`)
- `names_to` — name for the new column that will hold the **old column names**
- `values_to` — name for the new column that will hold the **cell values**

```{r}
#| echo: TRUE
#| eval: FALSE
bp_wide |>
  pivot_longer(cols = c(before, after), names_to = "when", values_to = "sbp")
```

::: {.callout-warning}
`names_to` and `values_to` are **quoted** — they're names you're inventing, so they don't exist in the data yet.
:::

## A real one: when names ARE data {.scrollable .nostretch}

The `billboard` data (in tidyr) has one column **per week** (`wk1`, `wk2`, ...). The week is a variable trapped in the column names:

```{r}
#| echo: TRUE
billboard |>
  pivot_longer(
    cols      = starts_with("wk"),
    names_to  = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

`values_drop_na = TRUE` drops the empty weeks (songs that had already fallen off the chart).

## Now it's plottable {.scrollable .nostretch}

In wide form you *couldn't* put "week" on an axis, it wasn't a variable. After pivoting, it is:

```{r}
#| echo: TRUE
#| fig-height: 3
#| warning: FALSE
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week", values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week)) |>     # "wk3" -> 3
  ggplot(aes(x = week, y = rank, group = track)) +
  geom_line(alpha = 0.25) +
  scale_y_reverse()                        # rank 1 on top
```

::: {.callout-note}
This is *why* we pivot: getting a variable out of the column names and into a real column is often the step that makes a plot possible. (`parse_number()` pulls the number out of `"wk3"`.)
:::

# pivot_wider() {background-color="#001E44"}

## The opposite move {.scrollable .nostretch}

`pivot_wider()` spreads one column's values across **new columns** (R4DS §5.4, *DC* §12.1). Its two key arguments are the mirror of before:

- `names_from` — the column whose **values become new column names**
- `values_from` — the column whose values **fill the new cells**

```{r}
#| echo: TRUE
bp_narrow |>
  pivot_wider(names_from = when, values_from = sbp)
```

We're back to `bp_wide`. **`pivot_longer()` and `pivot_wider()` are inverses.**

## Where you'll actually reach for it {.scrollable .nostretch}

Remember Wednesday: `group_by() |> summarise()` produces a **long** summary. Widening it makes a readable comparison table.

```{r}
#| echo: TRUE
#| message: FALSE
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(avg_mass = mean(body_mass_g, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = avg_mass)
```

## ...and comparison becomes a mutate {.scrollable .nostretch}

With `female` and `male` now side by side in each row, the difference is one `mutate()`, exactly *DC*'s before/after point:

```{r}
#| echo: TRUE
#| message: FALSE
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(avg_mass = mean(body_mass_g, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = avg_mass) |>
  mutate(diff = male - female)
```

::: {.callout-tip}
Long form is for **computing and plotting across** a category; wide form is for **reading and comparing within** a row. Same data, you pick the shape that makes your next step a one-liner.
:::

# Choosing a shape {background-color="#001E44"}

## Which way do I pivot? {.scrollable .nostretch}

| Your goal | Shape you want | Verb to get there |
|---|---|---|
| Put a "trapped" variable on a plot axis | narrow / long | `pivot_longer()` |
| Group or summarise over a category | narrow / long | `pivot_longer()` |
| Compare two values **per row** (a − b) | wide | `pivot_wider()` |
| A human-readable comparison table | wide | `pivot_wider()` |

: Match the shape to the next step. {tbl-colwidths="[44,28,28]"}

::: {.callout-note}
R4DS §5.5: there isn't always one "tidy" shape, it depends what you're doing next. Don't be afraid to **untidy, transform, and re-tidy** as the task changes.
:::

## What we covered today {.scrollable .nostretch}

- **Tidy data:** variables in columns, observations in rows, values in cells
- **Wide vs. narrow:** the same values, different meaning of a "row"
- **`pivot_longer()`:** gather many columns into `names_to` + `values_to`; the fix when a variable is hiding in column names
- **`pivot_wider()`:** spread `names_from` + `values_from` into new columns; the way to a comparison table
- **They're inverses**, and the right shape is whichever makes your *next* step easiest

# Activity {background-color="#7C0A02"}

## Activity: reshaping

Roughly 25 minutes. Individually or with a neighbor.

- [Activity Canvas Link](https://psu.instructure.com/courses/2487468/assignments/)

**Format:** you'll diagnose whether tables are tidy/wide/narrow, pivot a tiny table by hand-prediction then check, pivot `billboard` longer and plot it, widen a `penguins` summary into a comparison table, confirm the two verbs undo each other, and debug a pivot that went "too narrow."

**To turn in:** save your `.qmd` and rendered HTML.

## Wrap-up & looking ahead

**Tomorrow (Friday):** **Quiz 3** in the first 20 minutes, then version control, Git and GitHub concepts plus a live demo, so your project team can collaborate on shared code.

:::: {.columns}
::: {.column width="55%"}
**Upcoming Deadlines**

- **Tonight, Thu 6/4, 11:59 p.m.**: [Project: Team formation](https://psu.instructure.com/courses/2487468/assignments/)
- **Fri 6/5, before class**: [*DC* §9.1–9.2](https://psu.instructure.com/courses/2487468/assignments/18325372)
- **Fri 6/5, in class**: Quiz 3
- **Fri 6/5, 11:59 p.m.**: HW 3
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 5](https://r4ds.hadley.nz/data-tidy) · [*DC* Ch. 12](https://dtkaplan.github.io/DataComputingEbook/chap-wide-vs-narrow.html#chap:wide-vs-narrow) · [tidyr reference](https://tidyr.tidyverse.org/reference)
:::
::::
