---
title: "Reshaping: pivot_longer, pivot_wider"
subtitle: "Thursday, June 4, 2026"
format:
  html:
    toc: true
    code-fold: false
editor: source
---

::: callout-note
## Today's plan
- [ ] TODO: list 2-4 concrete goals for the class
:::

## Before class

- Read: Data Computing Ch 12 (§12.1)
- TODO: any prep materials, data files, etc.

## Outline

### 1. TODO: opening segment (~15 min)

TODO: concept exposition or motivating example.

```{r}
#| eval: false
# TODO: live-coded example
```

### 2. TODO: main segment (~25 min)

TODO: hands-on practice — students follow along.

```{r}
#| eval: false
# TODO: example to walk through
```

### 3. In-class activity (~20 min)

TODO: brief description, or link to a separate handout.

### 4. Wrap-up (~5 min)

- TODO: key takeaway
- TODO: preview next class

## Resources

- TODO: slides PDF
- TODO: in-class code .R file
- TODO: activity handout

## After class

- TODO: due tomorrow
- TODO: optional supplementary reading

::: {.callout-tip collapse="true"}
## Instructor notes

TODO: notes to self — pacing, common student questions, things to watch for.
:::
