---
title: "STAT 184 — Homework [N]"
author: "Your Name Here"
date: today
format:
  pdf:
    toc: false
    code-line-numbers: true
  html:
    toc: true
    code-fold: false
editor: source
---

# AI Disclosure

*Check off all boxes that apply to your usage and provide a brief description of how the tool was utilized.*

- [ ] No AI used
- [ ] Syntax / Debugging (e.g., finding a missing comma, fixing an error code)
- [ ] Conceptual explanation (e.g., asking how a `left_join` works)
- [ ] Code Generation (e.g., asking AI to write a function from scratch)

*Briefly describe your use (1-2 sentences):* [Type your explanation here]

---

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
# Style guide followed: [Tidyverse / Google / BOAST — pick one]
```

# Question 1

## Part (a)

Write your prose answer here.

```{r q1a}
# Your code here
```

## Part (b)

```{r q1b}
# Your code here
```

# Question 2

```{r q2}
# Your code here
```

# Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
