### ------------------------------------------------------------------------ ###
### WARNING: DO NOT RUN THIS - IT IS NOT A .R FILE
### SIMPLY USE THIS TO READ THROUGH AND COPY/PASTE CODE AS NEEDED
### ------------------------------------------------------------------------ ###
## Activity: your first Quarto report

You'll build a short Quarto report from scratch. Take 25 minutes for this activity.

**Setup:**

1. Open RStudio.
2. `File > New File > Quarto Document...`
3. Title: **"Penguin Exploration"**, Author: your name.
4. Pick **HTML** as the format.
5. **Uncheck** "Use visual markdown editor" if you want to see the raw source (recommended for learning today).
6. Click **Create**.

You'll get a template document with example content. **Delete everything below the YAML header.** Keep the YAML.

**Step 1 - Save the file**
  
  `File > Save As...`: call it `penguin-report.qmd`, save somewhere you'll find it.

**Step 2 - Add a setup chunk**

Insert a code chunk (`Cmd/Ctrl + Alt + I`). Make it look like this:

```` markdown
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(palmerpenguins)
```
````

**Step 3 - Write an intro**

Below the setup chunk, add a heading and a paragraph:

```markdown
# Introduction

This report explores the **Palmer Penguins** data set, which contains
measurements on three species of penguins from islands in Antarctica.
```

**Step 4 - Add a plot**

Insert another chunk and put a histogram in it:

```` markdown
```{r}
#| label: mass-histogram
#| message: false
#| warning: false

ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of penguin body mass",
    x = "Body mass (g)",
    y = "Count"
  )
```
````

**Step 5 - Add commentary**

Below the plot, write 2–3 sentences (in Markdown) describing what you see.

**Step 6 - Render**

Press `Cmd/Ctrl + Shift + K`. Look at the HTML file that pops up.

If you finish early, try one or more of these:

- Add a second plot using a different variable (`flipper_length_mm`, `bill_length_mm`, ...).
- Try changing `format: html` to `format: docx` in the YAML and re-render.
- Add a **table** of the first few rows: insert a chunk with `knitr::kable(head(penguins))`.
- Set `echo: false` in your YAML under `execute:` to hide *all* code from the output:

```yaml
execute:
  echo: false
```

**To turn in:** Save your `.qmd` file and the rendered HTML. We'll check at the end of class.
