---
title: "Penguin Exploration"
author: "YOUR NAME HERE"
date: "2026-05-21"
format: html
---

<!--
=============================================================================
Day 4 Activity: Your First Quarto Report
STAT 184 - Summer 2026

Goal: Build a short Quarto report from scratch and render it to HTML.
Time: ~25 minutes.

INSTRUCTIONS:
1. Open this file in RStudio. (File > Open File...)
2. Save it somewhere you'll remember (e.g., a "stat184" folder).
3. Replace "YOUR NAME HERE" in the YAML header above with your actual name.
4. Work through the steps below. Replace each "TODO" with your own content.
5. Render the document by pressing Cmd/Ctrl + Shift + K.
6. Re-render whenever you make a change (often).

WHAT TO TURN IN:
- This .qmd file (saved as `penguin-report.qmd`)
- The rendered HTML file
=============================================================================
-->

# Introduction

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(palmerpenguins)
```

<!--
STEP 1: Replace this comment with a 2-3 sentence introduction.

Tell the reader what this report is about. The data set you're using is
called `penguins` and contains measurements on three species of penguins
from islands in Antarctica.

Use **bold** or *italic* to emphasize at least one word.
-->

TODO: Write your introduction here.

# A first plot

<!--
STEP 2: The chunk below makes a histogram of penguin body mass.
You don't need to change this code -- just look at it and run it.
-->

```{r}
#| label: mass-histogram
#| message: false
#| warning: false

ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of penguin body mass",
    x = "Body mass (g)",
    y = "Count"
  )
```

<!--
STEP 3: Replace this comment with 2-3 sentences describing what you see
in the plot above. Some things to mention:

- Roughly what range of body masses do penguins have?
- Is the distribution symmetric? Skewed? Does it have one peak or more?
- Anything else interesting?
-->

TODO: Write your description of the plot here.

# Render

<!--
STEP 4: Press Cmd/Ctrl + Shift + K to render this document to HTML.
Look at the rendered file that pops up. Does it look right?

If you get an error:
- Read the error message carefully.
- Check that every code chunk has matching ```{r} and ``` lines.
- Check that you've installed palmerpenguins (run install.packages("palmerpenguins") in the CONSOLE if needed).
- Wave me down or ask a neighbor.
-->

---

# Optional Part (Time Permitting)

If you finish early, try one or more of these.

## Option A: A second plot

Pick a different numeric variable from `penguins` and make a histogram of it.
Some options: `flipper_length_mm`, `bill_length_mm`, `bill_depth_mm`.

Add a new code chunk (`Cmd/Ctrl + Alt + I`) and copy the histogram code,
then change the variable name and the labels.

## Option B: Render to Word instead

Change `format: html` to `format: docx` in the YAML header at the top,
then re-render. What happens?

## Option C: Add a table

Add a new code chunk somewhere and put this code in it:

```r
knitr::kable(head(penguins))
```

Re-render and see what shows up.

## Option D: Hide the code

In the YAML header, add:

```yaml
execute:
  echo: false
```

Re-render. What changed?
