### ------------------------------------------------------------------------ ###
### WARNING: DO NOT RUN or RENDER THIS
### SIMPLY USE THIS TO READ THROUGH AND TYPE OUT OR COPY/PASTE CODE AS NEEDED
### ------------------------------------------------------------------------ ###
## Activity: Your First Quarto Report

You'll build a short Quarto report from scratch. Take 20 minutes for this 
activity.

**Setup:**

1. Open RStudio.
2. `File > New File > Quarto Document...`
3. Title: **"Penguin Exploration"**, Author: your name.
4. Pick **HTML** as the format.
5. **Uncheck** "Use visual markdown editor" if you want to see the raw source 
   (recommended for learning today).
6. Click **Create**.

You'll get a template document, potentially with example content. **Delete 
anything below the YAML header if present.** Keep the YAML.

**Step 1 - Save the file**
  
  `File > Save As...`: call it `penguin-report.qmd`, save somewhere you'll 
  find it.

**Step 2 - Upgrade your YAML**

Let's make the HTML output look professional right away. Modify the top of 
your document (between the `---` lines) to add a Table of Contents (`toc`) 
and a custom theme. **Watch your indentation—YAML is picky.**

```yaml
---
title: "Penguin Exploration"
author: "Your Name"
format: 
  html:
    toc: true
    theme: litera
---
```

DON'T FORGET: *Render and open your html file to see if it worked.*

**Step 3 - Add a setup chunk**

Insert a code chunk (`Ctrl + Alt + I` on Windows or `Cmd + Option + I` on Mac). 
Make it look like this:

````markdown
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(palmerpenguins)
```
````

DON'T FORGET: *Render and open your html file to see if it worked.*

**Step 4 - Write an intro with a link, an image, and inline code**

Below the setup chunk, add a heading and a paragraph. 

Notice three new Markdown features here:

  1. **Images:** Formatted as `![alt text](url)`. We're pulling the package 
     hex logo straight from the web and using Quarto's `{width}` and 
     `fig-align` attributes to float it to the right.
  2. **Hyperlinks:** Formatted as `[text](url)`.
  3. **Inline code:** The backticks around `r nrow(penguins)` will 
     automatically calculate the number of rows in the dataset when you render.

Copy this into your document:

```markdown
# Introduction

![]([https://allisonhorst.github.io/palmerpenguins/logo.png](https://allisonhorst.github.io/palmerpenguins/logo.png)){width="120" fig-align="right"}

This report explores the [**Palmer Penguins**](https://allisonhorst.github.io/palmerpenguins/) 
data set, which contains measurements on `r nrow(penguins)` penguins from 
three islands in Antarctica.
```

DON'T FORGET: *Render and open your html file to see if it worked.*

**Step 5 - Add a plot**

Insert appropriate headings and another chunk (see if you can remember the 
keyboard shortcut for adding a chunk.) Then put a histogram in your new chunk. 
We are adding a fig-cap to label the figure, which lets us automatically 
reference it in our text.

````markdown
# Exploratory Data Analysis

## Basic Histogram

```{r}
#| label: fig-mass-histogram
#| fig-cap: "Distribution of penguin body mass"
#| message: false
#| warning: false

ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    x = "Body mass (g)",
    y = "Count"
  )
```
````

DON'T FORGET: *Render and open your html file to see if it worked.*

**Step 6 - Add commentary**

Below the plot, write 2–3 sentences (in Markdown) describing what you see.
Include this exact text somewhere in your paragraph to see how Quarto 
handles cross-references:

"As shown in @fig-mass-histogram, the data is..."

Note, you do not need the quotes. Just write the text and @fig-mass-histogram.

DON'T FORGET: *Render and open your html file to see if it worked.*

**Step 7 - Add a Table**

Let's look at the raw data. Add a new section heading and a code chunk that 
uses knitr::kable() to draw a clean table of the first 5 rows:

````markdown
## Data Preview

```{r}
#| label: tbl-preview
#| tbl-cap: "First 5 rows of the penguins dataset"

knitr::kable(head(penguins, 5))
```
````

DON'T FORGET: *Render and open your html file to see if it worked.*

**Step 8 - Final Render**

Press Cmd/Ctrl + Shift + K. Look at the HTML file that pops up. Notice your 
table of contents, your styled theme, the evaluated inline code, and the 
figure numbering!

If you finish early, try one or more of these:

- **Create a Tabset:** Want to hide multiple plots behind clickable tabs? Wrap 
  two headings and code chunks in a tabset block like this:

```markdown
::: {.panel-tabset}

### Mass Plot
(Put your mass plot code chunk here)

### Flipper Plot
(Put a new flipper_length_mm plot code chunk here)

:::
```

- **Change the Output:** Try changing `format: html` to `format: docx` in the 
  YAML and re-render to get a Word document. (Note: themes and tabsets are 
  HTML-only).
- **Hide the Code:** Set `echo: false` in your YAML under `execute:` to hide 
  *all* code from the output, leaving only the text, plots, and tables:

```yaml
---
title: "Penguin Exploration"
author: "Your Name"
format: 
  html:
    toc: true
    theme: litera
execute:
  echo: false
---
```

**To turn in:** Save your `.qmd` file and the rendered HTML. We'll check at 
the end of class.