---
title: "`%>%` The pipe operator `|>`"
subtitle: "Day 5 - Friday, May 22, 2026"
lang: en
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day5-pipe-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day5-pipe-page.html"
  pdf:
    fig-pos: "H"
editor: source
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(palmerpenguins)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Friday May 22]{style="font-size: 1.3em;"}

**Before class**

- Read *Data Computing* §3.6 ([Click here](https://psu.instructure.com/courses/2487468/assignments/18307712)).

**Plan for today**

- Quiz 1 (20 min)
- The pipe operator
- Chaining commands together
- In-class activity

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*Data Computing* §3.6](https://dtkaplan.github.io/DataComputingEbook/chap-basic-r-commands.html)
:::

# Quiz 1 {background-color="#7C0A02"}

## Quiz 1 — instructions

::: {.callout-important style="font-size: 1.1em;"}
## Closed-book, no AI, no electronics
- Put away phones and laptops.
- You may use a pen or pencil.
- 20 minutes.
- I'll hand out the quiz now and collect it at the end.
:::

**Topics covered:**

- R as a calculator and basic syntax
- Data types and structures
- Objects and assignment
- Reading R code (functions, arguments, variables, constants)
- Packages: installing vs. loading
- Histograms in `ggplot2` (reading code, not writing from scratch)

When you finish early, flip your quiz face-down and wait quietly.

## Quiz debrief

A few quick notes:

- We'll go over common stumbling points on Monday after I've graded them.
- If you finished with time to spare: nice.
- If you didn't finish: don't panic. Quizzes are 15% of the grade, weighted across the semester.

Now let's get into something new.

# The pipe operator {background-color="#001E44"}

## A problem we've already seen

Remember this slide from Wednesday?

```{r}
#| echo: TRUE
sqrt(mean(c(1, 4, 9, 16, 25)))
```

To read this, your eye has to jump **inside-out**:

1. First `c(1, 4, 9, 16, 25)` makes a vector.
2. Then `mean(...)` of that vector is `11`.
3. Then `sqrt(11)` is `≈ 3.317`.

That's already tough. Imagine four or five nested functions.

## A real example

Suppose we want to:

1. Take the `penguins` data,
2. Keep only rows where `body_mass_g` is not missing,
3. Group by `species`,
4. Compute the mean body mass for each species.

Written with nested function calls:

```{r}
#| eval: FALSE
#| echo: TRUE
summarize(
  group_by(
    filter(penguins, !is.na(body_mass_g)),
    species
  ),
  avg_mass = mean(body_mass_g)
)
```

You have to read it from the **inside out**, and the arguments get separated from their functions. Ugh.

## Enter the pipe

The **pipe operator** lets you read code left-to-right, top-to-bottom — the way you actually think about it.

```{r}
#| eval: FALSE
#| echo: TRUE
penguins |>
  filter(!is.na(body_mass_g)) |>
  group_by(species) |>
  summarize(avg_mass = mean(body_mass_g))
```

Read this as:

> "Take `penguins`, **then** filter out missing values, **then** group by species, **then** summarize the average mass."

That's it. That's the whole idea.

## What the pipe does

```r
x |> f()
```

is **exactly the same** as

```r
f(x)
```

The pipe takes whatever's on the **left** and inserts it as the **first argument** to the function on the **right**.

```{r}
#| echo: TRUE
# These do the same thing:
sqrt(16)

16 |> sqrt()
```

## With multiple arguments

```r
x |> f(arg2, arg3)
```

is the same as

```r
f(x, arg2, arg3)
```

The pipe always fills the **first** argument; the rest go inside the parentheses as usual.

```{r}
#| echo: TRUE
# Without the pipe:
round(3.14159, digits = 2)

# With the pipe:
3.14159 |> round(digits = 2)
```

## Chaining pipes

The real power: you can chain pipes together to do many steps in order.

```{r}
#| echo: TRUE
c(1, 4, 9, 16, 25) |>
  mean() |>
  sqrt()
```

Compare to the nested version:

```{r}
#| echo: TRUE
sqrt(mean(c(1, 4, 9, 16, 25)))
```

Same answer. The piped version reads in the order you do the work.

## Style: one pipe per line

::: {.callout-tip}
## Convention
Put each pipe step on its **own line**, with the `|>` at the **end** of the previous line.

Indent the continuation by 2 spaces.
:::

Good:

```{r}
#| eval: FALSE
#| echo: TRUE
penguins |>
  filter(!is.na(body_mass_g)) |>
  group_by(species) |>
  summarize(avg_mass = mean(body_mass_g))
```

Not so good (works, but harder to read):

```{r}
#| eval: FALSE
#| echo: TRUE
penguins |> filter(!is.na(body_mass_g)) |> group_by(species) |> summarize(avg_mass = mean(body_mass_g))
```

# Two pipes? {background-color="#001E44"}

## `|>` vs. `%>%`

You will see **two** pipe operators in R code online:

| Pipe | Where it comes from | When introduced |
|------|---------------------|-----------------|
| `%>%` | The `magrittr` package (and tidyverse) | 2014 |
| `\|>` | Base R | 2021 |

: The two pipes in R. {tbl-colwidths="[20,50,30]"}

They do **almost the same thing**. For everything we'll cover in this course, they behave identically.

## Why two pipes?

- `%>%` came first. It's the original tidyverse pipe.
- Base R added `|>` later because the pipe became so popular they wanted it built in.
- The textbook (*Data Computing*) uses `%>%`.
- I'll use `|>` in lecture going forward — it's the modern default.

::: {.callout-note}
**You can use either one.** I'll accept both on assignments. Just be consistent within a single document.
:::

## Typing the pipe quickly

You don't have to type `|>` character-by-character. RStudio has a shortcut:

- **Mac:** `Cmd + Shift + M`
- **Windows/Linux:** `Ctrl + Shift + M`

::: {.callout-tip}
By default, this shortcut inserts `%>%`. To make it insert `|>` instead:

`Tools > Global Options... > Code > Use native pipe operator`
:::

# Reading piped code {background-color="#001E44"}

## Translating between styles

You'll see code in both styles all over the internet. Practice translating in your head.

**Nested:**

```{r}
#| eval: FALSE
#| echo: TRUE
head(arrange(filter(penguins, species == "Gentoo"), body_mass_g), 5)
```

**Piped:**

```{r}
#| eval: FALSE
#| echo: TRUE
penguins |>
  filter(species == "Gentoo") |>
  arrange(body_mass_g) |>
  head(5)
```

Same result. **Way** easier to read the piped version.

## What the pipe is NOT

::: {.callout-warning}
## The pipe doesn't run things in parallel
It just rewrites your code. Each step still runs one at a time, in order.
:::

::: {.callout-warning}
## The pipe is NOT the same as `+` in ggplot2
- `dplyr` and other data verbs: use `|>` or `%>%`.
- `ggplot2` layers: use `+`.

This trips everyone up. Keep them straight.
:::

```{r}
#| eval: FALSE
#| echo: TRUE
# Correct:
penguins |>
  filter(!is.na(body_mass_g)) |>
  ggplot(aes(x = body_mass_g)) +
    geom_histogram()
```

## A peek at what's coming

You'll see the pipe everywhere from here on out. Tomorrow we start the **dplyr** verbs:

- `filter()` — pick rows
- `select()` — pick columns
- `arrange()` — sort
- `mutate()` — make new columns
- `summarize()` — collapse to a single number
- `group_by()` — split into groups before summarizing

Every one of these is designed to be chained together with the pipe.

# In-class activity {background-color="#7C0A02"}

## Activity: pipe practice (~ 15 min)

Work individually or with a neighbor.

**Setup:**

1. Open RStudio.
2. New R script. Save as `day5-activity.R`.
3. At the top:

```{r}
#| eval: FALSE
#| echo: TRUE
library(tidyverse)
library(palmerpenguins)
```

**Goal:** Translate between nested and piped code, then write your own.

## Activity: Part 1 — translate to piped

Rewrite each of these using the pipe `|>`. Don't run them yet — just rewrite the code.

```{r}
#| eval: FALSE
#| echo: TRUE
# 1
sqrt(64)

# 2
mean(c(2, 4, 6, 8, 10))

# 3
round(sqrt(50), digits = 2)

# 4
head(arrange(penguins, body_mass_g), 3)
```

Then run both the original and your piped version. Confirm you get the same answer.

## Activity: Part 2 — translate to nested

Rewrite each of these **without** the pipe. Then run both versions.

```{r}
#| eval: FALSE
#| echo: TRUE
# 5
c(1, 2, 3, 4, 5) |> sum()

# 6
penguins |> head(2)

# 7
9 |> sqrt() |> round(1)
```

## Activity: Part 3 — write your own

Write a single piped expression that:

1. Takes `penguins`,
2. Keeps only the rows where `species` is `"Adelie"`,
3. Picks just the first 5 rows.

::: {.callout-tip}
## Hint
You'll need `filter()` and `head()`. Chain them with `|>`.
:::

**To turn in:** Save your script. We'll check at the end of class.

## Activity debrief

Common things to watch for:

- **The pipe goes at the END of the line**, not the start. `x |>` then a new line — not `\n |> f()`.
- **The first argument is filled automatically.** Don't write `penguins |> filter(penguins, ...)` — drop the first `penguins`.
- **`+` vs. `|>` again.** dplyr verbs chain with `|>`; ggplot layers chain with `+`.
- **Parentheses still matter.** `f()` not `f`. The pipe doesn't change that.

## Wrap-up: what we covered today

- The pipe operator (`|>` or `%>%`) takes the value on the left and feeds it as the first argument to the function on the right.
- Pipes let you read code top-to-bottom in the order you'd actually do the work.
- Each step gets its own line; the pipe lives at the **end** of the previous line.
- `|>` and `%>%` are nearly interchangeable for our purposes.
- dplyr verbs chain with `|>`; ggplot layers chain with `+`.

## Wrap-up & looking ahead

**Monday:** dplyr verbs — `filter()`, `select()`, `arrange()`.

:::: {.columns}
::: {.column width="55%"}
**Upcoming Deadlines**

- **Today, 11:59 p.m.** — [HW 1: Basic R & Plots](https://psu.instructure.com/courses/2487468/assignments/18311694)
- **Mon 5/25, before class** — DC §7 reading (Canvas link TBA)
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*Data Computing*](https://dtkaplan.github.io/DataComputingEbook/)
:::
::::
