---
title: "Clustering & decision trees"
subtitle: "Monday, June 22, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day24-clustering-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day24-clustering-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(palmerpenguins)
library(rpart)

set.seed(184)
theme_set(theme_minimal(base_size = 15))

peng <- penguins |>
  drop_na(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Monday June 22]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Reading** (Canvas): *Data Computing* [§18.2 (clustering)](https://psu.instructure.com/courses/2487468/assignments/18325380)

**Plan for today**

- **Quiz 4** (rescheduled from Friday).
- **Supervised vs. unsupervised** learning.
- **k-means** and **hierarchical** clustering.
- **Decision trees / CART**.
- In-class activity.

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd) · [*DC* Ch. 18](https://dtkaplan.github.io/DataComputingEbook/chap-machine-learning.html)
:::

# Big picture {background-color="#001E44"}

## Two kinds of learning {.scrollable .nostretch}

| | **Unsupervised** | **Supervised** |
|---|---|---|
| You have... | features only, **no labels** | features **and** a known answer |
| Goal | find **structure / groups** | **predict** the answer for new cases |
| Examples | **clustering** (k-means, hierarchical) | **decision trees** (CART) |
| "Right answer"? | no; you judge usefulness | yes; you can measure accuracy |

Today, we'll be using the **Palmer penguins** body measurements to explore supervised and unsupervised learning techniques.

## Penguins data {.scrollable .nostretch}

```{r}
#| echo: TRUE
peng |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  glimpse()
```

For **clustering** we'll hide `species` and see if we can "rediscover" these categories. For **trees** we'll use `species` as the answer to predict.

# Clustering {background-color="#001E44"}

## Overview {.scrollable .nostretch}

**Clustering** puts rows into groups so that rows in a group are **similar** and rows in different groups are **different**, using only the feature columns.

- No labels go into the model, i.e., you don't tell it "species."
- "Similar" means **close in distance** once features are on a comparable scale.
- How we measure distance depends on the method. 

::: {.callout-warning}
**Scale first.** Body mass (~3000-6000 g) would swamp bill depth (~13-22 mm) in any distance calculation. We almost always want to standardize so every feature gets an equal vote.
:::

```{r}
#| echo: TRUE
X <- peng |>
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  scale() # subtract mean, divide by sd --> unit-less

round(colMeans(X), 14) # column means = zero
apply(X, 2, var) # column variances = one
```

## k-means clustering {.scrollable .nostretch}

`kmeans()` finds **k** cluster centers and assigns each point to its nearest one, iterating until things stop moving. **k** is a hyperparameter you pick.

```{r}
#| echo: TRUE
set.seed(184) # random algorithm; set a seed so results are reproducible
km <- kmeans(X, centers = 3, nstart = 25) # nstart = try 25 random starts, keep best
km$size # how many penguins per cluster
```

`nstart` matters here: k-means depends on random starting centers, so we try several and keep the best.

- `set.seed()` makes it reproducible (we'll get the same results if we re-run our code).

## Clustering results {.scrollable .nostretch}

```{r}
#| echo: FALSE
#| code-fold: TRUE

X_df <- as.data.frame(X)
X_df$Cluster <- as.factor(km$cluster)

centers_df <- as.data.frame(km$centers)
centers_df$Cluster <- as.factor(1:nrow(centers_df))

ggplot(X_df, aes(x = bill_length_mm, y = bill_depth_mm, color = Cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_point(data = centers_df, aes(fill = Cluster), color = "black",
             size = 8, shape = 21, stroke = 2.5, alpha = 0.7) +
             
  scale_color_manual(values = c("#E69F00", "#009E73", "#56B4E9")) +
  scale_fill_manual(values = c("#E69F00", "#009E73", "#56B4E9")) +
  labs(title = "K-Means Clustering on Penguins",
       subtitle = "Large open circles mark the center of each cluster",
       x = "Bill Length (Scaled)",
       y = "Bill Depth (Scaled)") +
  theme_minimal()
```

## Did it recover the species categories? {.scrollable .nostretch}

We clustered **without** species. Now compare the clusters to the real species:

```{r}
#| echo: FALSE
#| code-fold: TRUE
library(ggplot2)

conf_matrix <- table(cluster = km$cluster, species = peng$species)
heatmap_df <- as.data.frame(conf_matrix)

ggplot(heatmap_df, aes(x = species, y = cluster, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "white", high = "#3399FF") +
  labs(title = "Cluster Assignments vs. True Penguin Species",
       x = "True Species",
       y = "K-Means Cluster",
       fill = "Count") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = "bottom")
```

## Did it recover the species categories? {.scrollable .nostretch}

```{r}
#| echo: FALSE
#| code-fold: TRUE
#| layout-ncol: 2
#| fig-width: 5
#| fig-height: 4
peng |>
  mutate(cluster = factor(km$cluster)) |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = cluster, shape = species)) +
  geom_point(size = 2, alpha = .8) +
  labs(title = "Clusters (color) vs. true species (shape)", color = "cluster",
       x = "Bill length (Scaled)", y = "Bill depth (Scaled)") +
  scale_color_manual(values = c("#E69F00", "#009E73", "#56B4E9"),
                     name = "Cluster (Model Classification)") +
  scale_fill_manual(values = c("#E69F00", "#009E73", "#56B4E9"),
                    name = "Species (Ground Truth)") +
  theme(legend.position = "bottom")

X_df$True_Species <- peng$species
X_df$Predicted_Species <- NA
X_df$Predicted_Species[X_df$Cluster == "1"] <- "Adelie"
X_df$Predicted_Species[X_df$Cluster == "2"] <- "Chinstrap"
X_df$Predicted_Species[X_df$Cluster == "3"] <- "Gentoo"
X_df$Classification <- ifelse(X_df$True_Species == X_df$Predicted_Species, 
                              "Correct", "Misclassified")

ggplot(X_df, aes(x = bill_length_mm, y = bill_depth_mm, color = Classification)) +
  geom_point(size = 2, alpha = 0.8) +
  
  scale_color_manual(values = c("Correct" = "#000000", "Misclassified" = "#D55E00")) +
  
  labs(title = "K-Means Misclassifications",
       x = "Bill Length (Scaled)",
       y = "Bill Depth (Scaled)",
       color = "") + 
  theme_minimal() +
  theme(legend.position = "bottom")
```

Each cluster is mostly one species, found with **no labels**. (Gentoo separates cleanly; the other two overlap a bit.)

## How many clusters? The elbow {.scrollable .nostretch}

There's no "correct" **k**. A common heuristic is to plot total within-cluster spread as **k** grows, and look for the **elbow** where adding clusters stops helping. (Uses `purrr::map_dbl()` from last week.)

```{r}
#| echo: TRUE
#| fig-width: 7.5
#| fig-height: 3.4
tibble(k = 1:8) |>
  mutate(wss = map_dbl(k, ~ kmeans(X, centers = .x, nstart = 10)$tot.withinss)) |>
  ggplot(aes(k, wss)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:8) +
  labs(x = "k (number of clusters)", y = "Error",
       title = "Elbow around k = 3")
```

Here, "Error" is measured by the total within-cluster sum of squares:

$$\text{Error}_k = \sum_{i=1}^{N} ||\mathbf{x}_i - \boldsymbol{\mu}_{c(i)}||^2$$

Equivalently,

$$\text{Error}_k = \sum_{j=1}^{k} \sum_{\mathbf{x}_i \in C_j} ||\mathbf{x}_i - \boldsymbol{\mu}_j||^2$$

## Hierarchical clustering {.scrollable .nostretch}

Another approach is to not pick **k** up front. Instead, you can start with every point alone, then repeatedly **merge the two closest groups**, and record the whole tree (a **dendrogram**).

```{r}
#| echo: TRUE
#| fig-width: 9
#| fig-height: 3.4
d  <- dist(X) # pairwise distances
hc <- hclust(d, method = "complete")
plot(hc, labels = FALSE, main = "Dendrogram", xlab = "", sub = "")
rect.hclust(hc, k = 3, border = "#7C0A02") # cut into 3 groups
```

**Height** = how different two groups were when merged. **Cut** the tree at a height to get clusters, then you choose *after* seeing the structure.

## Hierarchical: cut and compare {.scrollable .nostretch}

```{r}
#| echo: TRUE
clusters_h <- cutree(hc, k = 3)
table(cluster = clusters_h, species = peng$species)
```

k-means needs **k** up front and is fast on big data; hierarchical shows the **whole nesting structure** and lets you decide **k** afterward.

# Decision trees (CART) {background-color="#001E44"}

## Now we *have* labels {.scrollable .nostretch}

Switch to **supervised**: use the measurements to **predict species**, and get a model a human can read.

A **decision tree** asks a sequence of yes/no questions about the features, splitting the data at each step, until each leaf is (mostly) one class. **CART** = Classification And Regression Trees.

```{r}
#| echo: TRUE
fit <- rpart(
  species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
  data = peng, method = "class"
)
```

## Read the tree {.scrollable .nostretch}

```{r}
#| echo: TRUE
#| fig-width: 9
#| fig-height: 4.2
plot(fit, margin = 0.1)
text(fit, use.n = TRUE, cex = 0.9)
```

Follow a penguin from the top: at each node, answer the question and go **left if yes**. The leaf is the predicted species. Notice it often needs only **two measurements**, which is much more interpretable.

## How does it pick the splits? {.scrollable .nostretch}

At each step the tree tries every feature and every cut point and keeps the split that makes the two sides **most "pure"** (closest to a single class).

- Purity is measured by **impurity** (Gini index): low impurity = the node is almost all one species.

- It splits **greedily**, top-down, then **stops** when splits no longer help enough.

::: {.callout-tip}
This is why trees are easy to explain: each split is just a threshold on one variable, like `flipper_length_mm < 206`.
:::

## How good is it? {.scrollable .nostretch}

Because species is known, we can **score** predictions with a confusion table:

```{r}
#| echo: TRUE
pred <- predict(fit, peng, type = "class")
table(prediction = pred, actual = peng$species)
mean(pred == peng$species) # overall accuracy
```

## Visualize with a heatmap

```{r}
#| echo: FALSE
#| code-fold: TRUE
pred <- predict(fit, peng, type = "class")
conf_matrix <- table(prediction = pred, actual = peng$species)
heatmap_df <- as.data.frame(conf_matrix)

ggplot(heatmap_df, aes(x = actual, y = prediction, fill = Freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "white", high = "#66CC99") +
  labs(title = "Model Predictions vs. True Penguin Species",
       subtitle = paste("Overall Accuracy:", 
                        round(mean(pred == peng$species) * 100, 1), "%"),
       x = "True Species",
       y = "Predicted Species",
       fill = "Count") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = "bottom")
```


::: {.callout-warning}
This is accuracy on the **same data the tree learned from**, which is overly optimistic. A more realistic evaluation uses a held-out **test set** (the model must perform on data it has never seen). Deep trees can **overfit**, i.e., memorize the training data, fail on new data.
:::

## Trees vs. clustering {.scrollable .nostretch}

:::: {.columns}
::: {.column width="50%"}
**Clustering (unsupervised)**

- No labels available
- Question: *"what natural groups exist?"*
- Output: a group ID per row
- Judged by usefulness, not accuracy
- e.g. customer segments, document topics
:::
::: {.column width="50%"}
**Decision tree (supervised)**

- Labels available
- Question: *"predict this known outcome"*
- Output: a prediction + a readable rule
- Judged by accuracy on held-out data
- e.g. spam / not-spam, species, churn
:::
::::

An unsupervised cluster is **not** a verified truth, it's a hypothesis you still have to check.

# Activity {background-color="#7C0A02"}

## In-class activity (~20 min) {.scrollable .nostretch}

Work in `day24-activity.R` (or a fresh script) on the `penguins` data.

1. **Cluster.** Standardize the four numeric measurements, run `kmeans()` with `centers = 2`, then `4`. How does `table(cluster, species)` change vs. `k = 3`?
2. **Elbow.** Rebuild the elbow plot using only **two** features. Does the elbow still say 3?
3. **Tree.** Fit a tree predicting `species` from **only** `bill_length_mm` and `bill_depth_mm`. Plot it and report accuracy. Did dropping two features hurt much?
4. **Discuss.** Which method would you prefer if the dataset had **no** `species` column at all? Why?

## Wrap-up & looking ahead {.scrollable .nostretch}

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- **Unsupervised** (clustering) finds groups with no labels; **supervised** (trees) predicts a known label.
- **k-means**: pick `k`, scale features, `nstart`, read the elbow.
- **Hierarchical**: merge-closest, read the dendrogram, cut afterward.
- **Trees**: greedy splits on single features $\rightarrow$ readable rules; score on **held-out** data, beware overfitting.

**Upcoming**

- **Tue Jun 23:** using R with LLMs · HW 5 due 11:59 p.m.
- **Wed Jun 24:** case study · Checkpoint 2 due
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

- [*DC* Ch. 18](https://dtkaplan.github.io/DataComputingEbook/chap-machine-learning.html)
- [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd)
:::
::::
