---
title: "Using R with LLMs"
subtitle: "Tuesday, June 23, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day25-r-with-llms-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day25-r-with-llms-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Tuesday June 23]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Reading** (Canvas): *Data Computing* [§18.3 (trees)](https://psu.instructure.com/courses/2487468/assignments/18325381)

**Plan for today**

- Three ways to use a **large language model (LLM)** with R: **chat**, **in the IDE**, and **from code**.
- How to **prompt well** for R help.
- Calling an LLM **from R** with the **`ellmer`** package.

**Helpful links**

- [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [ellmer](https://ellmer.tidyverse.org/) · [reprex](https://reprex.tidyverse.org/) · [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd)
:::

# Overview {background-color="#001E44"}

## LLMs are part of the R workflow now {.scrollable .nostretch}

An **LLM** (ChatGPT, Claude, Gemini, etc.) predicts likely text from a prompt. For an R user that's useful for:

- **Explaining errors**: paste a cryptic message, get a plain-English cause.
- **Drafting code**: "write a `ggplot` of X by Y, faceted by Z."
- **Refactoring**: "make this tidyverse instead of base R."
- **Learning**: "what does `across()` do, with a small example?"
- **As a data tool itself**: classify, extract, or summarize text *inside* a pipeline.

::: {.callout-tip}
The skill that matters most is **not** "getting an answer." It's judging whether the answer is **right**.
:::

## Three places you'll use one {.scrollable .nostretch}

| Where | Examples | Best for |
|---|---|---|
| **Chat window** | Claude, ChatGPT, Gemini | one-off questions, explaining errors, learning |
| **In the IDE** | Positron / RStudio assistants, GitHub Copilot | inline autocomplete, "fix this" on selected code |
| **From R code** | the **`ellmer`** package | doing the *same LLM call on every row* |

# Prompting well for R {background-color="#001E44"}

## Be specific {.scrollable .nostretch}

A vague prompt gets a vague (often wrong) answer. A good R prompt includes:

1. **The goal**: what the output should be.
2. **The data shape**: paste `str()`, `glimpse()`, or `dput(head(x))`.
3. **Anything you tried** and (if applicable) the **exact error**.
4. **Constraints**: "use the tidyverse," "base R only," "no new packages."

::: {.callout-warning}
**Don't paste private or sensitive data.** Anything you send leaves your machine and may be stored. Share a tiny **fake** example with the same structure, not the real dataset.
:::

## A reproducible example beats a paragraph {.scrollable .nostretch}

The single most useful thing you can hand an LLM (or a human, or Stack Overflow) is a **reprex**, which is a tiny, self-contained, runnable snippet that reproduces the problem.

```{r}
#| echo: TRUE
#| eval: FALSE
# install.packages("reprex")
library(reprex)

# copy this to your clipboard, then run reprex() 
# it runs the code and formats code + output together that's ready to paste:
reprex({
  library(dplyr)
  tibble(x = c(1, 2, NA)) |> summarise(m = mean(x)) # why is m NA?
})
```

Built-in datasets (`mtcars`, `iris`, `penguins`) make great reprexes. This helper can run them too.

## Weak prompt vs. strong prompt {.scrollable .nostretch}

:::: {.columns}
::: {.column width="50%"}
**Weak**

> "my ggplot doesn't work, fix it"

You'll get a generic guess that may not match your situation.
:::
::: {.column width="50%"}
**Stronger**

> "Using `penguins`, I want a scatter of `bill_length_mm` vs `flipper_length_mm` colored by `species`. My code `ggplot(penguins) + geom_point(bill_length_mm, flipper_length_mm)` errors with *'object not found'*. Use tidyverse to tackle this."

Goal + data + code + error + constraint.
:::
::::

# Calling an LLM from R: `ellmer` {background-color="#001E44"}

## What `ellmer` is {.scrollable .nostretch}

[`ellmer`](https://ellmer.tidyverse.org/) is the **tidyverse package** that provides a unified interface for accessing various LLMs from R:

- `chat_anthropic()`: Claude
- `chat_openai()`: GPT
- `chat_google_gemini()`: Gemini

```{r}
#| echo: TRUE
#| eval: FALSE
# install.packages("ellmer")
library(ellmer)

chat <- chat_anthropic(
  model = "claude-opus-4-8", # current default Claude model (RIP Fable 5)
  system_prompt = "You are a concise R tutor. Prefer the tidyverse."
)

chat$chat("In one sentence, what does dplyr::across() do?")
```

::: {.callout-note}
The newest, most capable Claude models are **Fable 5** (`claude-fable-5`) and the **Claude 4.x** family: **Opus 4.8** (`claude-opus-4-8`), **Sonnet 4.6** (`claude-sonnet-4-6`), **Haiku 4.5** (`claude-haiku-4-5`). Bigger models are smarter but cost more per call.
:::

## API keys: never in your code {.scrollable .nostretch}

Calling a provider needs an **API key** (a secret password tied to a paid account).

```{r}
#| echo: TRUE
#| eval: FALSE
# DON'T: a key pasted in a script gets committed to GitHub forever.
chat_anthropic(api_key = "sk-ant-REAL-KEY")

# INSTEAD: store it once in an environment variable ellmer can read automatically.
usethis::edit_r_environ()    # opens ~/.Renviron; add: ANTHROPIC_API_KEY=sk-ant-...
# restart R, then:
chat <- chat_anthropic(model = "claude-opus-4-8")  # key picked up from the env
```

::: {.callout-warning}
A leaked key can run up real charges. Keep it out of `.R`, `.qmd`, and anything you push to git (`.Renviron` is git-ignored by default).
:::

## LLM as a *data step* {.scrollable .nostretch}

The real power for us isn't chatting, but rather running the **same prompt over every row** of a data frame. This approach can help turn messy text into structured columns since `ellmer` can return **typed, structured output**:

```{r}
#| echo: TRUE
#| eval: FALSE
library(ellmer)
chat <- chat_anthropic(model = "claude-opus-4-8")

reviews <- tibble(text = c(
  "Loved it, arrived early and works great!",
  "Broke after two days. Very disappointed.",
  "It's fine. Does the job, nothing special."
))

# ask for a schema; ellmer enforces the shape
sentiment <- type_object(
  mood = type_string("one of: positive, negative, neutral"),
  reason = type_string("a 5-word justification")
)

reviews |>
  mutate(parsed = map(text, ~ chat$chat_structured(.x, type = sentiment))) |>
  unnest_wider(parsed)
```

This is `purrr` + a prompt: text in, tidy columns out. *That's* the data-science use.

# The catch {background-color="#7C0A02"}

## LLMs are confident, not always correct {.scrollable .nostretch}

An LLM generates **plausible**, not 100% true, text. You should expect:

- **Hallucinated functions / packages** like `dplyr::summarise_with()` that don't exist.
- **Subtly wrong code** that runs without error, gives the wrong number.
- **Outdated idioms**, i.e., deprecated `funs()`, old `gather()`/`spread()`.
- **Made-up facts and citations** stated with total confidence.

## Verification {.scrollable .nostretch}

For any code an LLM gives you:

::: {.callout-tip}
1. **Read it.** Can you explain every line?
2. **Run it.** Does it error on real input data? Does it even do what you asked?
3. **Check the result.** Spot-check against a value you can compute by hand or already know.
4. **Test the edge cases.** Things to try include empty input, `NA`s, one row, or a weird category.
:::

## Use it wisely {.scrollable .nostretch}

- **Learn from it, don't just paste.** Ask *why*, ask for a small example, then reproduce it yourself.
- **Keep prompts small and specific.** 
- **Prefer reprexes and built-in data.**
- **Track what you used.**

::: {.callout-note}
Treat LLMs like very fast, occasionally-wrong collaborators. They're not to be blindly trusted.
:::

# Activity {background-color="#7C0A02"}

## In-class activity (~20 min) {.scrollable .nostretch}

Use whatever LLM you have access to (chat window is fine, no API key needed).

1. **Explain an error.** Run this, then ask the LLM to explain *and* fix it. Verify the fix runs:
   ```r
   library(dplyr)
   penguins |> group_by(species) |> summarize(avg = mean(bill_length_mm))
   ```
   (Why `NaN`/`NA`? What's the one-argument fix?)
2. **Generate a function.** Ask for an R function that takes a numeric vector and returns its mean, median, and SD as a named tibble. Then **break it**: feed it `c(1, 2, NA)` and a length-1 vector. Did the LLM handle those?
3. **Catch a mistake.** Find **at least one** thing the LLM got wrong, outdated, or could improve.

## Wrap-up & looking ahead {.scrollable .nostretch}

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- Three surfaces: **chat**, **IDE assistant**, **`ellmer` from code**.
- **Prompt well**: goal + data shape + error + constraints; hand over a **reprex**, never private data.
- **`ellmer`** calls Claude/GPT/Gemini from R; keep API keys in `.Renviron`; structured output = LLM as a tidy **data step**.

**Upcoming**

- **Wed Jun 24:** benzoyl peroxide case study: data-driven investigation uncovering carcinogens in everyday skincare products· Checkpoint 2 due
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

- [ellmer docs](https://ellmer.tidyverse.org/) · [reprex](https://reprex.tidyverse.org/)
- [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd)
:::
::::
