---
title: "Using R with LLMs"
subtitle: "Tuesday, June 23, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day25-r-with-llms-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day25-r-with-llms-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Tuesday June 23]{style="font-size: 1.3em;"}

**Before class**

- No new reading. Skim the course [**AI policy**](../../syllabus/academic-integrity-and-ai-policy.qmd).

**Plan for today**

- Three ways to use a **large language model (LLM)** with R: **chat**, **in the IDE**, and **from code**.
- How to **prompt well** for R help (and give a reproducible example).
- Calling an LLM **from R** with the **`ellmer`** package.
- The non-negotiable habit: **verify everything** (yesterday's *and* tomorrow's theme).
- Doing all of this **within the course AI policy**.

**Helpful links**

- [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [ellmer](https://ellmer.tidyverse.org/) · [reprex](https://reprex.tidyverse.org/) · [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd)
:::

# Why this matters {background-color="#001E44"}

## LLMs are part of the R workflow now {.scrollable .nostretch}

An **LLM** (ChatGPT, Claude, Gemini, …) predicts likely text from a prompt. For an R user that's useful for:

- **Explaining errors** — paste a cryptic message, get a plain-English cause.
- **Drafting code** — "write a `ggplot` of X by Y, faceted by Z."
- **Refactoring** — "make this tidyverse instead of base R."
- **Learning** — "what does `across()` do, with a small example?"
- **As a data tool itself** — classify, extract, or summarize text *inside* a pipeline.

::: {.callout-tip}
The skill that matters is **not** "getting an answer." It's judging whether the answer is **right** — which is exactly what a data course trains you to do.
:::

## Three places you'll use one {.scrollable .nostretch}

| Where | Examples | Best for |
|---|---|---|
| **Chat window** | claude.ai, ChatGPT, Gemini | one-off questions, explaining errors, learning |
| **In the IDE** | Positron / RStudio assistants, GitHub Copilot | inline autocomplete, "fix this" on selected code |
| **From R code** | the **`ellmer`** package | doing the *same LLM call on every row* — a data step |

Today we touch all three, but spend the most time on the third — because that's the one unique to a programming course.

# Prompting well for R {background-color="#001E44"}

## Give it what a human helper would need {.scrollable .nostretch}

A vague prompt gets a vague (often wrong) answer. A good R prompt includes:

1. **The goal** — what the output should be.
2. **The data shape** — paste `str()`, `glimpse()`, or `dput(head(x))`.
3. **What you tried** and the **exact error**.
4. **Constraints** — "use the tidyverse," "base R only," "no new packages."

::: {.callout-warning}
**Don't paste private or sensitive data.** Anything you send leaves your machine and may be stored. Share a tiny **fake** example with the same structure, not the real dataset — especially anything with names, grades, or PII.
:::

## A reproducible example beats a paragraph {.scrollable .nostretch}

The single most useful thing you can hand an LLM (or a human, or Stack Overflow) is a **reprex** — a tiny, self-contained, runnable snippet that reproduces the problem.

```{r}
#| echo: TRUE
#| eval: FALSE
# install.packages("reprex")
library(reprex)

# copy this to your clipboard, then run reprex() — it runs the code and
# formats code + output together, ready to paste:
reprex({
  library(dplyr)
  tibble(x = c(1, 2, NA)) |> summarise(m = mean(x))   # why is m NA?
})
```

Built-in datasets (`mtcars`, `iris`, `penguins`) make great reprexes — the helper can run them too.

## Weak prompt vs. strong prompt {.scrollable .nostretch}

:::: {.columns}
::: {.column width="50%"}
**Weak** ❌

> "my ggplot doesn't work, fix it"

No data, no code, no error. You'll get a generic guess that probably doesn't match your situation.
:::
::: {.column width="50%"}
**Strong** ✅

> "Using `penguins`, I want a scatter of `bill_length_mm` vs `flipper_length_mm` colored by `species`. My code `ggplot(penguins) + geom_point(bill_length_mm, flipper_length_mm)` errors with *'object not found'*. Tidyverse please."

Goal + data + code + error + constraint.
:::
::::

# Calling an LLM from R: `ellmer` {background-color="#001E44"}

## What `ellmer` is {.scrollable .nostretch}

[`ellmer`](https://ellmer.tidyverse.org/) is the **tidyverse package** for talking to LLMs from R. One interface, many providers:

- `chat_anthropic()` — Claude
- `chat_openai()` — GPT
- `chat_google_gemini()` — Gemini

```{r}
#| echo: TRUE
#| eval: FALSE
# install.packages("ellmer")
library(ellmer)

chat <- chat_anthropic(
  model = "claude-opus-4-8",                 # current default Claude model
  system_prompt = "You are a concise R tutor. Prefer the tidyverse."
)

chat$chat("In one sentence, what does dplyr::across() do?")
```

::: {.callout-note}
The newest, most capable Claude models are **Fable 5** (`claude-fable-5`) and the **Claude 4.x** family — **Opus 4.8** (`claude-opus-4-8`), **Sonnet 4.6** (`claude-sonnet-4-6`), **Haiku 4.5** (`claude-haiku-4-5`). Bigger models are smarter but cost more per call.
:::

## API keys: never in your code {.scrollable .nostretch}

Calling a provider needs an **API key** (a secret password tied to a paid account). Treat it like one.

```{r}
#| echo: TRUE
#| eval: FALSE
# DON'T: a key pasted in a script gets committed to GitHub forever.
chat_anthropic(api_key = "sk-ant-REAL-KEY")        # ❌

# DO: store it once in an environment variable; ellmer reads it automatically.
usethis::edit_r_environ()    # opens ~/.Renviron — add: ANTHROPIC_API_KEY=sk-ant-...
# restart R, then:
chat <- chat_anthropic(model = "claude-opus-4-8")  # ✅ key picked up from the env
```

::: {.callout-warning}
A leaked key can run up real charges. Keep it out of `.R`, `.qmd`, and anything you push to git (`.Renviron` is git-ignored by default).
:::

## The interesting part: an LLM as a *data step* {.scrollable .nostretch}

The real power for us isn't chatting — it's running the **same prompt over every row** of a data frame, turning messy text into structured columns. `ellmer` can return **typed, structured output**:

```{r}
#| echo: TRUE
#| eval: FALSE
library(ellmer)
chat <- chat_anthropic(model = "claude-opus-4-8")

reviews <- tibble(text = c(
  "Loved it, arrived early and works great!",
  "Broke after two days. Very disappointed.",
  "It's fine. Does the job, nothing special."
))

# ask for a schema; ellmer enforces the shape
sentiment <- type_object(
  mood   = type_string("one of: positive, negative, neutral"),
  reason = type_string("a 5-word justification")
)

reviews |>
  mutate(parsed = map(text, ~ chat$chat_structured(.x, type = sentiment))) |>
  unnest_wider(parsed)
```

This is `purrr` + a prompt: text in, tidy columns out. *That's* the data-science use.

# The catch {background-color="#7C0A02"}

## LLMs are confident, not correct {.scrollable .nostretch}

An LLM generates **plausible** text. Plausible is not the same as true. Expect:

- **Hallucinated functions / packages** — `dplyr::summarise_with()` that doesn't exist.
- **Subtly wrong code** — runs without error, gives the wrong number.
- **Outdated idioms** — deprecated `funs()`, old `gather()`/`spread()`.
- **Made-up facts and citations** — stated with total confidence.

::: {.callout-important}
This is the **same lesson as tomorrow's case study**: a thing can *sound* authoritative and still be wrong. You are the reviewer.
:::

## The verify loop (non-negotiable) {.scrollable .nostretch}

For any code an LLM gives you:

::: {.callout-tip}
1. **Read it.** Can you explain every line? If not, don't ship it.
2. **Run it.** On real input. Does it error? Does it even do what you asked?
3. **Check the result.** Spot-check against a value you can compute by hand or already know.
4. **Test the edges.** Empty input, `NA`s, one row, a weird category.
:::

If you can't verify it, you can't use it — for a grade, a paper, or a job.

## Use it well {.scrollable .nostretch}

- **Learn from it, don't just paste.** Ask *why*, ask for a small example, then reproduce it yourself.
- **Keep prompts small and specific.** One problem at a time beats "build my whole project."
- **Prefer reprexes and built-in data.** Better answers, no privacy risk.
- **Track what you used.** You'll need it for the next slide.

# Academic integrity {background-color="#001E44"}

## What the course policy asks {.scrollable .nostretch}

Using an LLM is **allowed within the course [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd)** — read it, but the spirit is simple:

- **You are responsible** for everything you submit, AI-assisted or not. "The AI wrote it" is not a defense for a bug or a wrong claim.
- **Disclose** AI use where the policy asks (and don't pass AI text off as your own writing).
- **You must understand your submission** — be ready to explain any line in office hours or an oral check-in.
- On a **quiz/exam**, only the tools the policy permits.

::: {.callout-note}
Treat it like a very fast, occasionally-wrong collaborator: great for momentum, never the final authority. The judgment stays yours.
:::

# Activity {background-color="#7C0A02"}

## In-class activity (~20 min) {.scrollable .nostretch}

Use whatever LLM you have access to (chat window is fine — no API key needed).

1. **Explain an error.** Run this, then ask the LLM to explain *and* fix it. Verify the fix runs:
   ```r
   library(dplyr)
   penguins |> group_by(species) |> summarize(avg = mean(bill_length_mm))
   ```
   (Why `NaN`/`NA`? What's the one-argument fix?)
2. **Generate a function.** Ask for an R function that takes a numeric vector and returns its mean, median, and SD as a named tibble. Then **break it**: feed it `c(1, 2, NA)` and a length-1 vector. Did the LLM handle those?
3. **Catch a mistake.** Find **at least one** thing the LLM got wrong, outdated, or could improve — and write one sentence on how you knew.

## Wrap-up & looking ahead {.scrollable .nostretch}

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- Three surfaces: **chat**, **IDE assistant**, **`ellmer` from code**.
- **Prompt well**: goal + data shape + error + constraints; hand over a **reprex**, never private data.
- **`ellmer`** calls Claude/GPT/Gemini from R; keep API keys in `.Renviron`; structured output = LLM as a tidy **data step**.
- **Verify everything** — read, run, check, test. Confident ≠ correct.
- Stay inside the **course AI policy**; you own what you submit.

**Upcoming**

- **Tonight:** HW 5 due 11:59 p.m.
- **Wed Jun 24:** case study — bad science · Checkpoint 2 due
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

- [ellmer docs](https://ellmer.tidyverse.org/) · [reprex](https://reprex.tidyverse.org/)
- [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd)
:::
::::
