---
title: "Midterm exam review"
subtitle: "Wednesday, June 10, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day17-exam-review-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day17-exam-review-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(palmerpenguins)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Wednesday June 10]{style="font-size: 1.3em;"}

**Plan for today**

- Exam is **tomorrow, Thursday June 11, in class**
- Review
- Practice problems

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 1–5](https://r4ds.hadley.nz) · [*DC* Ch. 3–5, 11–12](https://dtkaplan.github.io/DataComputingEbook/)
:::

# The exam {background-color="#7C0A02"}

## Logistics {.scrollable .nostretch}

- **When:** Thursday June 11, start of class (9:35 a.m.)
- **How long:** 75 minutes
- **Points:** 100 total, 8 parts (A-H)
- **Closed book**

You may assume `tidyverse` and `palmerpenguins` are loaded. You'll see these datasets: `penguins`, `mpg`, `diamonds`, `msleep`.

**Format breakdown:**

| Part | Type | Points |
|---|---|---|
| A | Multiple choice (12 items) | 24 |
| B | Predict the output | 12 |
| C | Code anatomy | 10 |
| D | Code writing | 20 |
| E | Position adjustments | 8 |
| F | Facets and scales | 6 |
| G | Tidy data | 8 |
| H | Short answer & debugging | 12 |

## Study advice {.scrollable .nostretch}

The exam tests whether you can **read, write, and reason about** code. Some study suggestions:

- For multiple choice, know *why* each answer is correct or incorrect, not just the letter
- For predict-the-output, run the expressions yourself and understand the rule
- For code writing, practice writing from scratch instead of just reading
- For debugging think about "what rule does this code violate?"

# R basics {background-color="#001E44"}

## Assignment and objects {.scrollable .nostretch}

The standard assignment operator is `<-`. `=` works too, but `<-` is the course convention and the idiomatic R style.

```{r}
#| echo: TRUE
x <- 7.3 # preferred
x

x = 8.1 # works, not preferred
x

1.2 -> x # also works, really not preferred
x
```

`==` is a **comparison**, not assignment; it tests whether two things are equal and returns `TRUE` or `FALSE`.

```{r}
#| echo: TRUE
x == 1.2     # is x equal to 1.2?
x == pi    # is x equal to pi (~3.14)?
```

## install.packages() vs library() {.scrollable .nostretch}

| | `install.packages()` | `library()` |
|---|---|---|
| What it does | Downloads the package to your computer | Loads it into the current session |
| When to run | **Once per computer** | **Every session** |
| In a script? | Never; do it in the Console | Yes: at the top |

::: {.callout-warning}
Installed once $\neq$ available always. You must call `library()` at the start of every new R session.
:::

## Data frames {.scrollable .nostretch}

A **data frame** is a two-dimensional table:

- Each **row** is one case/observation
- Each **column** is one variable
- Columns can each hold a **different type** (numeric, character, logical, factor, etc.)
- Within a column, every value is the same type

This is what distinguishes a data frame from a **matrix** (all values must be the same type) and from a **vector** (one-dimensional).

```{r}
#| echo: TRUE
glimpse(penguins)  # see the structure: rows, columns, types
```

## Named vs positional arguments {.scrollable .nostretch}

Functions take **arguments**. You can pass them **positionally** (by order) or **by name**.

```{r}
#| echo: TRUE
round(3.14159, 2)              # positional: x first, digits second
round(x = 3.14159, digits = 2) # named
round(digits = 2, x = 3.14159) # named; shows that order doesn't matter
```

Named arguments are safer and more readable, but both styles are correct R.

## Arithmetic and vectors {.scrollable .nostretch}

R operates **element-wise** on vectors of the same length:

```{r}
#| echo: TRUE
c(1, 2, 3) + c(10, 20, 30)
c(2, 4, 6) * 3
```

Key operators to know:

| Expression | What it does |
|---|---|
| `%%` | Modulo (remainder after division) |
| `%/%` | Integer division |
| `^` | Exponent |

```{r}
#| echo: TRUE
17 %% 5  # remainder when 17 is divided by 5
17 %/% 5 # how many times does 5 go into 17?
```

## Practice: predict the output {.scrollable .nostretch}

Write your answer before clicking.

**(a)** `sqrt(81)`

::: {.fragment}
**9**
:::

**(b)** `c(3, 6, 9) - c(1, 2, 3)`

::: {.fragment}
**2 4 6**: element-wise subtraction
:::

**(c)** `10 %% 3`

::: {.fragment}
**1**: 10 = 3 × 3 + 1
:::

**(d)** `seq(0, 12, by = 3)`

::: {.fragment}
**0 3 6 9 12**
:::

**(e)** `length(c("alpha", "beta", "gamma", "delta"))`

::: {.fragment}
**4**: four elements in the vector
:::

**(f)** `mean(c(5, 10, 15, 20), na.rm = TRUE)`

::: {.fragment}
**12.5**: (5+10+15+20)/4
:::

## NA and na.rm {.scrollable .nostretch}

`NA` in R means "not available", i.e., a missing value. Almost any calculation with `NA` returns `NA`:

```{r}
#| echo: TRUE
mean(c(2, 4, NA, 8))               # NA
mean(c(2, 4, NA, 8), na.rm = TRUE) # drop NAs first
```

Any function that summarizes values across a vector, such as `mean()`, `sum()`, `min()`, `max()`, `sd()`, accepts `na.rm = TRUE`.

# Grammar of graphics {background-color="#001E44"}

## Mapping vs setting aesthetics {.scrollable .nostretch}

| | Inside `aes()` | Outside `aes()` |
|---|---|---|
| What it does | **Maps** a variable to a visual property | **Sets** a fixed visual property |
| Changes by row? | Yes; different rows get different values | No; every point gets the same value |
| Example | `aes(color = species)` | `color = "red"` |

```{r}
#| echo: TRUE
#| eval: FALSE
# color varies by species (mapping):
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species))

# every point is red (setting):
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = "red")
```

::: {.callout-warning}
`geom_point(aes(color = "red"))` maps the *string* `"red"` as if it were a variable. You get one default color and a legend that says "red". **Set constants outside `aes()`.**
:::

## Glyphs {.scrollable .nostretch}

In the **grammar of graphics**, a **glyph** is the basic mark that represents one case.

- In a scatterplot: one dot per row
- In a bar chart: one bar segment per group
- In a line plot: one point connected by a line segment

The glyph is what `geom_*` functions draw. Each row of data produces one glyph (or contributes to one group glyph).

## Choosing the right geom {.scrollable .nostretch}

| Goal | Appropriate Geom |
|---|---|
| Distribution of one quantitative variable | `geom_histogram()` or `geom_density()` |
| Distribution of one categorical variable | `geom_bar()` |
| Relationship between two quantitative variables | `geom_point()` |
| Trend over time or ordered x | `geom_line()` |
| Relationship between quant and categorical | `geom_boxplot()` |
| Counts of two categoricals | `geom_bar()` + fill + position |

## Zooming: coord_cartesian() vs xlim() {.scrollable .nostretch}

Both zoom in on a plot, but the difference matters when you have fitted curves:

- `xlim(0, 50)` **drops** data outside the range before any computation, which changes what a smoother sees
- `coord_cartesian(xlim = c(0, 50))` only adjusts the **viewport**. Here, all data is still used (you're just zooming in).

```{r}
#| echo: TRUE
#| eval: FALSE
# Safer zoom since it doesn't affect statistics:
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(1.5, 4))
```

::: {.callout-tip}
Default: use `coord_cartesian()` when you want to zoom without affecting calculations.
:::

## Position adjustments {.scrollable .nostretch}

With `geom_bar(aes(x = drv, fill = class))`:

```{r}
#| echo: FALSE
#| fig-height: 2.5
#| fig-width: 9
p <- ggplot(mpg, aes(x = drv, fill = class))

p1 <- p + geom_bar(position = "stack") + labs(title = '"stack" (default)')  + theme(legend.position = "none")
p2 <- p + geom_bar(position = "dodge") + labs(title = '"dodge"') + theme(legend.position = "none")
p3 <- p + geom_bar(position = "fill") + labs(title = '"fill"') + theme(legend.position = "none")

library(patchwork)
p1 | p2 | p3
```

| Position | What it does |
|---|---|
| `"stack"` | Bars stacked; total height = total count |
| `"dodge"` | Bars side by side within each group |
| `"fill"` | Bars all same height; shows **proportions** |

`"fill"` is best when you want to **compare proportions** across groups.

## Facets {.scrollable .nostretch}

Faceting splits the plot into panels, one per level of a variable.

```{r}
#| echo: TRUE
#| eval: FALSE
facet_wrap(~ class)    # one variable; auto-wraps into a grid
facet_grid(drv ~ year) # drv in ROWS, year in COLUMNS
```

Key rules for `facet_grid(row_var ~ col_var)`:

- Left of `~` $\rightarrow$ **rows**
- Right of `~` $\rightarrow$ **columns**
- Number of panels = levels(row_var) $\times$ levels(col_var)
- By default all panels **share the same axis scales**

```{r}
#| echo: TRUE
#| eval: FALSE
# 3 drv values × 2 year values = 6 panels
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(bins = 15) +
  facet_grid(drv ~ year)
```

## Practice: ggplot debugging {.scrollable .nostretch}

What is wrong with this code? There are two separate problems.

```{r}
#| echo: TRUE
#| eval: FALSE
ggplot(diamonds, aes(x = carat, y = price))
  geom_point()
```

::: {.fragment}
**Problem 1:** The `ggplot()` line is missing a trailing `+`. Without it, R treats the two lines as separate, unrelated expressions.

**Problem 2:** If `tidyverse` (or `ggplot2`) was never loaded with `library()` in this session, `geom_point` won't be found, even if the package is installed.

Fixed:
```{r}
#| echo: TRUE
#| eval: FALSE
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```
:::

## Practice: skewed data {.scrollable .nostretch}

A scatterplot of `msleep` with `bodywt` on the x-axis and `sleep_total` on the y-axis shows almost all points crammed into the bottom-left corner, with one or two animals far to the right.

**(a)** Why does this happen?

::: {.fragment}
`bodywt` is extremely right-skewed, i.e., a few very large animals (elephants, giraffes) pull the x-axis far right, compressing everything else into the left.
:::

**(b)** Write one line of code to fix it:

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: FALSE
+ scale_x_log10()
```
A log scale spaces the values multiplicatively so the small animals spread out instead of piling up.
:::

# dplyr verbs {background-color="#001E44"}

## The six core verbs {.scrollable .nostretch}

| Verb | What it does | Rows changed? | Columns changed? |
|---|---|---|---|
| `filter()` | Keep rows matching a condition | Yes; fewer rows | No |
| `select()` | Keep (or drop) columns | No | Yes; fewer columns |
| `mutate()` | Add or modify a column | No | Yes; more/changed columns |
| `arrange()` | Sort rows | Yes; reordered | No |
| `group_by()` | Tag groups for later | No visible change | No |
| `summarise()` | Collapse to one row per group | Yes; fewer rows | Yes; new summary columns |

`group_by()` + `summarise()` always go together.

## Reduction functions {.scrollable .nostretch}

A **reduction** (or summary) function takes a whole column (vector) and returns **one value**:

```{r}
#| echo: TRUE
#| eval: FALSE
mean(c(10, 20, 30))
sum(c(10, 20, 30))
n() # only works inside summarise(): counts rows in the group
```

Compare to a **transformation** function — it returns one value *per row*:

```{r}
#| echo: TRUE
c(10, 20, 30) * 2  # three inputs, three outputs
log(c(10, 20, 30)) # still three outputs
```

Use reductions in `summarise()`. Use transformations in `mutate()`.

## Pipeline order matters {.scrollable .nostretch}

The pipe (`|>`) feeds the output of one step into the next. **Order determines what each step sees.**

::: {.callout-warning}
You cannot filter on a column that `summarise()` already collapsed away. Always `filter()` first when the condition is on individual rows.
:::

```{r}
#| echo: TRUE
#| eval: FALSE
# CORRECT: filter rows first, then summarise
penguins |>
  filter(body_mass_g > 4000) |>
  summarise(avg = mean(body_mass_g))

# WRONG: summarise removes the rows — nothing left to filter
penguins |>
  summarise(avg = mean(body_mass_g)) |>
  filter(body_mass_g > 4000) # body_mass_g no longer exists!
```

## Reading code: anatomy of a pipeline {.scrollable .nostretch}

```{r}
#| echo: TRUE
#| eval: FALSE
mpg |>
  filter(class == "compact") |>
  group_by(manufacturer) |>
  summarise(avg_hwy = mean(hwy))
```

Be able to identify each piece:

- **Data table:** `mpg`
- **Functions:** `filter`, `group_by`, `summarise`, `mean`, `==` (operator)
- **Variables (columns):** `class`, `manufacturer`, `hwy`, and the new `avg_hwy`
- **Constants:** `"compact"` (a character string literal)
- **Verb that keeps only some rows:** `filter()`

## Practice: write a pipeline {.scrollable .nostretch}

Write a pipeline on `penguins` that:

1. Keeps only rows where `sex` is not `NA`
2. Groups by `species`
3. Reports the count (`n`) and mean flipper length (`avg_flipper`) per species
4. Sorts from longest to shortest average flipper

Try it before looking.

::: {.fragment}
```{r}
#| echo: TRUE
penguins |>
  filter(!is.na(sex)) |>
  group_by(species) |>
  summarise(n = n(), avg_flipper = mean(flipper_length_mm, na.rm = TRUE)) |>
  arrange(desc(avg_flipper))
```
:::

## Practice: pipes and nested calls {.scrollable .nostretch}

**Nested $\rightarrow$ pipe:** Rewrite with `|>`, one step per line.

```{r}
#| echo: TRUE
#| eval: FALSE
head(arrange(filter(mpg, year == 2008), desc(hwy)), 3)
```

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: FALSE
mpg |>
  filter(year == 2008) |>
  arrange(desc(hwy)) |>
  head(3)
```
The innermost call becomes the first step. `head()` is outermost, so it goes last.
:::

**Pipe $\rightarrow$ nested:** Rewrite without the pipe.

```{r}
#| echo: TRUE
#| eval: FALSE
penguins |>
  filter(!is.na(bill_length_mm)) |>
  summarise(avg_bill = mean(bill_length_mm))
```

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: FALSE
summarise(filter(penguins, !is.na(bill_length_mm)), avg_bill = mean(bill_length_mm))
```
:::

## The over-grouping pitfall {.scrollable .nostretch}

```{r}
#| echo: TRUE
penguins |>
  group_by(species, island, sex) |>
  summarise(avg_mass = mean(body_mass_g, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(avg_mass)) |>
  head(3)
```

The top row has the highest average, but how many penguins is it based on?

::: {.fragment}
Grouping by three variables (species + island + sex) can produce groups with very few penguins. A high average from 2 penguins is not trustworthy.

**Fix:** always add `n = n()` inside `summarise()` so you can see the group size.
:::

# Tidy data {background-color="#001E44"}

## Wide vs narrow (long) {.scrollable .nostretch}

The **same data** can be stored in two shapes:

**Wide**: one row per student, one column per exam:

```
student   midterm   final
Ava            88      91
Ben            72      85
Cy             95      90
```

**Narrow (long)**: one row per measurement:

```
student   exam       score
Ava       midterm       88
Ava       final         91
Ben       midterm       72
Ben       final         85
Cy        midterm       95
Cy        final         90
```

Wide has **3 rows** and **3 columns**. Narrow has **6 rows** and **3 columns**. They contain the same data, just in a different shape.

## pivot_longer() and pivot_wider() {.scrollable .nostretch}

```{r}
#| echo: TRUE
#| eval: FALSE
# Wide to narrow (long)
wide_scores |>
  pivot_longer(
    cols = c(midterm, final), # which columns to stack
    names_to  = "exam",       # new column holding the old names
    values_to = "score"       # new column holding the values
  )

# Narrow to wide
narrow_scores |>
  pivot_wider(
    names_from  = exam, # column whose values become new column names
    values_from = score # column whose values fill the new cells
  )
```

**Memory trick:** `pivot_longer()` makes the table *taller* (more rows). `pivot_wider()` makes it *wider* (more columns). They are exact inverses.

## Practice: identify the pivot {.scrollable .nostretch}

**(a)** You have one row per country and columns `pop_2000`, `pop_2010`, `pop_2020`. You want to plot population over time. Should you `pivot_longer()` or `pivot_wider()`?

::: {.fragment}
**`pivot_longer()`**: the years are trapped in column names and need to become a `year` variable so you can put it on the x-axis.
:::

**(b)** You have a long table with one row per (student, subject) pair. You want to compute each student's math score minus their reading score. Should you `pivot_longer()` or `pivot_wider()`?

::: {.fragment}
**`pivot_wider()`**: put each subject in its own column so that math and reading are side by side in the same row, enabling a `mutate(diff = math - reading)`.
:::

# Putting it together {background-color="#001E44"}

## Writing a ggplot {.scrollable .nostretch}

The template:

```{r}
#| echo: TRUE
#| eval: FALSE
ggplot(DATA, aes(x = XVAR, y = YVAR)) +
  geom_SOMETHING(aes(MAPPED_AESTHETICS), FIXED_AESTHETICS)
```

**Mapped** (inside `aes()`): changes row by row based on a column
**Fixed** (outside `aes()`): same for every point/bar/line

Practice: Write a scatterplot of `penguins` with `bill_length_mm` on x, `bill_depth_mm` on y, points **colored by** `species`, and a **fixed** size of 2.

::: {.fragment}
```{r}
#| echo: TRUE
#| fig-height: 2.8
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species), size = 2)
```

`color = species` is a mapping $\rightarrow$ inside `aes()`.
`size = 2` is a constant $\rightarrow$ outside `aes()`.
:::

## Quick-reference: things to know cold {.scrollable .nostretch}

**R basics**

- `<-` is assignment; `==` is equality test
- `install.packages()` once; `library()` every session
- Data frame: 2D, columns can have different types
- `na.rm = TRUE` to ignore NAs in summary functions

**ggplot**

- `aes()` maps data; constants go outside `aes()`
- `geom_histogram` for distribution, `geom_point` for scatter, `geom_bar` for counts
- `coord_cartesian()` to zoom without dropping data
- `facet_grid(row ~ col)`: rows left of `~`, columns right
- `position = "fill"` for proportions

**dplyr**

- `filter` rows, `mutate` columns, `summarise` collapses with `group_by`
- `filter()` must come before `summarise()` if the condition is on individual rows
- Always add `n = n()` to summaries so you know the group size
- `arrange(desc(...))` for descending order

**Tidy data**

- `pivot_longer()`: wide to narrow (more rows)
- `pivot_wider()`: narrow to wide (more columns)

# Good luck tomorrow {background-color="#7C0A02"}

## Exam {.nostretch}

**Thursday June 11, 9:35 a.m.**

Bring a pen or pencil.

If you finish early, double-check your code writing answers.

If something is unclear during the exam, please raise your hand.
