---
title: "Importing & cleaning data"
subtitle: "Tuesday, June 9, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day16-import-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day16-import-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Tuesday June 9]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Assignment**: [*DC* §16.1--16.2, 16.4--16.5](https://psu.instructure.com/courses/2487468/assignments/18325374)

**Plan for today**

- `read_csv()`
- Data cleaning: NA values, column names, column types
- Other file formats and reading from URLs
- Writing data
- Dates with lubridate
- In-class activity

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 7](https://r4ds.hadley.nz/data-import) · [*DC* Ch. 16](https://dtkaplan.github.io/DataComputingEbook/chap-scraping.html) · [readr reference](https://readr.tidyverse.org/reference)
:::

# Getting data into R {background-color="#001E44"}

## Pre-loaded data {.scrollable .nostretch}

Every dataset we've used this semester was already in R: `penguins`, `flights`, `diamonds`. That won't be true in real work.

Real data lives in files, including CSV spreadsheets, Excel workbooks, tab-delimited exports from survey software, files someone emails you.

Today's goal: get a file off disk (or the web) and into a tidy data frame that we can easily work with.

::: {.callout-note}
`readr` is part of the tidyverse, so `library(tidyverse)` is all you need for CSV and most text files.
:::

## read_csv() {.scrollable .nostretch}

The most common file type you'll encounter is **CSV** (comma-separated values), which can be read in using`read_csv()`:

```{r}
#| echo: TRUE
#| eval: FALSE
students <- read_csv("data/students.csv")
```

We can also read in CSV files directly from a URL, without having to download anything:

```{r}
#| echo: TRUE
#| message: TRUE
students <- read_csv("https://pos.it/r4ds-students-csv")
```

After reading in the CSV, readr prints a **column specification message** that tells you how many rows and columns it found, and what type it guessed for each column.

## Column specification message {.scrollable .nostretch}

```
Rows: 6 Columns: 5
── Column specification ─────────────────
Delimiter: ","
chr (4): Full Name, favourite.food, mealPlan, AGE
dbl (1): Student ID
```

- `chr` — character string
- `dbl` — double (numeric)
- `lgl` — logical (TRUE/FALSE)
- `dttm` / `date` — date-time / date

readr samples up to 1,000 rows and works through a set of rules to guess each column's type. It's usually right, but not always. The rest of today is largely about what to do when it's not.

## What we actually got {.scrollable .nostretch}

```{r}
#| echo: TRUE
students
```

A few things to fix:

- `favourite.food` has `"N/A"` as a string, not a real `NA`
- `Student ID` and `Full Name` have spaces, which are awkward to type
- `AGE` is a character column even though it's mostly numbers (`"five"` snuck in)
- `mealPlan` is categorical and should be a factor

## Problem 1: NA values {.scrollable .nostretch}

By default, readr only treats empty cells as `NA`. Strings like `"N/A"`, `"."`, `"missing"`, or `"-"` are left as-is.

Use the `na` argument to tell readr what else counts as missing:

```{r}
#| echo: TRUE
students <- read_csv(
  "https://pos.it/r4ds-students-csv",
  na = c("N/A", "")
)
students
```

`favourite.food` row 3 is now `NA` instead of the string `"N/A"`.

## Problem 2: column names {.scrollable .nostretch}

Columns with spaces are **non-syntactic**, and you can't type them without backticks:

```{r}
#| echo: TRUE
#| eval: FALSE
students$Student ID   # Error
students$`Student ID` # Works, but annoying
```

Two fixes. Rename manually:

```{r}
#| echo: TRUE
#| eval: FALSE
students |> rename(student_id = `Student ID`, full_name = `Full Name`)
```

Or use `janitor::clean_names()` to convert **everything** to snake_case at once:

```{r}
#| echo: TRUE
students |> janitor::clean_names()
```

`janitor` isn't part of the tidyverse; run `install.packages("janitor")` once.

## Problem 3: wrong column types {.scrollable .nostretch}

`mealPlan` is categorical, but we want to make it a factor. `age` is numeric, so we need adjust the (incorrect) character string `"five"` to be numeric first, then parse:

```{r}
#| echo: TRUE
students <- students |>
  janitor::clean_names() |>
  mutate(
    meal_plan = factor(meal_plan),
    age = parse_number(if_else(age == "five", "5", age)) # vectorized ifelse()
  )
students
```

`if_else(test, yes, no)`: where `age == "five"`, substitute `"5"`; otherwise leave it alone. Then `parse_number()` converts the string to a number.

::: {.callout-tip}
`parse_number()` is also great for values like `"$1,234"` or `"42%"` since it strips non-numeric characters automatically.
:::

# Column type detection {background-color="#001E44"}

## How readr guesses {.scrollable .nostretch}

For each column, readr samples up to 1,000 values and asks, in order:

1. Only `TRUE`, `FALSE`, `T`, `F`? $\rightarrow$ **logical**
2. Only numbers? $\rightarrow$ **double**
3. Matches ISO 8601 (`2026-06-09`)? $\rightarrow$ **date**
4. Otherwise $\rightarrow$ **character**

```{r}
#| echo: TRUE
read_csv("
  logical,numeric,date,string
  TRUE,1,2021-01-15,abc
  false,4.5,2021-02-15,def
  T,Inf,2021-02-16,ghi
")
```

This works well with clean data, but fails when unexpected values like `"N/A"` or `"five"` are present.

## Overriding types: col_types {.scrollable .nostretch}

We can tell readr what type to expect with the `col_types` argument:

```{r}
#| echo: TRUE
#| eval: FALSE
read_csv(
  file,
  col_types = list(
    student_id = col_integer(),
    meal_plan  = col_factor(),
    age        = col_double()
  )
)
```

Available column types:

| Function | Type |
|---|---|
| `col_double()` / `col_integer()` | numbers |
| `col_character()` | always a string (good for IDs, phone numbers) |
| `col_logical()` | TRUE/FALSE |
| `col_factor()` | categorical |
| `col_date()` / `col_datetime()` | dates |
| `col_skip()` | drop this column entirely |

## When guessing goes wrong: problems() {.scrollable .nostretch}

If you force a type and readr finds values that don't fit, it warns you and stores the details:

```{r}
#| echo: TRUE
simple_csv <- "x\n10\n.\n20\n30"

df <- read_csv(simple_csv, col_types = list(x = col_double()))
problems(df)
```

Row 3 has `"."` where readr expected a number. Now that we have identified the culprit, we can fix it:

```{r}
#| echo: TRUE
read_csv(simple_csv, na = ".")
```

::: {.callout-tip}
The workflow: read with forced types $\rightarrow$ `problems()` to locate oddities $\rightarrow$ decide whether they're true missing values or something else.
:::

# Other file formats {background-color="#001E44"}

## Other read_ functions {.scrollable .nostretch}

Once you know `read_csv()`, the others follow the same pattern, but have a different delimiter:

| Function | Use when |
|---|---|
| `read_csv()` | comma-separated (`.csv`) |
| `read_csv2()` | semicolon-separated (common in Europe where `,` is the decimal) |
| `read_tsv()` | tab-delimited (`.tsv`, `.txt`) |
| `read_delim()` | any delimiter; guesses if you don't specify |
| `read_fwf()` | fixed-width fields |

All of them accept a **file path or a URL** as the first argument.

```{r}
#| echo: TRUE
#| eval: FALSE
# From disk
read_tsv("data/survey_export.txt")

# From the web — no download needed
read_csv("https://some-data-portal.gov/dataset.csv")
```

## Reading from a URL {.scrollable .nostretch}

Any `read_*` function accepts a URL. This is useful for:

- Data portals that publish CSVs with a stable link
- Course data files hosted on GitHub or a course server
- Reproducibility: your script works on any machine without manual downloads

```{r}
#| echo: TRUE
#| eval: FALSE
houses <- read_csv(
  "https://mdbeckman.github.io/dcSupplement/data/houses-for-sale.csv"
)
```

::: {.callout-note}
If the URL is long, assign it to a variable first:
```r
url <- "https://..."
df  <- read_csv(url)
```
:::

# Writing data {background-color="#001E44"}

## write_csv() {.scrollable .nostretch}

To save a data frame to CSV:

```{r}
#| echo: TRUE
#| eval: FALSE
write_csv(students, "data/students-clean.csv")
```

**The catch:** column types are lost. When you read the file back, readr guesses types fresh and `meal_plan` goes back to character, not factor.

```{r}
#| echo: TRUE
#| eval: FALSE
read_csv("data/students-clean.csv")   # meal_plan is <chr> again
```

CSV is a plain-text format, so it stores values, not type metadata.

## Preserving types: write_rds() {.scrollable .nostretch}

For saving intermediate results within R, use **RDS**, which is R's native binary format. In this format, types are preserved exactly.

```{r}
#| echo: TRUE
#| eval: FALSE
write_rds(students, "data/students-clean.rds")
read_rds("data/students-clean.rds")   # meal_plan is <fct>, age is <dbl>
```

| Format | Readable outside R | Preserves types | File size |
|---|---|---|---|
| CSV | Yes | No | Medium |
| RDS | No | Yes | Small |

Use CSV to share data with others or for archiving. Use RDS when you're saving an intermediate result for your own pipeline and care about types.

# Data entry {background-color="#001E44"}

## tribble() and tibble() {.scrollable .nostretch}

Sometimes you need to key in a small table by hand. You have two options:

**`tibble()`**: column by column:

```{r}
#| echo: TRUE
tibble(
  x = c(1, 2, 5),
  y = c("h", "m", "g"),
  z = c(0.08, 0.83, 0.60)
)
```

**`tribble()`**: row by row (easier to read for small tables):

```{r}
#| echo: TRUE
tribble(
  ~x, ~y,   ~z,
   1, "h", 0.08,
   2, "m", 0.83,
   5, "g", 0.60
)
```

Column headers start with `~`. This is what we used to define `students` and `departments` for the join activity yesterday.

# Cleaning data {background-color="#001E44"}

## Strings to numbers {.scrollable .nostretch}

Real data often stores numbers as strings, such as formatted currency, percentages, European decimals.

**`as.numeric()`**: works on clean numeric strings:

```{r}
#| echo: TRUE
as.numeric(c("42", "3.14", "NA"))
```

**`parse_number()`**: ignores non-numeric characters:

```{r}
#| echo: TRUE
parse_number(c("$1,234.56", "17%", "9.8 m/s"))
```

`parse_number()` is more forgiving and usually the right choice for messy real-world values.

::: {.callout-warning}
`as.numeric("nine")` returns `NA` with a warning. If you have words, fix them first with `if_else()` or `case_when()` before parsing.
:::

## Recoding integer codes {.scrollable .nostretch}

Survey exports and government data often store categories as integers. They will typically provide a codebook that tells you what the numbers mean.

Strategy: build a lookup table, then `left_join()`:

```{r}
#| echo: TRUE
fuel_codes <- tribble(
  ~code, ~fuel_type,
      2, "gas",
      3, "electric",
      4, "oil"
)

houses_raw <- tibble(fuel = c(2, 3, 2, 4))

houses_raw |> left_join(fuel_codes, join_by(fuel == code))
```

## Dates with lubridate {.scrollable .nostretch}

Dates stored as strings need to be converted before R can do date arithmetic or plot them in order.

`lubridate` (part of the tidyverse, but needs `library(lubridate)`) provides functions named after the order of the components:

```{r}
#| echo: TRUE
library(lubridate)

mdy("June 9, 2026")          # month-day-year
ymd("2026-06-09")            # ISO 8601
dmy("9 June 2026")           # day-month-year
mdy_hm("06/09/2026 09:35")   # with time
```

Once a date is a proper `<date>` object, arithmetic works as you would expect:

```{r}
#| echo: TRUE
ymd("2026-06-26") - ymd("2026-06-09")  # days until end of course
```

::: {.callout-note}
Always check what order the components are in the original string. `mdy()` vs `dmy()` on the same string give different (wrong) dates.
:::

## Factors vs. character strings {.scrollable .nostretch}

R has a special type for categorical data: **factors**. They store categories efficiently and let you control the order of levels.

- `readr::read_csv()` reads text as **character** by default.
- The older `read.csv()` (base R) converts text to **factors** by default — this causes problems.

**When to convert to factor:**

```{r}
#| echo: TRUE
#| eval: FALSE
df |> mutate(region = factor(region))                        # unordered
df |> mutate(size = factor(size, levels = c("S","M","L")))  # ordered
```

**Watch out:** if you accidentally have a factor when you expect a string, `as.numeric()` gives the level number, not the value. Use `as.character()` first.

# Activity {background-color="#7C0A02"}

## In-class activity

Work through `day16-activity.R` in RStudio.

- **Part 1**: read in and clean the students CSV step by step
- **Part 2**: houses-for-sale data: integer codes, missing values, writing out
- **Part 3**: Saratoga dates (challenge)

[Activity file](day16-activity.R){download="day16-activity.R"}

## Wrap-up & looking ahead

:::: {.columns}
::: {.column width="55%"}
**What we covered**

- `read_csv()` and its key arguments: `na`, `col_types`, `col_names`
- `janitor::clean_names()`, `parse_number()`, `factor()`
- Other file types: `read_tsv()`, `read_delim()`, `read_csv2()`
- `write_csv()` vs `write_rds()`
- `tribble()` for hand-keyed tables
- Recoding integers via `left_join()`
- Dates with `lubridate`

**Upcoming**

- **Tomorrow (Wed Jun 10):** exam review
- **Thursday (Thu Jun 11):** in-class midterm exam
- **HW 4** due Mon Jun 15
:::
::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 7](https://r4ds.hadley.nz/data-import) · [readr reference](https://readr.tidyverse.org/reference)
:::
::::
