---
title: "Joining tables"
subtitle: "Monday, June 8, 2026"
lang: en
editor: source
engine: knitr
format:
  revealjs:
    slide-number: true
    embed-resources: true
    smaller: true
    highlight-style: a11y
    output-file: "day15-joins-slides.html"
  html:
    toc: true
    embed-resources: true
    highlight-style: a11y
    axe: true
    output-file: "day15-joins-page.html"
  pdf:
    fig-pos: "H"
---

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(tidyverse)
library(nycflights13)

LEGO_YELLOW <- "#F7C400"
LEGO_BLUE   <- "#006CB7"

students <- tribble(
  ~student, ~major,
  "A", "STAT",
  "B", "STAT",
  "C", "ECON",
  "D", "PSY",
  "E", NA,
  "F", "BIO"
)

departments <- tribble(
  ~code,  ~dept,
  "STAT", "Statistics",
  "ECON", "Economics",
  "PSY",  "Psychology",
  "ENG",  "Engineering"
)

plot_plate <- function(data) {
  data <- data |>
    mutate(text_col = if_else(fill == LEGO_YELLOW, "#1a1a1a", "white"))
  ggplot(data, aes(x = col, y = row)) +
    annotate("rect",
             xmin = min(data$col) - 0.6, xmax = max(data$col) + 0.6,
             ymin = 0.4,                 ymax = max(data$row) + 0.6,
             fill = "#00852B") +
    geom_tile(aes(fill = fill), width = 0.9, height = 0.9,
              color = "white", linewidth = 1.5) +
    geom_text(aes(label = label, color = text_col),
              fontface = "bold", size = 5) +
    scale_fill_identity() +
    scale_color_identity() +
    scale_x_continuous(expand = expansion(add = 0.7)) +
    scale_y_continuous(expand = expansion(add = 0.5)) +
    theme_void() +
    theme(plot.background = element_rect(fill = "white", color = NA))
}
```

::: {.callout-note style="font-size: 1.3em;"}
## [Today: Monday June 8]{style="font-size: 1.3em;"}

**Before class**

- **Perusall Assignment**: [*DC* §11.1--11.3](https://psu.instructure.com/courses/2487468/assignments/18325373)

**Plan for today**

- LEGO joins activity (opener, ~20 min)
- Why data gets split across tables
- Keys: what connects one table to another
- `left_join()`: the join you'll use most
- `join_by()`: specifying the match
- `anti_join()` and `semi_join()`: filtering by membership

**Helpful links**

- [Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 19](https://r4ds.hadley.nz/joins) · [*DC* Ch. 11](https://dtkaplan.github.io/DataComputingEbook/chap-join.html) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::

# Brick University Joins {background-color="#7C0A02"}

## Activity Overview {.scrollable .nostretch}

Each of you will get two sets of bricks, a baseplate, and a handout showing the two corresponding datasets. The two kinds of bricks are:

- **Yellow** — departments (STAT, ECON, PSY, ENG) — the **right table**
- **Blue** — students (A–F), each with a declared major — the **left table**

You'll work through **4 joins**, one at a time. Each slide describes a scenario and asks you to physically build the answer on your baseplate. After a minute or two, I'll show the answer on the projector.

::: {.callout-tip}
The goal here is to build intuition for what different joins actually *do*. In all four cases, `students` is the left table and `departments` is the right table.
:::

## The data {.scrollable .nostretch}

Each person gets 6 student bricks + 4 department bricks + 1 baseplate.

<iframe src="day15-activity.html" width="100%" height="500px" style="border: 1px solid #ccc; border-radius: 4px;"></iframe>

## Join 1 {.scrollable .nostretch}

*"The registrar is printing a student directory that pairs each student with their department. A student only appears if they've declared a major that maps to one of our four departments. Build the directory."*

**Left table: `students` &nbsp;·&nbsp; Right table: `departments`**

What join does this? Build it on the baseplate, then hold it up.

::: {.fragment}
**INNER JOIN** — keep only rows that match in *both* tables.

On the plate: 4 stacks -- A with STAT, B with STAT (stack them on the same brick), C with ECON, D with PSY.

Left off: E, F, and ENG.

E never declared, F declared a major we don't have, and ENG has no students. The inner join only keeps mutual matches.
:::

## Inner join {.scrollable .nostretch}

```{r}
#| echo: FALSE
#| fig-width: 9
#| fig-height: 3.5
tibble(
  col   = c(1, 1, 1, 2, 2, 3, 3),
  row   = c(1, 2, 3, 1, 2, 1, 2),
  label = c("STAT", "A", "B", "ECON", "C", "PSY", "D"),
  fill  = c(LEGO_YELLOW, LEGO_BLUE, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE)
) |> plot_plate()
```

Yellow = department (right table), blue = student (left table). E, F, and ENG are off the plate.

```{r}
#| echo: TRUE
#| eval: FALSE
inner_join(students, departments, join_by(major == code))
```

## Join 2 {.scrollable .nostretch}

*"Academic advising needs to schedule a check-in with every student on the roster -- including students who haven't declared, or whose major isn't one of ours. Show each student's department if we have it."*

**Left table: `students` &nbsp;·&nbsp; Right table: `departments`**

What join does this? Build it, then hold it up.

::: {.fragment}
**LEFT JOIN** — keep every row from the left table; attach matches from the right where they exist.

On the plate: the same 4 stacks, plus E solo and F solo -- all 6 students.

Left off: ENG.

E and F sit on the plate with no department brick attached. Those are exactly the students advising most needs to reach. Engineering, with no students, disappears entirely.

*Before revealing: try an inner join for this prompt. Ask the room -- did every student get a slot? When E and F are missing, that's the gap the left join fixes.*
:::

## Left join {.scrollable .nostretch}

```{r}
#| echo: FALSE
#| fig-width: 9
#| fig-height: 3.5
tibble(
  col   = c(1, 1, 1, 2, 2, 3, 3, 4, 5),
  row   = c(1, 2, 3, 1, 2, 1, 2, 1, 1),
  label = c("STAT", "A", "B", "ECON", "C", "PSY", "D", "E", "F"),
  fill  = c(LEGO_YELLOW, LEGO_BLUE, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_BLUE,   LEGO_BLUE)
) |> plot_plate()
```

All 6 students (left table) are on the plate. E and F have no yellow department brick -- `dept` is `NA` for them. ENG is off the plate.

```{r}
#| echo: TRUE
#| eval: FALSE
left_join(students, departments, join_by(major == code))
```

## Join 3 {.scrollable .nostretch}

*"The activities fair needs a table for every department -- staffed or not -- so prospective students can learn about programs that have few or no current majors. Show each department's students if it has any."*

**Left table: `students` &nbsp;·&nbsp; Right table: `departments`**

What join does this? Build it, then hold it up.

::: {.fragment}
**RIGHT JOIN** — keep every row from the right table; attach matches from the left where they exist.

On the plate: STAT (with A and B stacked), ECON (with C), PSY (with D), ENG solo -- all 4 departments.

Left off: E and F.

Engineering gets a table even with zero students. E and F drop out because they don't belong to any listed department.
:::

## Right join {.scrollable .nostretch}

```{r}
#| echo: FALSE
#| fig-width: 9
#| fig-height: 3.5
tibble(
  col   = c(1, 1, 1, 2, 2, 3, 3, 4),
  row   = c(1, 2, 3, 1, 2, 1, 2, 1),
  label = c("STAT", "A", "B", "ECON", "C", "PSY", "D", "ENG"),
  fill  = c(LEGO_YELLOW, LEGO_BLUE, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_YELLOW)
) |> plot_plate()
```

All 4 departments (right table) are on the plate. ENG has no blue student brick -- `student` is `NA` for it. E and F are off the plate.

```{r}
#| echo: TRUE
#| eval: FALSE
right_join(students, departments, join_by(major == code))
```

## Join 4 {.scrollable .nostretch}

*"The Provost wants a complete audit -- every student and every department must appear at least once, match or no match. Build the full picture."*

**Left table: `students` &nbsp;·&nbsp; Right table: `departments`**

What join does this? Build it, then hold it up.

::: {.fragment}
**FULL JOIN** — keep every row from both tables; fill in `NA` where there's no match.

On the plate: the 4 matched stacks + E solo + F solo + ENG solo. Every brick is used.

Solo student bricks (E, F) have `NA` for department. The solo ENG brick has `NA` for student. A left join and a right join at the same time.
:::

## Full join {.scrollable .nostretch}

```{r}
#| echo: FALSE
#| fig-width: 9
#| fig-height: 3.5
tibble(
  col   = c(1, 1, 1, 2, 2, 3, 3, 4, 5, 6),
  row   = c(1, 2, 3, 1, 2, 1, 2, 1, 1, 1),
  label = c("STAT", "A", "B", "ECON", "C", "PSY", "D", "E", "F", "ENG"),
  fill  = c(LEGO_YELLOW, LEGO_BLUE, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_YELLOW, LEGO_BLUE,
            LEGO_BLUE,   LEGO_BLUE,  LEGO_YELLOW)
) |> plot_plate()
```

Every brick is on the plate. Solo blue bricks (E, F) have `NA` for dept; solo yellow brick (ENG) has `NA` for student.

```{r}
#| echo: TRUE
#| eval: FALSE
full_join(students, departments, join_by(major == code))
```

## What we just built {.scrollable .nostretch}

| Join | On the baseplate | Left off |
|---|---|---|
| Inner | 4 matched stacks | E, F, ENG |
| Left | 4 stacks + E solo + F solo | ENG |
| Right | 4 stacks + ENG solo | E, F |
| Full | 4 stacks + E solo + F solo + ENG solo | nothing |

: {tbl-colwidths="[20,45,35]"}

In R the same four joins are `inner_join()`, `left_join()`, `right_join()`, and `full_join()`. The one you'll reach for most is `left_join()` -- keep everything in your main table, attach what you can from the right.

# Why multiple tables? {background-color="#001E44"}

## Data is split up on purpose {.scrollable .nostretch}

The `flights` dataset from nycflights13 has 336,776 rows, one per flight. Each row has a `carrier` column (`"UA"`, `"AA"`, etc.), but not the airline's full name.

The full names live in a separate `airlines` table:

```{r}
#| echo: TRUE
airlines
```

Storing the name separately isn't an oversight. The name belongs to the airline, not to each individual flight. Keeping it in one place means you only update it once, and you don't repeat "United Air Lines Inc." a hundred thousand times.

## Joins bring the tables back together {.scrollable .nostretch}

A **join** matches rows from one table to rows in another and combines them side by side:

```{r}
#| echo: TRUE
flights |>
  select(month, day, carrier, flight, origin, dest) |>
  left_join(airlines, join_by(carrier))
```

The `name` column is now attached to each flight. The airline name appears once in `airlines`, but as many times as needed in the result.

# Keys {background-color="#001E44"}

## The columns that connect tables {.scrollable .nostretch}

A **key** is a column (or set of columns) that identifies rows and links tables together.

A **primary key** uniquely identifies each row in its own table:

- `airlines$carrier`: each two-letter code appears exactly once
- `airports$faa`: each airport code appears exactly once
- `planes$tailnum`: each tail number appears exactly once

A **foreign key** refers to a primary key in another table:

- `flights$carrier` refers to `airlines$carrier`
- `flights$dest` (and `flights$origin`) refer to `airports$faa`
- `flights$tailnum` refers to `planes$tailnum`

The key columns often share the same name in both tables, which makes the join straightforward. When they don't, you say so explicitly.

## How the nycflights13 tables connect {.scrollable .nostretch}

`flights` is the central table. The others each add detail about one piece of a flight:

| Column(s) in `flights` | Connects to | Table |
|---|---|---|
| `carrier` | `carrier` | `airlines` |
| `tailnum` | `tailnum` | `planes` |
| `origin`, `dest` | `faa` | `airports` |
| `origin` + `time_hour` | `origin` + `time_hour` | `weather` |

: {tbl-colwidths="[38,22,40]"}

Both `origin` and `dest` refer to the same column in `airports`. You can join on either one depending on what you're asking.

## Checking a primary key {.scrollable .nostretch}

Before joining, it's worth verifying that a column is actually a primary key -- no duplicates. Count and look for any `n > 1`:

```{r}
#| echo: TRUE
airlines |>
  count(carrier) |>
  filter(n > 1)
```

An empty result confirms `carrier` is unique in `airlines`. If you got rows back, the column isn't a primary key and the join could produce more rows than you expected.

# left_join() {background-color="#001E44"}

## The join you'll use most {.scrollable .nostretch}

`left_join(x, y)` keeps every row from `x` and attaches matching columns from `y`. If there's no match in `y`, the new columns get `NA`.

```{r}
#| echo: TRUE
flights |>
  select(month, day, tailnum, carrier) |>
  left_join(planes |> select(tailnum, manufacturer, model),
            join_by(tailnum))
```

::: {.callout-tip}
Use `left_join()` when you want to add information to a main table. The left table stays intact; you're just pulling in extra columns from the right.
:::

## When there's no match: NA {.scrollable .nostretch}

Some tail numbers in `flights` don't appear in `planes`. `left_join()` keeps those rows anyway and fills the new columns with `NA`:

```{r}
#| echo: TRUE
flights |>
  filter(tailnum == "N3ALAA") |>
  select(month, day, tailnum, carrier) |>
  left_join(planes |> select(tailnum, manufacturer, model),
            join_by(tailnum))
```

That's intentional: the flight happened whether or not we have plane records. The missing information shows up as `NA` rather than disappearing silently.

# join_by() {background-color="#001E44"}

## When the column names match {.scrollable .nostretch}

When the key has the same name in both tables, `join_by()` is simple:

```{r}
#| echo: TRUE
#| eval: FALSE
flights |> left_join(airlines, join_by(carrier))   # carrier == carrier
flights |> left_join(planes,   join_by(tailnum))   # tailnum == tailnum
```

dplyr prints which columns it matched on, so you can check it did the right thing.

## When the column names differ {.scrollable .nostretch}

`airports$faa` is the primary key, but `flights` uses `origin` and `dest` to refer to airports. You tell R which columns correspond with `==` inside `join_by()`:

```{r}
#| echo: TRUE
flights |>
  select(month, day, origin, dest) |>
  left_join(airports |> select(faa, name, lat, lon),
            join_by(dest == faa))
```

`join_by(dest == faa)` says: match rows where `dest` in the left table equals `faa` in the right table.

## The natural join pitfall {.scrollable .nostretch}

If you omit `join_by()`, dplyr matches on **all** shared column names. This can go wrong quietly.

`flights` and `planes` both have a `year` column, but `flights$year` is the flight year and `planes$year` is the year the plane was built:

```{r}
#| echo: TRUE
# joins on both tailnum AND year -- almost certainly not what you want
flights |>
  select(year, month, day, tailnum) |>
  left_join(planes)
```

Because almost no plane was built in 2013 with a matching tailnum, the result is full of `NA`. Always specify `join_by()` when columns share a name but mean different things.

# Other mutating joins {background-color="#001E44"}

## left, inner, right, full {.scrollable .nostretch}

All four mutating joins add columns from the right table. The difference is which rows survive -- the same four cases you just built with LEGO:

| Join | Rows kept |
|---|---|
| `left_join()` | all rows from the **left** table |
| `inner_join()` | only rows with a match in **both** tables |
| `right_join()` | all rows from the **right** table |
| `full_join()` | all rows from **either** table |

: {tbl-colwidths="[38,62]"}

`left_join()` is the right choice for adding detail to a main table. `inner_join()` is useful when you only want rows that can be fully matched on both sides.

## left vs. inner: what changes {.scrollable .nostretch}

```{r}
#| echo: TRUE
# left_join: keeps all 336,776 flights; unmatched rows get NA
nrow(flights |> left_join(planes, join_by(tailnum)))

# inner_join: drops any flight with no matching plane record
nrow(flights |> inner_join(planes, join_by(tailnum)))
```

About 50,000 flights disappear with the inner join. Whether that's right depends on your question.

# Filtering joins {background-color="#001E44"}

## semi_join() and anti_join() {.scrollable .nostretch}

Filtering joins don't add columns. They filter rows based on whether a match exists in another table:

| Join | Keeps... |
|---|---|
| `semi_join(x, y)` | rows in `x` that **have** a match in `y` |
| `anti_join(x, y)` | rows in `x` that **don't** have a match in `y` |

: {tbl-colwidths="[42,58]"}

Only the columns of `x` appear in the result -- nothing from `y` is added. This is handy for checking what's present or absent in one table relative to another.

## anti_join(): finding what's missing {.scrollable .nostretch}

Which flights have a tail number that doesn't appear in `planes` at all?

```{r}
#| echo: TRUE
flights |>
  anti_join(planes, join_by(tailnum)) |>
  distinct(tailnum)
```

Over 700 unique tail numbers in `flights` have no matching record in `planes`. Worth knowing before you join in plane characteristics for an analysis.

## semi_join(): filtering by membership {.scrollable .nostretch}

`semi_join()` keeps only the rows from `x` that appear in `y`. Which airports actually appear as destinations in `flights`?

```{r}
#| echo: TRUE
airports |>
  semi_join(flights, join_by(faa == dest))
```

Only 101 of the 1,458 airports in `airports` are actual destinations in this dataset.

## What we covered today {.scrollable .nostretch}

- **Why joins**: data is split by design; joins combine it on demand without duplicating storage
- **Keys**: a primary key uniquely identifies rows in its table; a foreign key points to a primary key elsewhere
- **`left_join()`**: keeps all left rows, pulls in columns from the right; unmatched rows get `NA`
- **`join_by()`**: always specify the key; use `a == b` when column names differ between tables
- **Natural join pitfall**: shared column names get joined automatically and can silently give wrong results
- **Mutating join comparison**: left keeps all left rows; inner keeps only matches; right and full less common
- **`anti_join()`**: rows with no match; useful for finding data quality gaps
- **`semi_join()`**: rows with a match; useful for filtering by membership in another table

## Wrap-up & looking ahead

**Tomorrow (Tuesday):** importing data from CSV files, Excel, and web URLs, plus a look at factors vs. strings.

:::: {.columns}
::: {.column width="55%"}
**Upcoming Deadlines**

- **Tue 6/9, before class**: [*DC* §16.1--16.2, 16.4--16.5](https://psu.instructure.com/courses/2487468/assignments/18325374)
- **Mon 6/15, 11:59 p.m.**: [HW 4](https://psu.instructure.com/courses/2487468/assignments/18333004)
- **Thu 6/11, in class**: Exam
:::

::: {.column width="45%"}
**Reach out**

Stuck? Office hours, or email me ([cgc5478@psu.edu](mailto:cgc5478@psu.edu)) or Xinyue ([xpw5228@psu.edu](mailto:xpw5228@psu.edu)).

**Reference**

[Syllabus](../../syllabus/syllabus.qmd) · [AI policy](../../syllabus/academic-integrity-and-ai-policy.qmd) · [Schedule](../../schedule.qmd) · [*R4DS* Ch. 19](https://r4ds.hadley.nz/joins) · [*DC* Ch. 11](https://dtkaplan.github.io/DataComputingEbook/chap-join.html) · [dplyr reference](https://dplyr.tidyverse.org/reference)
:::
::::
