---
title: "STAT 184 — Homework 0: Setup Confirmation"
subtitle: "Due Wednesday, May 20 at 11:59 PM on Canvas"
author: "Your Name Here"
format:
  html:
    toc: true
    code-fold: false
  pdf:
    toc: false
editor: source
---

**AI Disclosure**

- [ ] No AI used
- [ ] Syntax / Debugging (e.g., finding a missing comma, fixing an error code)
- [ ] Conceptual explanation (e.g., asking how a `left_join` works)
- [ ] Code Generation (e.g., asking AI to write a function from scratch)

*Briefly describe your use (1-2 sentences):* 

[Type your explanation here]

---

## Welcome to STAT 184! 

The goal of this assignment is to ensure your RStudio environment is configured correctly, your core packages are installed, and you know how to render (knit) a Quarto document. 

## Question 1: Setup

Uncomment and run the `install.packages(tidyverse")` line ONCE by deleting the `# ` to install the "tidyverse" package on your machine. Make sure to remove any leading whitespace. Then re-comment by adding back the `# `. It is important to re-comment the line (otherwise, your file won't render).

```{r setup}
#| message: false
#| warning: false
# install.packages("tidyverse") 
library(tidyverse)
```

## Question 2: Global Options

In lecture, we discussed changing your RStudio Global Options under `Tools > Global Options > General` to the following:

- Restore .RData into workspace at startup: *Unchecked*
- Save workspace to .RData on exit: *Never*

### Part (a): Why is it important to change these specific settings when working reproducibly in R?

[Write your answer here.]

### Part (b): By typing your name below, you confirm that you have successfully changed these settings on the computer you will use for this course.

[Type your name here to confirm.]

## Question 3: Package Check

To ensure that your R environment is working and the `tidyverse` package is loaded correctly in the setup chunk, run the code below to print out the first few rows of a built-in dataset called `mpg`. 

```{r}
# Replace the comment below with the following code: head(mpg)
# (DELETE this line and write your code here.)
```

## Question 4: Rendering

Now that you have answered the questions, click the **Render** button at the top of your RStudio script window. It should generate a HTML or PDF file.

Submit that rendered file to Canvas.