---
title: "STAT 184 - Homework 4: Joining Tables"
subtitle: "Due Monday, June 15 at 11:59 PM on Canvas"
author: "Your Name Here"
format:
  html:
    toc: TRUE
    code-fold: FALSE
    embed-resources: TRUE
  pdf:
    toc: FALSE
    include-in-header:
      text: |
        \usepackage[labelfont=bf, format=plain, labelsep=colon]{caption}
execute:
  enabled: true
editor: source
---

**AI Disclosure**

- [ ] No AI used
- [ ] Syntax / Debugging (e.g., finding a missing comma, fixing an error code)
- [ ] Conceptual explanation (e.g., asking how `left_join()` works)
- [ ] Code Generation (e.g., asking AI to write a join pipeline from scratch)

*Briefly describe your use (1-2 sentences):*

[Type your explanation here]

---

### Setup

Run the chunk below to load the packages. `nycflights13` should already be installed from HW 3. If not, run `install.packages("nycflights13")` in the **Console** once.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
```

This homework uses the five related tables in `nycflights13`: `flights`, `airlines`, `airports`, `planes`, and `weather`. Take a moment to look at `flights`:

```{r}
glimpse(flights)
```

Throughout this homework, the key relationship to keep in mind: `flights` is the central table. The other four tables each add detail about one piece of a flight -- the airline, the airports, the plane, or the weather conditions.

---

## Question 1 -- left_join basics

`left_join()` keeps every row from the left table and attaches matching columns from the right. If there's no match, the new columns get `NA`.

### Part (a): add airline names

The `carrier` column in `flights` holds two-letter codes. Use `left_join()` to attach the full airline name from `airlines`, then build a summary: **for each airline, how many flights departed, and what was the average departure delay?** Include only carriers with at least 1,000 flights. Arrange by average delay, highest first. Show the result.

```{r}
# Write your R code here
```

In one sentence: which airline had the worst average departure delay, and does anything about that surprise you?

[Write your answer here.]

### Part (b): add plane details

Use `left_join()` to attach `manufacturer` and `seats` from `planes` to a filtered version of `flights`. Narrow `flights` down to just `month`, `day`, `carrier`, `tailnum`, `origin`, and `dest` first, then join. Show the first six rows with `head()`.

```{r}
# Write your R code here
```

### Part (c): when there's no match

Not every tail number in `flights` has a record in `planes`. After joining in Part (b), filter the result to only rows where `manufacturer` is `NA`. How many such flights are there?

```{r}
# Write your R code here
```

In one sentence: what does it mean that `manufacturer` is `NA` here?

[Write your answer here.]

---

## Question 2 -- join_by() with different column names

Sometimes the key column has a different name in the two tables. `join_by(a == b)` says: match rows where column `a` in the left table equals column `b` in the right table.

The `airports` table has a primary key called `faa`. In `flights`, both `origin` and `dest` refer to airport codes -- but they're named differently from `faa`.

### Part (a): destination airport names

Join `airports` onto `flights` using `dest` as the key. Select only `month`, `day`, `origin`, `dest`, and `name` (the destination airport's full name) from the result. Show the first six rows.

```{r}
# Write your R code here
```

### Part (b): most popular destinations

Using the join from Part (a), find the **10 most common destination airports by number of flights**. Your result should show the airport code, the full airport name, and the flight count. Arrange by count descending.

```{r}
# Write your R code here
```

### Part (c): destination coordinates

Join the latitude (`lat`) and longitude (`lon`) from `airports` onto flights using `dest`. Then compute the **average arrival delay by destination airport**, keeping only airports with at least 100 flights. Join the airport name in as well so the table is readable. Arrange by average delay descending and show the top 10.

```{r}
# Write your R code here
```

In one sentence: do the airports with the worst average arrival delays share anything in common (region, size, etc.)?

[Write your answer here.]

---

## Question 3 -- anti_join and semi_join

Filtering joins don't add columns -- they filter rows based on whether a match exists in another table.

- `anti_join(x, y)`: keeps rows in `x` with **no** match in `y`
- `semi_join(x, y)`: keeps rows in `x` **with** a match in `y`

### Part (a): flights with no plane record

Use `anti_join()` to find all flights whose `tailnum` does not appear in `planes`. How many such flights are there? Then use `count()` on `carrier` to see which airlines account for most of the missing records.

```{r}
# Write your R code here
```

In one or two sentences: which carriers account for most of the unmatched tail numbers? What might explain this?

[Write your answer here.]

### Part (b): filtering airports by flight activity

The `airports` table has 1,458 airports, but most of them never appear as a destination in `flights`. Use `semi_join()` to keep only the airports that appear as a `dest` in `flights`. How many airports remain?

```{r}
# Write your R code here
```

### Part (c): thinking about the joins

In one or two sentences each, explain:

1. Why would you use `anti_join()` instead of just `filter()` to find unmatched rows?
2. If `anti_join(flights, planes, join_by(tailnum))` gives you rows, what does that tell you about the data?

[Write your answer here.]

---

## Question 4 -- joins in a pipeline

Joins fit naturally into the same pipelines you've been building all semester. This question asks you to chain joins with the wrangling verbs from weeks 3 and 4.

### Part (a): average delay by airline name

The goal: a table showing **average arrival delay by full airline name**, for airlines with at least 500 flights, arranged from worst to best. You'll need to join `airlines` to get the name, group by it, summarise, filter, and arrange -- all in one pipeline.

```{r}
# Write your R code here
```

### Part (b): delay by plane age

The `planes` table has a `year` column giving the year each plane was manufactured. Join `planes` onto `flights`, compute the plane's age as `2013 - year` (since `flights` is from 2013), then group by age and compute average departure delay. Show only planes where age is known (not `NA`) and where the age group has at least 100 flights.

Make a scatter plot with plane age on the x-axis and average departure delay on the y-axis.

```{r}
# Write your R code here
```

In one or two sentences: is there a visible relationship between plane age and departure delay in this data?

[Write your answer here.]

---

# Code appendix

```{r ref.label=knitr::all_labels()}
#| echo: TRUE
#| eval: FALSE
```
